---
title: "1. QC and Data Preparation - Concatenation vs Integration Comparison"
output: html_document
date: "`r Sys.Date()`"
---

## Load Libraries

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tibble)

# Memory monitoring
check_memory <- function(stage = "") {
  mem_info <- gc()
  used_mb <- (mem_info[1, 2] * 8) / (1024^2)  # Convert to MB
  cat("Memory usage", stage, ":", round(used_mb, 1), "MB\n")
  return(used_mb)
}

# Create output directories (local to QC folder)
dir.create("data", recursive = TRUE, showWarnings = FALSE)
dir.create("figures", recursive = TRUE, showWarnings = FALSE)

cat("=== QC AND DATA PREPARATION ===\n")
initial_memory <- check_memory("at start")
```

## Load Spatial Data

```{r load-data}
# Data loading
# Visium SpaceRanger sample outputs (path configured in config.R)
source(here::here("config.R"))  # defines DATA_DIR, RESULTS_DIR, ANALYSIS_SUBTYPES_DIR
data_dir <- file.path(DATA_DIR, "SPT/samples_3.3.1")

# Grab every "outs/" directory one level down
all_dirs  <- list.dirs(data_dir, full.names = TRUE, recursive = 1)
outs_dirs <- all_dirs[basename(all_dirs) == "outs"]

# Keep only valid Visium directories
valid_outs <- Filter(function(d) {
  has_h5   <- file.exists(file.path(d, "filtered_feature_bc_matrix.h5"))
  has_dir  <- dir.exists(file.path(d, "filtered_feature_bc_matrix"))
  has_spat <- dir.exists(file.path(d, "spatial"))
  (has_h5 || has_dir) && has_spat
}, outs_dirs)

if (length(valid_outs) == 0) {
  stop("No valid Visium 'outs/' directories found in ", data_dir)
}

cat("Found", length(valid_outs), "valid Visium folders\n")

# Smart loader function (from original QC script)
fix_spatial_scales_safe <- function(outs_path) {
  tryCatch({
    # Pick the right filename argument
    if (file.exists(file.path(outs_path, "filtered_feature_bc_matrix.h5"))) {
      so <- Load10X_Spatial(outs_path, filename = "filtered_feature_bc_matrix.h5")
    } else {
      so <- Load10X_Spatial(outs_path, filename = "filtered_feature_bc_matrix")
    }
    
    # Fix any spatial coordinate scaling issues
    if ("spatial" %in% names(so@images)) {
      scale_factors <- so@images$spatial@scale.factors
      if (is.null(scale_factors$tissue_hires_scalef) || scale_factors$tissue_hires_scalef == 0) {
        so@images$spatial@scale.factors$tissue_hires_scalef <- 0.1
      }
      if (is.null(scale_factors$tissue_lowres_scalef) || scale_factors$tissue_lowres_scalef == 0) {
        so@images$spatial@scale.factors$tissue_lowres_scalef <- 0.05
      }
    }
    
    return(so)
  }, error = function(e) {
    cat("Error loading", outs_path, ":", e$message, "\n")
    return(NULL)
  })
}

# Load all samples
seurat_list <- list()
for (outs_path in valid_outs) {
  sample_name <- basename(dirname(outs_path))
  cat("Loading sample:", sample_name, "\n")
  
  so <- fix_spatial_scales_safe(outs_path)
  
  if (!is.null(so)) {
  
  # Add sample metadata
  so$sample_id <- sample_name
  so$orig.ident <- sample_name
  
  # Extract cell line using SAME pattern as integration script
  extracted_cell_line <- gsub(".*_(FaDu|UT5|Cal33|UT8|SAS|UT45|SAT).*", "\\1", sample_name)
  
  # If extraction failed, mark as "Unknown" instead of using the full sample name
  if (extracted_cell_line == sample_name) {
    cat(" Cell line extraction failed for:", sample_name, "\n")
    so$cell_line <- "Unknown"
  } else {
    so$cell_line <- extracted_cell_line
  }
  
  # NO TREATMENT INFO - this is just spatial transcriptomics data from different cell lines
  # "SPT" likely means "Spatial" not a treatment
  
  seurat_list[[sample_name]] <- so
  }
}

# Remove any NULL entries
seurat_list <- Filter(Negate(is.null), seurat_list)

# DIAGNOSTIC: Check all cell lines detected during loading
all_cell_lines_detected <- sapply(seurat_list, function(so) unique(so$cell_line)[1])
cat("\n=== CELL LINE EXTRACTION DIAGNOSTIC ===\n")
cat("Total samples loaded:", length(seurat_list), "\n")
for(i in seq_along(seurat_list)) {
  sample_name <- names(seurat_list)[i]
  cell_line <- all_cell_lines_detected[i]
  cat("Sample:", sample_name, "→ Cell line:", cell_line, "\n")
}
cat("Unique cell lines detected:", paste(unique(all_cell_lines_detected), collapse = ", "), "\n")
cat("Number of unique cell lines:", length(unique(all_cell_lines_detected)), "\n")

cat("Loaded", length(seurat_list), "samples successfully\n")
check_memory("after loading samples")
```


## QC Metrics Analysis

```{r qc-metrics, fig.width=15, fig.height=12}
# Calculate comprehensive QC metrics
cat("=== CALCULATING QC METRICS ===\n")

# Patterns for gene types
mt_patterns <- c("^MT-", "^mt-", "^Mt-", "MITO", "mito")
ribo_patterns <- c("^RPS", "^RPL", "^rps", "^rpl")

for (i in seq_along(seurat_list)) {
  sample_name <- names(seurat_list)[i]
  so <- seurat_list[[sample_name]]
  
  cat("Processing QC for:", sample_name, "\n")
  
  # Check for mitochondrial genes
  mt_found <- FALSE
  for (pattern in mt_patterns) {
    mt_genes <- grep(pattern, rownames(so), value = TRUE)
    if (length(mt_genes) > 0) {
      so <- PercentageFeatureSet(so, pattern = pattern, col.name = "percent_mt")
      mt_found <- TRUE
      break
    }
  }
  if (!mt_found) {
    so$percent_mt <- 0
    cat("  No mitochondrial genes found\n")
  }
  
  # Check for ribosomal genes
  ribo_found <- FALSE
  for (pattern in ribo_patterns) {
    ribo_genes <- grep(pattern, rownames(so), value = TRUE)
    if (length(ribo_genes) > 0) {
      so <- PercentageFeatureSet(so, pattern = pattern, col.name = "percent_ribo")
      ribo_found <- TRUE
      break
    }
  }
  if (!ribo_found) {
    so$percent_ribo <- 0
    cat("  No ribosomal genes found\n")
  }
  
  # Additional QC metrics
  so$log10_nCount <- log10(so$nCount_Spatial + 1)
  so$log10_nFeature <- log10(so$nFeature_Spatial + 1)
  
  seurat_list[[sample_name]] <- so
}

# Create comprehensive QC plots
qc_metrics <- c("nCount_Spatial", "nFeature_Spatial", "percent_mt", "percent_ribo")

# Plot 1: Individual sample QC distributions
qc_plots_individual <- list()
for (metric in qc_metrics) {
  plots_per_metric <- list()
  
  for (i in seq_len(min(6, length(seurat_list)))) {  # Show first 6 samples
    sample_name <- names(seurat_list)[i]
    so <- seurat_list[[sample_name]]
    
    if (metric %in% colnames(so@meta.data)) {
      p <- VlnPlot(so, features = metric, pt.size = 0.1) +
        ggtitle(paste(sample_name, metric)) +
        theme(axis.text.x = element_blank(),
              axis.title.x = element_blank())
      plots_per_metric[[sample_name]] <- p
    }
  }
  
  if (length(plots_per_metric) > 0) {
    combined_plot <- wrap_plots(plots_per_metric, ncol = 3)
    qc_plots_individual[[metric]] <- combined_plot
  }
}

# Display individual QC plots
for (metric in names(qc_plots_individual)) {
  print(qc_plots_individual[[metric]])
}
```

## Combined QC Summary

```{r qc-summary, fig.width=16, fig.height=10}
# Combine metadata from all samples
combined_metadata <- do.call(rbind, lapply(names(seurat_list), function(sample_name) {
  so <- seurat_list[[sample_name]]
  data.frame(
    sample_id = sample_name,
    cell_line = so$cell_line,
    nCount_Spatial = so$nCount_Spatial,
    nFeature_Spatial = so$nFeature_Spatial,
    percent_mt = so$percent_mt,
    percent_ribo = so$percent_ribo,
    stringsAsFactors = FALSE
  )
}))

# QC summary statistics
cat("=== QC SUMMARY STATISTICS ===\n")
qc_summary <- combined_metadata %>%
  group_by(cell_line) %>%
  summarise(
    n_spots = n(),
    mean_UMI = round(mean(nCount_Spatial)),
    median_UMI = round(median(nCount_Spatial)),
    mean_genes = round(mean(nFeature_Spatial)),
    median_genes = round(median(nFeature_Spatial)),
    mean_mt = round(mean(percent_mt), 2),
    mean_ribo = round(mean(percent_ribo), 2),
    .groups = 'drop'
  )

print(qc_summary)

# Comprehensive QC visualization
p1 <- ggplot(combined_metadata, aes(x = cell_line, y = nCount_Spatial, fill = cell_line)) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.1, alpha = 0.5) +
  scale_y_log10() +
  labs(title = "UMI Counts by Cell Line", 
       y = "UMI Count (log10)", x = "Cell Line") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- ggplot(combined_metadata, aes(x = cell_line, y = nFeature_Spatial, fill = cell_line)) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.1, alpha = 0.5) +
  scale_y_log10() +
  labs(title = "Gene Count by Cell Line",
       y = "Gene Count (log10)", x = "Cell Line") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p3 <- ggplot(combined_metadata, aes(x = nCount_Spatial, y = nFeature_Spatial, color = cell_line)) +
  geom_point(alpha = 0.6, size = 0.8) +
  scale_x_log10() + scale_y_log10() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "UMI vs Gene Count Correlation",
       x = "UMI Count (log10)", y = "Gene Count (log10)") +
  theme_minimal()

p4 <- ggplot(combined_metadata, aes(x = cell_line, fill = cell_line)) +
  geom_bar() +
  labs(title = "Sample Distribution by Cell Line",
       y = "Number of Spots", x = "Cell Line") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Combine all QC plots
combined_qc_plot <- (p1 | p2) / (p3 | p4)
print(combined_qc_plot)

# Save QC plot
ggsave("figures/01_comprehensive_qc_analysis.pdf", combined_qc_plot, 
       width = 16, height = 10, dpi = 300)
```

## Filtering Assessment

```{r filtering-assessment, fig.width=14, fig.height=8}
cat("=== FILTERING ASSESSMENT ===\n")

# Assess potential filtering thresholds
filtering_stats <- combined_metadata %>%
  summarise(
    total_spots = n(),
    low_umi_100 = sum(nCount_Spatial < 100),
    low_umi_200 = sum(nCount_Spatial < 200),
    low_genes_50 = sum(nFeature_Spatial < 50),
    low_genes_100 = sum(nFeature_Spatial < 100),
    high_umi_q99 = sum(nCount_Spatial > quantile(nCount_Spatial, 0.99)),
    high_genes_q99 = sum(nFeature_Spatial > quantile(nFeature_Spatial, 0.99))
  )

cat("Filtering assessment:\n")
cat("Total spots:", filtering_stats$total_spots, "\n")
cat("Spots with <100 UMIs:", filtering_stats$low_umi_100, 
    "(", round(filtering_stats$low_umi_100/filtering_stats$total_spots*100, 1), "%)\n")
cat("Spots with <200 UMIs:", filtering_stats$low_umi_200,
    "(", round(filtering_stats$low_umi_200/filtering_stats$total_spots*100, 1), "%)\n")
cat("Spots with <50 genes:", filtering_stats$low_genes_50,
    "(", round(filtering_stats$low_genes_50/filtering_stats$total_spots*100, 1), "%)\n")
cat("Spots with <100 genes:", filtering_stats$low_genes_100,
    "(", round(filtering_stats$low_genes_100/filtering_stats$total_spots*100, 1), "%)\n")

# Filtering recommendation
if (filtering_stats$low_umi_100/filtering_stats$total_spots > 0.05) {
  cat("RECOMMEND FILTERING: >5% spots have very low UMIs\n")
  recommended_filter <- "nCount_Spatial > 100"
} else if (filtering_stats$low_genes_50/filtering_stats$total_spots > 0.05) {
  cat("CONSIDER FILTERING: >5% spots have very low gene counts\n")
  recommended_filter <- "nFeature_Spatial > 50"
} else {
  cat("NO FILTERING NEEDED: Most spots have reasonable expression\n")
  recommended_filter <- "none"
}

# Filtering visualization
p_filter1 <- ggplot(combined_metadata, aes(x = nCount_Spatial)) +
  geom_histogram(bins = 50, alpha = 0.7, fill = "steelblue") +
  geom_vline(xintercept = 100, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 200, color = "orange", linetype = "dashed") +
  scale_x_log10() +
  labs(title = "UMI Count Distribution", x = "UMI Count (log10)", y = "Frequency") +
  theme_minimal()

p_filter2 <- ggplot(combined_metadata, aes(x = nFeature_Spatial)) +
  geom_histogram(bins = 50, alpha = 0.7, fill = "forestgreen") +
  geom_vline(xintercept = 50, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 100, color = "orange", linetype = "dashed") +
  scale_x_log10() +
  labs(title = "Gene Count Distribution", x = "Gene Count (log10)", y = "Frequency") +
  theme_minimal()

filtering_plot <- p_filter1 | p_filter2
print(filtering_plot)

ggsave("figures/02_filtering_assessment.pdf", filtering_plot, 
       width = 14, height = 6, dpi = 300)
```

## Save Processed Data

```{r save-data}
cat("=== SAVING PROCESSED DATA ===\n")

# Save individual samples with metadata
for (sample_name in names(seurat_list)) {
  saveRDS(seurat_list[[sample_name]], 
          file = paste0("data/", sample_name, "_qc_with_metadata.rds"))
}

# Save QC summary
saveRDS(combined_metadata, "data/qc_metadata_summary.rds")
saveRDS(qc_summary, "data/qc_statistics_summary.rds")

# Save filtering recommendation
writeLines(recommended_filter, "data/filtering_recommendation.txt")

cat("Saved QC data and metadata\n")
cat("Recommended filtering:", recommended_filter, "\n")

final_memory <- check_memory("at end")
cat("Memory change:", round(final_memory - initial_memory, 1), "MB\n")
```

## Summary

**QC Analysis Complete:**

- Loaded `r length(seurat_list)` spatial samples
- Calculated comprehensive QC metrics
- Assessed mitochondrial/ribosomal gene content
- Generated filtering recommendations
- Created visualization plots
- Saved processed data for comparison analysis

**Next Steps:**
1. Run Script 2A (Concatenation approach)
2. Run Script 2B (Integration approach)  
3. Compare results from both methods
