---
title: "GSEA_LOOP_HPV_E220_Final"
author: Safayat Mahmud Khan
output: html_document
date: "2023-10-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File input with msigdb categories 

### ( Change the function settings if necessary). If you don't need all the categories, you can also edit it.

```{r include=FALSE}
library(clusterProfiler)
library(enrichplot)
library(DOSE)
library(org.Hs.eg.db)
library(msigdbr)
library(dplyr)
library(GSEAmining)
library(GSEABase)
# List of input files (replace with your actual file names)
input_files <- c("put your deseq2 output files here")

# Load OrgDb for humans
human_orgdb <- org.Hs.eg.db

# Specify the categories you want to analyze
categories_to_analyze <- c("C5","H") 

# Function to perform GSEA for a specific category and input file
perform_gsea <- function(category, input_data) {
  lfc_vector_c <- input_data$log2FoldChange
  names(lfc_vector_c) <- input_data$gene_symbol
  lfc_vector_c <- sort(lfc_vector_c, decreasing = TRUE)
  
  set.seed(2020)
  
  tryCatch({
    gsea_result <- GSEA(geneList = lfc_vector_c,
                        TERM2GENE = dplyr::select(category, gs_name, gene_symbol),
                        minGSSize = 10, # Minimum gene set size
                        maxGSSize = 1000, # Maximum gene set size
                        seed = TRUE,
                        pvalueCutoff = 0.05, # p-value cutoff
                        pAdjustMethod = "BH" # Benjamini-Hochberg correction
    )
    return(gsea_result)
  }, error = function(e) {
    return(NULL) # Return NULL for failed analyses
  })
}

# Create a list to store GSEA results for each category and each input file
all_results_list <- list()

# Perform GSEA analysis for each category and each input file
for (category_name in categories_to_analyze) {
  # Find the category data frame that matches the category name
  category <- msigdbr(species = "Homo sapiens", category_name)
  
  if (!is.null(category)) {
    category_results <- list()
    
    for (input_file in input_files) {
      data <- read.csv(input_file)
      gsea_result <- perform_gsea(category, data)
      category_results[[input_file]] <- gsea_result
    }
    
    all_results_list[[category_name]] <- category_results
  } else {
    print(paste("Category not found:", category_name))
  }
}

# Function to get a GSEA result data frame from all_results_list based on category name and input file name
get_gsea_result <- function(category_name, input_file) {
  # Construct the name of the list element
  result_name <- paste0("gsea_results_", category_name, "_", tools::file_path_sans_ext(basename(input_file)))
  
  # Retrieve the GSEA result data frame from all_results_list
  gsea_result <- all_results_list[[category_name]][[input_file]]
  
  # Check if the result is NULL (i.e., the GSEA analysis failed for this category and input file)
  if (is.null(gsea_result)) {
    warning(paste("No GSEA result found for category", category_name, "and input file", input_file))
    return(NULL)
  }
  
  return(as.data.frame(gsea_result))
}

# Define the results directory
results_directory <- "GSEAresults_loopcode"

# Check if the directory exists, if not, create it
if (!file.exists(results_directory)) {
  dir.create(results_directory, showWarnings = FALSE)
}
# Save GSEA results data frames to files for each category and each input file (optional)
for (category_name in names(all_results_list)) {
  category_results <- all_results_list[[category_name]]
  
  for (i in seq_along(input_files)) {
    input_file <- input_files[i]
    result <- category_results[[input_file]]
    
    if (!is.null(result)) {
      output_file <- file.path(results_directory, paste0("gsea_results_", category_name, "_", tools::file_path_sans_ext(basename(input_file)), ".csv"))
      write.csv(as.data.frame(result), file = output_file, row.names = FALSE)
    }
  }
}

# New function to save merged GSEA results per input file
save_merged_results <- function(all_results, input_file) {
  merged_results <- data.frame()
  for (category_name in names(all_results)) {
    result <- all_results[[category_name]][[input_file]]
    if (!is.null(result) && nrow(result) > 0) {
      result_df <- as.data.frame(result)
      result_df$Category <- category_name
      merged_results <- rbind(merged_results, result_df)
    }
  }
  # Define directory and file names
  input_file_directory <- file.path("GSEAresults_loopcode", tools::file_path_sans_ext(basename(input_file)))
  dir.create(input_file_directory, recursive = TRUE, showWarnings = FALSE)
  output_file <- file.path(input_file_directory, "merged_GSEA_results.csv")
  
  # Write merged results to CSV
  write.csv(merged_results, file = output_file, row.names = FALSE)
  
  # Save merged GSEA results object
  saveRDS(merged_results, file = file.path(input_file_directory, "merged_GSEA_results.rds"))
}

# Now loop through each input file to save merged GSEA results
for (input_file in input_files) {
  save_merged_results(all_results_list, input_file)
}


```


# Create dotplots 



```{r echo=TRUE}
# Load necessary libraries
library(ggplot2)

# Create base directories
dir.create(file.path(results_directory, "doseplots"), showWarnings = FALSE, recursive = TRUE)

# Updated save_plots function
save_plots <- function(plots_list, base_directory, category_name, input_file_name, width = 10, height = 8) {
  # Create category directory
  category_directory <- file.path(base_directory, category_name)
  dir.create(category_directory, showWarnings = FALSE)
  
  # Create input file directory
  input_file_directory <- file.path(category_directory, input_file_name)
  dir.create(input_file_directory, showWarnings = FALSE)
  
  # Iterate over each plot in the list and save it
  for (geneSetID in names(plots_list)) {
    plot <- plots_list[[geneSetID]]
    
    # Define the file name
    file_name_base <- file.path(input_file_directory, geneSetID)
    
    # Save in PDF and PNG formats with specified dimensions
    ggsave(paste0(file_name_base, ".pdf"), plot, width = width, height = height)
    ggsave(paste0(file_name_base, ".png"), plot, width = width * 100, height = height * 100, units = "px")
  }
}

# Loop through each category and input file
for (category_name in names(all_results_list)) {
  category_results <- all_results_list[[category_name]]
  
  for (input_file in input_files) {
    result <- category_results[[input_file]]
    
    if (!is.null(result) && nrow(result) > 0) {  # Check if result is not null and has rows
      
      input_file_name <- tools::file_path_sans_ext(basename(input_file))
      
      # Doseplot with custom title that includes category and input file name
      dose_plot <- dotplot(result, showCategory=15, split=".sign", font.size=8) + facet_grid(.~.sign) +
                   ggtitle(paste("Category:", category_name, "| Input File:", input_file_name))  # Add title
      
      # Determine the NES sign presence in the result
      has_positive_nes <- any(result$NES > 0)
      has_negative_nes <- any(result$NES < 0)
      
      # Apply facet_grid only if both positive and negative NES values are present
      if (has_positive_nes && has_negative_nes) {
        dose_plot <- dose_plot + facet_grid(.~.sign)
      }
      
      # Print the plot to show it in the knitted output
      print(dose_plot)
      
      # Save the plot to the directory
      dose_plot_list <- list(dose_plot = dose_plot)  # Wrap the dose plot in a list to be compatible with save_plots function
      save_plots(dose_plot_list, file.path(results_directory, "doseplots"), category_name, input_file_name)
    }
  }
}

```

# code with enriched and dendoplot but still the text sizes need to be changed. This code will sometimes generate error when it doesn't find enough dendograms. I couldn't fix it, as the package is not so well defined. Still, it's good if you want to see overall summary in the enriched plots. You can ignore it.

```{r eval=FALSE, include=FALSE}
# Function to generate plots
generate_plots <- function(data, input_file_directory) {
  
   #Generate dendrogram plot
  png(filename = file.path(input_file_directory, "dendrogram_plot.png"))
  gm_dendplot(data$gs_filtered, data$gs_clustered)
  dev.off()
  
   #Save dendrogram plot to PDF as well
  pdf(file = file.path(input_file_directory, "dendrogram_plot.pdf"))
  gm_dendplot(data$gs_filtered, data$gs_clustered)
  dev.off()
  
  # Generate enriched terms plot
  enriched_terms_plot <- gm_enrichterms(data$gs_filtered, data$gs_clustered)
  
  # Generate enrichment cores plot
  enrichment_cores_plot <- gm_enrichcores(data$gs_filtered, data$gs_clustered)
  
  # Save individual plots (if they are ggplot objects)
  if ("gg" %in% class(enriched_terms_plot) || "ggplot" %in% class(enriched_terms_plot))
    ggsave(filename = file.path(input_file_directory, "enriched_terms_plot.png"), plot = enriched_terms_plot)
  if ("gg" %in% class(enrichment_cores_plot) || "ggplot" %in% class(enrichment_cores_plot))
    ggsave(filename = file.path(input_file_directory, "enrichment_cores_plot.png"), plot = enrichment_cores_plot)
  
  # Save plots in PDF format
  if (!is.null(enriched_terms_plot)) {
    pdf(file = file.path(input_file_directory, "enriched_terms_plot.pdf"))
    print(enriched_terms_plot)
    dev.off()
  }
  
  if (!is.null(enrichment_cores_plot)) {
    pdf(file = file.path(input_file_directory, "enrichment_cores_plot.pdf"))
    print(enrichment_cores_plot)
    dev.off()
  }
}




# Loop through each input file
for (input_file in input_files) {
  
  # Use tryCatch to handle errors and continue to the next file
  tryCatch({
    
    # Define the directory and file names
    input_file_directory <- file.path("GSEAresults_loopcode", tools::file_path_sans_ext(basename(input_file)))
    merged_results_file <- file.path(input_file_directory, "merged_GSEA_results.csv")
    
    # Load the merged GSEA results
    merged_results <- read.csv(merged_results_file)
    
    # Assume that you have filtered and clustered your data
    # Replace the following line with your actual code for filtering and clustering
    gs_data <- list(gs_filtered = gm_filter(merged_results), gs_clustered = gm_clust(gm_filter(merged_results)))
    
    # Check if the filtered data is empty
    if (nrow(gs_data$gs_filtered) == 0) {
      message(paste("No significant pathways for", input_file, ". Skipping this file."))
      next
    }
    
    # Generate and save the plots
    generate_plots(gs_data, input_file_directory)
    
  }, error = function(e) {
    # If an error occurs, print a message and continue to the next file
    message(paste("Error processing", input_file, ": ", e$message, ". Continuing to the next file."))
  })
  
}



```







