---
title: "TCGA-HNSC RT-treated: K-means on the 5-gene CT panel + survival"
output: html_document
---

```{r setup, include=FALSE}
# K-means (k = 2) clustering of the RT-treated TCGA-HNSC subset on the
# 5-gene CT panel (RIBC2, TCAM1P, SMC1B, STAG3, SYCP2), followed by
# Kaplan-Meier survival analysis (OS, PFS, DSS) stratified on HPV.
source(here::here("config.R"))
knitr::opts_chunk$set(fig.width = 14, fig.height = 8)
```


##library
```{r include=FALSE}
library(DESeq2)

library(pheatmap)
library(dplyr)
library(Rlabkey)
library(tidyverse)
library(readxl)
library(TCGAbiolinks)
library(biomaRt)
library(SummarizedExperiment)
library(stringr)
library(tidyr)
library(limma)
```


###  only the counts results of significant genes from enrichment analysis

```{r }
sig <- read.csv(file.path(TCGA_TX_DATA, "Countsprimarytumormodified.csv"))
sig = aggregate(sig,FUN = max,by=list(sig$hgnc_symbol))
sig <- sig[,-1]
```




```{r}
read.csv(file.path(TCGA_TX_DATA, "TCGA_metadata_GDC_RPB_combined.csv"), row.names = 1) -> tcga_hnscc_all
```

# only radiation patients (both HPV+ and HPV-)
```{r}


# Helper to pull off the leading 0/1 from a "0:alive" or "1:dead" string
parse_status <- function(x) {
  ifelse(grepl("^[01]:", x),
         as.numeric(sub(":.*", "", x)),
         NA_real_)
}

LTC_tcga_HPV_pos <- tcga_hnscc_all %>%
  # 1) Convert numeric-month columns
  mutate(across(
    c(
      `Overall.Survival..Months.`,
      `Progress.Free.Survival..Months.`,
      `Months.of.disease.specific.survival`
    ),
    as.numeric
  )) %>%
  # 2) Convert status columns all with the same helper
  mutate(across(
    c(
      `Overall.Survival.Status`,
      `Progression.Free.Status`,
      `Disease.specific.Survival.status`
    ),
    parse_status
  )) %>%
  # 3) Filter for radiation therapy patients (both HPV+ and HPV-)
  filter(`Radiation.Therapy` == "Yes")

# Now you can plot, summarise, etc.
hist(LTC_tcga_HPV_pos$`Overall.Survival..Months.`)



```




```{r}
# Add a pseudo-count of 1 to all counts
sig <- sig %>%
  mutate(across(-1, ~ . + 1))
# Make the first column the row names
rownames(sig) <- sig$hgnc_symbol

# Remove the first column (optional)
sig <- sig[, -1]


# Now proceed with the variance stabilizing transformation and heatmap
vsdsig_tcga <- varianceStabilizingTransformation(object = as.matrix(sig))
expression_data_tcga_t <- t(vsdsig_tcga)
```

# Select radiation patients (already filtered above)
```{r}
# Check which patients are missing
patients_in_clinical <- rownames(LTC_tcga_HPV_pos)
samples_in_expression <- rownames(expression_data_tcga_t)

cat("Patients in clinical data (LTC_tcga_HPV_pos):", length(patients_in_clinical), "\n")
cat("Samples in expression data (expression_data_tcga_t):", length(samples_in_expression), "\n\n")

# Find common patients
common_samples <- intersect(patients_in_clinical, samples_in_expression)
cat("Common samples in both datasets:", length(common_samples), "\n\n")

# Keep only common samples in both datasets
LTC_tcga_HPV_pos <- LTC_tcga_HPV_pos[rownames(LTC_tcga_HPV_pos) %in% common_samples, ]

expression_data_tcga_t_radiation <- expression_data_tcga_t[
  rownames(expression_data_tcga_t) %in% common_samples,
  , drop = FALSE
]

cat("Final dimensions:\n")
cat("  LTC_tcga_HPV_pos:", nrow(LTC_tcga_HPV_pos), "rows\n")
cat("  expression_data_tcga_t_radiation:", nrow(expression_data_tcga_t_radiation), "rows\n")

```


```{r}
final_gene <- c("RIBC2", "TCAM1P", "SMC1B", "STAG3", "SYCP2")

# Scale expression data for the specified genes
scaled_patients <- scale(expression_data_tcga_t_radiation[, final_gene], center = TRUE, scale = TRUE)
# Verify scaling (variances should be ~1)
apply(scaled_patients, 2, var)
```

```{r}
# Define a k-means clustering function
kmeans_clustering <- function(data) {
  set.seed(42)  # For reproducibility
  kmeans_result <- kmeans(data, centers = 3, nstart = 100, iter.max = 10, algorithm = "Hartigan-Wong")
  cluster_means <- tapply(rowMeans(data), kmeans_result$cluster, mean)
  
  # Determine which cluster corresponds to high, medium, and low expression
  if (cluster_means[1] > cluster_means[2] && cluster_means[1] > cluster_means[3]) {
    high <- 1; medium <- ifelse(cluster_means[2] > cluster_means[3], 2, 3); low <- ifelse(cluster_means[2] > cluster_means[3], 3, 2)
  } else if (cluster_means[2] > cluster_means[1] && cluster_means[2] > cluster_means[3]) {
    high <- 2; medium <- ifelse(cluster_means[1] > cluster_means[3], 1, 3); low <- ifelse(cluster_means[1] > cluster_means[3], 3, 1)
  } else {
    high <- 3; medium <- ifelse(cluster_means[1] > cluster_means[2], 1, 2); low <- ifelse(cluster_means[1] > cluster_means[2], 2, 1)
  }
  
  # Replace factor levels with descriptive labels
  cluster <- as.factor(kmeans_result$cluster)
  levels(cluster)[high] <- "High CT Panel Expression"
  levels(cluster)[medium] <- "Intermediate CT Panel Expression"
  levels(cluster)[low] <- "Low CT Panel Expression"
  
  return(list(cluster_labels = cluster, centroids = kmeans_result$centers))
}

# Perform clustering and add the results to expression_data_tcga_t_radiation
clustering_results <- kmeans_clustering(scaled_patients[, final_gene])
expression_data_tcga_t_radiation <- as.data.frame(expression_data_tcga_t_radiation)
expression_data_tcga_t_radiation <- tibble::rownames_to_column(expression_data_tcga_t_radiation, "Patient_ID")
expression_data_tcga_t_radiation$DEG.kmeans <- clustering_results$cluster_labels

# Check overall mean expression per cluster
group_means <- aggregate(rowMeans(scaled_patients), 
                         by = list(expression_data_tcga_t_radiation$DEG.kmeans), 
                         mean)
colnames(group_means) <- c("Group", "Overall_Mean_Expression")
print(group_means)
```

```{r}
table(expression_data_tcga_t_radiation$DEG.kmeans)
```

```{r}
# Prepare data for plotting by merging scaled values and cluster labels
cluster <- cbind(scaled_patients[, final_gene], expression_data_tcga_t_radiation[, c("DEG.kmeans", "Patient_ID")])
dfm <- reshape2::melt(cluster, variable.name = "Gene", value.name = "Expression")

# Convert DEG.kmeans to ordered factor for proper display order
dfm$DEG.kmeans <- factor(dfm$DEG.kmeans, 
                         levels = c("High CT Panel Expression", 
                                   "Intermediate CT Panel Expression", 
                                   "Low CT Panel Expression"),
                         ordered = TRUE)

dfm_summary <- dfm %>% 
  dplyr::group_by(Gene, DEG.kmeans) %>% 
  dplyr::summarize(n = n(),
                   Mean = mean(Expression), 
                   sd = sd(Expression),
                   se = sd / sqrt(n),
                   .groups = "drop")

# Plot the mean expression values with error bars, faceted by k-means cluster
p <- ggplot(dfm_summary, aes(x = Gene, y = Mean, fill = DEG.kmeans)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = Mean - se, ymax = Mean + se),
                width = 0.2, position = position_dodge(0.9)) +
  scale_fill_manual(values = c("High CT Panel Expression" = "#7AB2D3", 
                               "Intermediate CT Panel Expression" = "#E7B800", 
                               "Low CT Panel Expression" = "#DE3163")) +
  coord_flip() +
  facet_grid(rows = vars(DEG.kmeans)) +
  theme_bw() +
  theme(axis.text = element_text(size = 10, face = "bold.italic"),
        axis.text.x = element_text(size = 10),
        axis.title = element_text(size = 10, face = "bold"),
        legend.position = "none",
        strip.text = element_text(size = 10),
        plot.title = element_text(size = 10, hjust = 0.5, face = "bold")) +
  ggtitle("Mean expression value of patients clustered \n into 'High/Intermediate/Low CT Panel Expression' groups")

# Save to PDF
#pdf(file.path(TCGA_KM_OUT, "kmeans_plot.pdf"), width = 3, height = 5)
print(p)
#dev.off()
```

# Optional testing

```{r}
library(cluster)
library(ggplot2)

# 1. Run kmeans clustering for k = 3
# Coerce cluster factor into numeric for metrics
# Coerce factor labels to numeric (for silhouette)
cluster_numeric <- as.numeric(expression_data_tcga_t_radiation$DEG.kmeans)
# Reuse your scaled matrix
sil <- silhouette(cluster_numeric, dist(scaled_patients[, final_gene]))
s <- plot(sil, main = "Silhouette Plot for K-means Clusters (k = 3)")
cat("Average silhouette width:", mean(sil[, 3]), "\n")
pdf(file.path(TCGA_KM_OUT, "kmeans_tcga_radiotherapy_silhouette.pdf"), width = 3, height = 5)
print(s)


```


```{r}
library(cluster)

# Coerce cluster factor into numeric for metrics
cluster_numeric <- as.numeric(clustering_results$cluster_labels)

# 1. Silhouette analysis
sil <- silhouette(cluster_numeric, dist(scaled_patients[, final_gene]))
plot(sil, main = "Silhouette Plot for K-means Clusters")
cat("Average silhouette width:", mean(sil[, 3]), "\n")
```


# K-means clustering with k = 2

```{r}
# --- K-means clustering function for k = 2 with expression-based labeling
kmeans_clustering_k2 <- function(data) {
  set.seed(42)
  kmeans_result <- kmeans(data, centers = 2, nstart = 100, iter.max = 10, algorithm = "Hartigan-Wong")
  cluster_means <- tapply(rowMeans(data), kmeans_result$cluster, mean)

  # Assign labels based on cluster mean expression
  if (cluster_means[1] > cluster_means[2]) {
    high <- 1; low <- 2
  } else {
    high <- 2; low <- 1
  }

  cluster <- as.factor(kmeans_result$cluster)
  levels(cluster)[high] <- "High CT Panel Expression"
  levels(cluster)[low]  <- "Low CT Panel Expression"

  return(list(cluster_labels = cluster, centroids = kmeans_result$centers))
}

# --- Run clustering
clustering_k2 <- kmeans_clustering_k2(scaled_patients)
clusters_k2 <- clustering_k2$cluster_labels
print(table(clusters_k2))

```

```{r}
library(cluster)
cluster_numeric_k2 <- as.numeric(clusters_k2)
sil_k2 <- silhouette(cluster_numeric_k2, dist(scaled_patients[, final_gene]))
plot(sil_k2, main = "Silhouette Plot for K-means Clusters (k = 2)")
cat("Average silhouette width (k = 2):", mean(sil_k2[, 3]), "\n")

```








#do a survival analysis

```{r}

# Use the Patient_ID from expression data and clusters from k=2 clustering
clusters_df <- data.frame(Patient_ID = expression_data_tcga_t_radiation$Patient_ID, 
                          Cluster = clusters_k2)

# Merge with clinical data using Patient_ID
merged_df <- merge(clusters_df, 
                   LTC_tcga_HPV_pos,
                   by.x = "Patient_ID", 
                   by.y = 0)

```

```{r}
library(survival)
library(survminer)

# Create survival object
surv_object <- Surv(time = merged_df$`Overall.Survival..Months.`, 
                    event = merged_df$`Overall.Survival.Status`)

# Fit Kaplan-Meier curve
fit <- survfit(surv_object ~ Cluster, data = merged_df)

# Plot survival curves
OS <- ggsurvplot(fit, data = merged_df, 
           pval = TRUE,                      # Display p-value of the log-rank test
           risk.table = TRUE,                # Add a table of the number of subjects at risk at different times
           conf.int = TRUE,                  # Show confidence intervals
           palette = c("#7AB2D3", "#DE3163"), # Colors for the groups (High, Low)
           xlab = "Time (months)",             # Customize X-axis label
           ylab = "Survival probability",    # Customize Y-axis label
           legend.labs = c("High CT Panel Expression", "Low CT Panel Expression"))  # Labels for the legend

print(OS)
```

#for poster - save to PDF
```{r}
# Save the plot to PDF
pdf(file.path(TCGA_KM_OUT, "tcga_OS_kmeans.pdf"), height = 6, width = 8, onefile = F)
print(OS)
dev.off()
```


# progression free

```{r}

# Create survival object
surv_object <- Surv(time = merged_df$`Progress.Free.Survival..Months.`, event = merged_df$`Progression.Free.Status`)

# Fit Kaplan-Meier curve
fit <- survfit(surv_object ~ Cluster, data = merged_df)

# Plot survival curves
ggsurvplot(fit, data = merged_df, 
           pval = TRUE,                      # Display p-value of the log-rank test
           risk.table = TRUE,                # Add a table of the number of subjects at risk at different times
           conf.int = TRUE,                  # Show confidence intervals
           palette = c("#7AB2D3", "#DE3163"), # Colors for the groups
           xlab = "Time (months)",             # Customize X-axis label
           ylab = "Progression Free Survival probability",    # Customize Y-axis label
           legend.title = "HPV Status",
           legend.labs = c("High", "Low"))  # Labels for the legend
```


# disease speciific

```{r}
# Create survival object
surv_object <- Surv(time = merged_df$`Months.of.disease.specific.survival`, event = merged_df$`Disease.specific.Survival.status`)

# Fit Kaplan-Meier curve
fit <- survfit(surv_object ~ Cluster, data = merged_df)

# Plot survival curves
ggsurvplot(fit, data = merged_df, 
           pval = TRUE,                      # Display p-value of the log-rank test
           risk.table = TRUE,                # Add a table of the number of subjects at risk at different times
           conf.int = TRUE,                  # Show confidence intervals
           palette = c("#7AB2D3", "#DE3163"), # Colors for the groups
           xlab = "Time (months)",             # Customize X-axis label
           ylab = "Disease Specific Survival probability",    # Customize Y-axis label
           legend.title = "HPV Status",
           legend.labs = c("High", "Low"))  # Labels for the legend
```


#### Heatmap with 5-gene panel and k-means clustering

```{r}
library(pheatmap)

# Define the 5-gene panel
gene_panel <- c("RIBC2", "TCAM1P", "SMC1B", "STAG3", "SYCP2")

# Get expression data for these genes and set rownames to Patient_ID
heatmap_data <- expression_data_tcga_t_radiation[, gene_panel]
rownames(heatmap_data) <- merged_df$Patient_ID

# Prepare annotations from merged_df (only Cluster and Subtype, no Patient_ID)
heatmap_anno <- merged_df[, c("Cluster", "Subtype")]
rownames(heatmap_anno) <- merged_df$Patient_ID

# Convert to character for consistency
heatmap_anno$Cluster <- as.character(heatmap_anno$Cluster)
heatmap_anno$Subtype <- as.character(heatmap_anno$Subtype)

# Create annotation colors
anno_colors <- list(
  Cluster = c("High CT Panel Expression" = "#7AB2D3", 
              "Low CT Panel Expression" = "#DE3163"),
  Subtype = c("HNSC_HPV+" = "#FF6B6B", 
              "HNSC_HPV-" = "#4ECDC4")
)

# Create the heatmap

p <- pheatmap(t(heatmap_data),
         annotation_col = heatmap_anno,
         annotation_colors = anno_colors,
         scale = "row",
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         show_rownames = TRUE,
         show_colnames = FALSE,
         main = "5-Gene Panel Expression Heatmap",
         fontsize = 10)
print(p)
```

```{r}
# Create the heatmap
pdf(file.path(TCGA_KM_OUT, "tcga_heatmap_5gene_panel.pdf"), width = 12, height = 8)
pheatmap(t(heatmap_data),
         annotation_col = heatmap_anno,
         annotation_colors = anno_colors,
         scale = "row",
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         show_rownames = TRUE,
         show_colnames = FALSE,
         main = "5-Gene Panel Expression Heatmap",
         fontsize = 10)
#dev.off()
```
#### Survival analysis - HPV+ patients only

```{r}
# Filter for HPV+ patients only
hpv_pos_df <- merged_df %>%
  filter(Subtype == "HNSC_HPV+")

# Create survival object for HPV+ patients
surv_object_hpv <- Surv(time = hpv_pos_df$`Overall.Survival..Months.`, 
                        event = hpv_pos_df$`Overall.Survival.Status`)

# Fit Kaplan-Meier curve
fit_hpv <- survfit(surv_object_hpv ~ Cluster, data = hpv_pos_df)

# Plot survival curves for HPV+ only
OS_hpv <- ggsurvplot(fit_hpv, data = hpv_pos_df, 
           pval = TRUE,
           risk.table = TRUE,
           conf.int = FALSE,
           palette = c("#7AB2D3", "#DE3163"),
           xlab = "Time (months)",
           ylab = "Survival probability",
           legend.labs = c("High CT Panel Expression", "Low CT Panel Expression"))

print(OS_hpv)
```

#for poster - save to PDF
```{r}
# Save the HPV+ survival plot to PDF
pdf(file.path(TCGA_KM_OUT, "tcga_OS_kmeans_HPV_positive.pdf"), height = 6, width = 8, onefile = F)
print(OS_hpv)
#dev.off()
```

#### Progression Free Survival - HPV+ patients only

```{r}
# Create survival object for PFS in HPV+ patients
surv_object_pfs_hpv <- Surv(time = hpv_pos_df$`Progress.Free.Survival..Months.`, 
                             event = hpv_pos_df$`Progression.Free.Status`)

# Fit Kaplan-Meier curve
fit_pfs_hpv <- survfit(surv_object_pfs_hpv ~ Cluster, data = hpv_pos_df)

# Plot PFS curves for HPV+ only
PFS_hpv <- ggsurvplot(fit_pfs_hpv, data = hpv_pos_df, 
           pval = TRUE,
           risk.table = TRUE,
           conf.int = FALSE,
           palette = c("#7AB2D3", "#DE3163"),
           xlab = "Time (months)",
           ylab = "Progression Free Survival probability",
           legend.labs = c("High CT Panel Expression", "Low CT Panel Expression"))

print(PFS_hpv)
```

```{r}
# Save the HPV+ PFS plot to PDF
pdf(file.path(TCGA_KM_OUT, "tcga_PFS_kmeans_HPV_positive.pdf"), height = 6, width = 8, onefile = F)
print(PFS_hpv)
dev.off()
```

#### Disease Specific Survival - HPV+ patients only

```{r}
# Create survival object for DSS in HPV+ patients
surv_object_dss_hpv <- Surv(time = hpv_pos_df$`Months.of.disease.specific.survival`, 
                             event = hpv_pos_df$`Disease.specific.Survival.status`)

# Fit Kaplan-Meier curve
fit_dss_hpv <- survfit(surv_object_dss_hpv ~ Cluster, data = hpv_pos_df)

# Plot DSS curves for HPV+ only
DSS_hpv <- ggsurvplot(fit_dss_hpv, data = hpv_pos_df, 
           pval = TRUE,
           risk.table = TRUE,
           conf.int = FALSE,
           palette = c("#7AB2D3", "#DE3163"),
           xlab = "Time (months)",
           ylab = "Disease Specific Survival probability",
           legend.labs = c("High CT Panel Expression", "Low CT Panel Expression"))

print(DSS_hpv)
```

```{r}
# Save the HPV+ DSS plot to PDF
pdf(file.path(TCGA_KM_OUT, "tcga_DSS_kmeans_HPV_positive.pdf"), height = 6, width = 8, onefile = F)
print(DSS_hpv)
dev.off()
```

#### Faceted Survival Plot with Risk Tables - HPV+ Patients Only

```{r}
# Add survival type label to HPV+ patients data
hpv_os <- hpv_pos_df %>%
  mutate(Detection = "Overall Survival",
         Time = `Overall.Survival..Months.`,
         Status = `Overall.Survival.Status`)

hpv_pfs <- hpv_pos_df %>%
  mutate(Detection = "Progression-Free Survival",
         Time = `Progress.Free.Survival..Months.`,
         Status = `Progression.Free.Status`)

hpv_dss <- hpv_pos_df %>%
  mutate(Detection = "Disease-Specific Survival",
         Time = `Months.of.disease.specific.survival`,
         Status = `Disease.specific.Survival.status`)

# Combine all into one dataset for HPV+ patients
cohort_combined <- bind_rows(hpv_os, hpv_pfs, hpv_dss)
cohort_combined$Detection <- factor(cohort_combined$Detection,
                                    levels = c("Overall Survival", "Progression-Free Survival", "Disease-Specific Survival"))
```

```{r}
# Fit survival models for all three types in HPV+ patients
fit_all <- survfit(Surv(Time, Status) ~ Cluster, data = cohort_combined)

# Create faceted survival plot
facet_plot <- ggsurvplot_facet(
  fit = fit_all,
  data = cohort_combined,
  facet.by = "Detection",
  palette = c("#7AB2D3", "#DE3163"),
  conf.int = FALSE,
  pval = TRUE,
  risk.table = TRUE,
  legend.title = "Groups",
  legend.labs = c("High CT Panel Expression", "Low CT Panel Expression"),
  xlab = "Months after diagnosis",
  ylab = "Survival Probability",
  short.panel.labs = TRUE,
  scales = "fixed"
) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold"),
    plot.margin = margin(10, 30, 10, 10)
  )

print(facet_plot)

# Save to PDF
ggsave(file.path(TCGA_KM_OUT, "TCGA_survival_CT_genes_faceted_HPV_positive.pdf"), facet_plot, width = 15, height = 6)
```
