---
title: "AllColumns_HPV"
output: html_document
date: "2023-07-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### This analysis is performed including all samples for which we also have methylation data, additionally excluding FaDu #86 (which has a very different tumor volume as well). Samples excluded: cal33 #248, fadu #127, fadu #268       
##loadpackage
```{r}
library(ashr) 
library(tidyverse)
library(ggplot2)
library(readxl)
library(pheatmap)
library(tibble)
library(dplyr)
```


## 3.2. Load individual .tsv files for each sample

```{r}



source(here::here("config.R"))  # defines DATA_DIR, RESULTS_DIR, ANALYSIS_SUBTYPES_DIR
FaDu.N150a.14rep <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_FaDu_014_rep/depl-wholetumor01-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor01-ffpe-x_HS_FaDu_014_rep.fpkm_tpm.featureCounts.tsv"))

FaDu.N150a.20 <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_FaDu_20/depl-wholetumor02-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor02-ffpe-x_HS_FaDu_20.fpkm_tpm.featureCounts.tsv"))

FaDu.N150d.320 <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_FaDu_320/depl-wholetumor03-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor03-ffpe-x_HS_FaDu_320.fpkm_tpm.featureCounts.tsv"))

UT5.N153a.83 <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_UT5_083/depl-wholetumor01-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor01-ffpe-x_HS_UT5_083.fpkm_tpm.featureCounts.tsv"))

UT5.N153a.84 <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_UT5_084/depl-wholetumor02-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor02-ffpe-x_HS_UT5_084.fpkm_tpm.featureCounts.tsv"))

UT5.N153b.219 <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_UT5_219/depl-wholetumor03-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor03-ffpe-x_HS_UT5_219.fpkm_tpm.featureCounts.tsv"))

Cal33.N154a.37rep <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_Cal33_037_rep/depl-wholetumor01-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor01-ffpe-x_HS_Cal33_037_rep.fpkm_tpm.featureCounts.tsv"))

Cal33.N154a.73 <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_Cal33_073/depl-wholetumor02-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor02-ffpe-x_HS_Cal33_073.fpkm_tpm.featureCounts.tsv"))

Cal33.N154a.38 <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_Cal33_038/depl-wholetumor03-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor03-ffpe-x_HS_Cal33_038.fpkm_tpm.featureCounts.tsv"))

UT8.N155a.131 <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_UT8_131/depl-wholetumor01-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor01-ffpe-x_HS_UT8_131.fpkm_tpm.featureCounts.tsv"))

UT8.N155b.199 <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_UT8_199/depl-wholetumor02-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor02-ffpe-x_HS_UT8_199.fpkm_tpm.featureCounts.tsv"))

UT8.N155b.200 <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_UT8_200/depl-wholetumor03-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor03-ffpe-x_HS_UT8_200.fpkm_tpm.featureCounts.tsv"))

SAS.N156a.74 <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_SAS_74/depl-wholetumor01-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor01-ffpe-x_HS_SAS_74.fpkm_tpm.featureCounts.tsv"))

SAS.N156b.140rep <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_SAS_140_rep/depl-wholetumor02-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor02-ffpe-x_HS_SAS_140_rep.fpkm_tpm.featureCounts.tsv"))

SAS.N156b.181 <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_SAS_181/depl-wholetumor03-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor03-ffpe-x_HS_SAS_181.fpkm_tpm.featureCounts.tsv"))

UT45.N157b.120 <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_UT45_120/depl-wholetumor01-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor01-ffpe-x_HS_UT45_120.fpkm_tpm.featureCounts.tsv"))

UT45.N157b.123 <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_UT45_123/depl-wholetumor02-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor02-ffpe-x_HS_UT45_123.fpkm_tpm.featureCounts.tsv"))

SAT.N165a.67rep <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_SAT_067_rep/depl-wholetumor01-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor01-ffpe-x_HS_SAT_067_rep.fpkm_tpm.featureCounts.tsv"))

SAT.N165b.149rep <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_SAT_149_rep/depl-wholetumor02-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor02-ffpe-x_HS_SAT_149_rep.fpkm_tpm.featureCounts.tsv"))

SAT.N165a.2 <- read_tsv(file = file.path(DATA_DIR, "rna_sequencing/view-by-pid/HS_SAT_02/depl-wholetumor03-ffpe-x/paired/merged-alignment/featureCounts/depl-wholetumor03-ffpe-x_HS_SAT_02.fpkm_tpm.featureCounts.tsv"))


```




#Create counts file
```{r}

Counts_FaDu.N178a.68 <- FaDu.N178a.68 %>% dplyr::select(num_reads)
colnames(Counts_FaDu.N178a.68) <- c("FaDu.N178a.68")

Counts_FaDu.N178a.72 <- FaDu.N178a.72 %>% dplyr::select(num_reads)
colnames(Counts_FaDu.N178a.72) <- c("FaDu.N178a.72")

Counts_FaDu.N178a.73 <- FaDu.N178a.73 %>% dplyr::select(num_reads)
colnames(Counts_FaDu.N178a.73) <- c("FaDu.N178a.73")

Counts_FaDu.N178a.74 <- FaDu.N178a.74 %>% dplyr::select(num_reads)
colnames(Counts_FaDu.N178a.74) <- c("FaDu.N178a.74")

Counts_FaDu.N178a.75 <- FaDu.N178a.75 %>% dplyr::select(num_reads)
colnames(Counts_FaDu.N178a.75) <- c("FaDu.N178a.75")

Counts_SAS.N180b.96 <- SAS.N180b.96 %>% dplyr::select(num_reads)
colnames(Counts_SAS.N180b.96) <- c("SAS.N180b.96")

Counts_SAS.N180b.97 <- SAS.N180b.97 %>% dplyr::select(num_reads)
colnames(Counts_SAS.N180b.97) <- c("SAS.N180b.97")

Counts_SAS.N180b.98 <- SAS.N180b.98 %>% dplyr::select(num_reads)
colnames(Counts_SAS.N180b.98) <- c("SAS.N180b.98")

Counts_SAS.N180b.105 <- SAS.N180b.105 %>% dplyr::select(num_reads)
colnames(Counts_SAS.N180b.105) <- c("SAS.N180b.105")

Counts_SAS.N180b.106 <- SAS.N180b.106 %>% dplyr::select(num_reads)
colnames(Counts_SAS.N180b.106) <- c("SAS.N180b.106")

Counts_SAS.N180b.107 <- SAS.N180b.107 %>% dplyr::select(num_reads)
colnames(Counts_SAS.N180b.107) <- c("SAS.N180b.107")

Counts_UDSCC2.N174b.124 <- UDSCC2.N174b.124 %>% dplyr::select(num_reads)
colnames(Counts_UDSCC2.N174b.124) <- c("UDSCC2.N174b.124")

Counts_UDSCC2.N174b.161 <- UDSCC2.N174b.161 %>% dplyr::select(num_reads)
colnames(Counts_UDSCC2.N174b.161) <- c("UDSCC2.N174b.161")

Counts_UDSCC2.N174b.167 <- UDSCC2.N174b.167 %>% dplyr::select(num_reads)
colnames(Counts_UDSCC2.N174b.167) <- c("UDSCC2.N174b.167")

Counts_UPC154.N189b.102 <- UPC154.N189b.102 %>% dplyr::select(num_reads)
colnames(Counts_UPC154.N189b.102) <- c("UPC154.N189b.102")

Counts_UPC154.N189b.103 <- UPC154.N189b.103 %>% dplyr::select(num_reads)
colnames(Counts_UPC154.N189b.103) <- c("UPC154.N189b.103")

Counts_UPC154.N189b.173 <- UPC154.N189b.173 %>% dplyr::select(num_reads)
colnames(Counts_UPC154.N189b.173) <- c("UPC154.N189b.173")

Counts_UPC154.N189b.174 <- UPC154.N189b.174 %>% dplyr::select(num_reads)
colnames(Counts_UPC154.N189b.174) <- c("UPC154.N189b.174")

Counts_UDSCC2.N174b.122 <- UDSCC2.N174b.122 %>% dplyr::select(num_reads)
colnames(Counts_UDSCC2.N174b.122) <- c("UDSCC2.N174b.122")

Counts_UDSCC2.N174b.123 <- UDSCC2.N174b.123 %>% dplyr::select(num_reads)
colnames(Counts_UDSCC2.N174b.123) <- c("UDSCC2.N174b.123")

Counts_UDSCC2.N174b.169 <- UDSCC2.N174b.169 %>% dplyr::select(num_reads)
colnames(Counts_UDSCC2.N174b.169) <- c("UDSCC2.N174b.169")


Counts_UPC154.N189b.105 <- UPC154.N189b.105 %>% dplyr::select(num_reads)
colnames(Counts_UPC154.N189b.105) <- c("UPC154.N189b.105")

Counts_UPC154.N189b.175 <- UPC154.N189b.175 %>% dplyr::select(num_reads)
colnames(Counts_UPC154.N189b.175) <- c("UPC154.N189b.175")

```


```{r}
Counts <- bind_cols(
  Counts_FaDu.N178a.68,
  Counts_FaDu.N178a.72,
  Counts_FaDu.N178a.73,
  Counts_FaDu.N178a.74,
  Counts_FaDu.N178a.75,
  Counts_SAS.N180b.96,
  Counts_SAS.N180b.97,
  Counts_SAS.N180b.98,
  Counts_SAS.N180b.105,
  Counts_SAS.N180b.106,
  Counts_SAS.N180b.107,
  Counts_UDSCC2.N174b.122,
  Counts_UDSCC2.N174b.123,
  Counts_UDSCC2.N174b.124,
  Counts_UDSCC2.N174b.161,
  Counts_UDSCC2.N174b.167,
  Counts_UDSCC2.N174b.169,
  Counts_UPC154.N189b.102,
  Counts_UPC154.N189b.103,
  Counts_UPC154.N189b.105,
  Counts_UPC154.N189b.173,
  Counts_UPC154.N189b.174,
  Counts_UPC154.N189b.175,
)

```



```{r}
gene_names <- FaDu.N178a.68 %>% dplyr::select(name)
head(gene_names)
```


```{r}
counts_v2 <- cbind(gene_names, Counts)
head(counts_v2)
```

```{r}
#if we a trget max of counts
counts_v2 = aggregate(counts_v2,FUN = max,by=list(counts_v2$name))
counts_v2 <- counts_v2[,-1]

rownames(counts_v2) <- counts_v2$name
head(counts_v2)
```


```{r}
counts_df <- as.data.frame(counts_v2)
head(counts_df)
counts_df <- counts_df %>% rename(Genes=name)
```

## 3.3. save counts file as csv 

```{r}
write_csv(counts_df, "countall.csv")
```

##save TPM files as csv
```{r}
TPM_v2 <- cbind(gene_names, TPM)
head(TPM)
```

```{r}
#if we a trget max of counts
TPM_v2 = aggregate(TPM_v2,FUN = max,by=list(TPM_v2$name))
TPM_v2 <- TPM_v2[,-1]

rownames(TPM_v2) <- TPM_v2$name
head(TPM)
```


```{r}
TPM_df <- as.data.frame(TPM_v2)
head(TPM_df)
```

## 3.3. save counts file as tsv 

```{r}
write_csv(TPM_df, "TPM_all_byR.csv")
```


























#### since we already saw that the cell lines cluster independently but also according to Nimorazole response (FaDu and SAS = responders, UT-SCC-45, SAT and Cal33 = Non-responders), we will combine FaDu and SAS samples in one group (responder) and SAT, cal33 and UT-SCC-45 in another group (non_responder)

#### Perform contrasts followed by Wald test (instead of doing pairwise comparison and normalization each time):

#### 1) responder vs non_responder --> CONTROL
#### 2) responder vs non_responder --> RCT
#### 3) responder vs non_responder --> RCT + NIMO

```{r echo=TRUE}
counts_all <- as.matrix(counts_tsv %>% dplyr::select(cal33_184, 
cal33_37, 
cal33_73, 
cal33_98, 
cal33_104, 
cal33_21,      
cal33_22,       
cal33_259,     
cal33_107,    
cal33_149,     
cal33_41,  
cal33_42,  
sat_149,         
sat_2,             
sat_201,         
sat_85,           
sat_107,        
sat_168, 
sat_82, 
sat_83, 
sat_133,      
sat_19,          
sat_26,      
sat_73,       
utscc45_11,
utscc45_119, 
utscc45_12,   
utscc45_16,   
utscc45_1,    
utscc45_2,
utscc45_231, 
utscc45_57, 
utscc45_58,
utscc45_183, 
utscc45_209, 
utscc45_228, 
utscc45_5,
fadu_125,       
fadu_270,       
fadu_320,       
fadu_100,
fadu_23,         
fadu_24,        
fadu_13,  
fadu_165,  
fadu_324,    
fadu_7,        
sas_140,         
sas_181,        
sas_74,           
sas_82,          
sas_10,
sas_207,         
sas_218,         
sas_31,           
sas_120,
sas_192,
sas_30,
sas_55,
sas_66,  
))
head(counts_all)
```

```{r echo=TRUE}
rownames(counts_all) <- counts_tsv$name
head(counts_all)
```

#### Create sample annotation 

```{r warning=FALSE}
samples <- c("cal33_184", 
"cal33_37", 
"cal33_73", 
"cal33_98", 
"cal33_104", 
"cal33_21",      
"cal33_22",       
"cal33_259",     
"cal33_107",    
"cal33_149",     
"cal33_41",  
"cal33_42",
"sat_149",         
"sat_2",             
"sat_201",         
"sat_85",           
"sat_107",        
"sat_168", 
"sat_82", 
"sat_83", 
"sat_133",      
"sat_19",          
"sat_26",      
"sat_73",       
"utscc45_11",
"utscc45_119", 
"utscc45_12",   
"utscc45_16",   
"utscc45_1",    
"utscc45_2",
"utscc45_231", 
"utscc45_57", 
"utscc45_58",
"utscc45_183", 
"utscc45_209", 
"utscc45_228", 
"utscc45_5",
"fadu_125",       
"fadu_270",       
"fadu_320",       
"fadu_100",
"fadu_23",         
"fadu_24",        
"fadu_13",  
"fadu_165",  
"fadu_324",    
"fadu_7",        
"sas_140",         
"sas_181",        
"sas_74",           
"sas_82",          
"sas_10",
"sas_207",         
"sas_218",         
"sas_31",           
"sas_120",
"sas_192",
"sas_30",
"sas_55",
"sas_66")

cell_line <- c(rep("cal33", 12), rep("sat", 12), rep("utscc45", 13), rep("fadu", 10), rep("sas", 13))

condition <- c(rep("control",4), rep("RCT", 4), rep("RCT-Nimo", 4), rep("control", 4), rep("RCT", 4), rep("RCT-Nimo", 4), rep("control", 4), rep("RCT", 5), rep("RCT-Nimo", 4), rep("control", 3), rep("RCT", 3), rep("RCT-Nimo", 4), rep("control", 4), rep("RCT", 4), rep("RCT-Nimo", 5))

response <- c(rep("non_responder", 37), rep("responder", 23))

#we create this merged column so then it will be easier to get contrasts done per nimo response and condition

response_condition <- c(rep("non_resp_control", 4), rep("non_resp_RCT", 4), rep("non_resp_RCT-Nimo", 4), rep("non_resp_control", 4), rep("non_resp_RCT", 4), rep("non_resp_RCT-Nimo", 4), rep("non_resp_control", 4), rep("non_resp_RCT", 5), rep("non_resp_RCT-Nimo", 4), rep("resp_control", 3), rep("resp_RCT", 3), rep("resp_RCT-Nimo", 4), rep("resp_control", 4), rep("resp_RCT", 4), rep("resp_RCT-Nimo", 5))

annotations <- as.data.frame(cbind(samples, cell_line, response, condition, response_condition))
rownames(annotations) <- annotations$samples
annotations

write.csv(annotations, file = "D:/RNAseq-xenografts-whole_tumors/results_31052023/annotation.csv")
```