---
title: "Beta values — extraction, variable CpG filtering, M-value conversion"
output: html_document
---

```{r setup, include=FALSE}
# Loads the preprocessed RnBeads object produced by 01_rnbeads_hpv.R,
# then extracts beta values, converts to M values, and plots heatmaps
# of the most variable CpG sites.
source(here::here("config.R"))
knitr::opts_chunk$set(echo = TRUE)
```



```{r, results='hide'}
library("BiocParallel")
library("GenomicFeatures")
library("RnBeads")
library("ggplot2")
library("reshape2")
library("grid")
library("gridExtra")
library("GOstats")
library("LOLA")
library("qvalue")
library('wordcloud')
library("hexbin")
library("RnBeads.hg38")
library("dplyr")
library(biomaRt)
```


```{r}
load(file.path(PRECLIN_METH_OUT, "rnb.set.PC.rda"))
rnb.options(identifiers.column="Sample")
```


```{r}
Genes <- meth(rnb.set.PC, row.names = T, type="genes")
Genes <- as.data.frame(Genes)
Promoters <- meth(rnb.set.PC, row.names = T, type="promoters")
Promoters <- as.data.frame(Promoters)
cpg_beta <- meth(rnb.set.PC, row.names=T)
#write.csv(cpg, "Beta_cpg_Untreated.csv", row.names=T)
```


#convert to gene symbols
```{r}

mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")

getGeneSymbols <- function(ensembl_ids) {
  gene_symbols <- getBM(
    attributes = c('ensembl_gene_id', 'hgnc_symbol'), 
    filters = 'ensembl_gene_id', 
    values = ensembl_ids, 
    mart = mart
  )
  return(gene_symbols)
}

```


```{r}
# Get the gene symbols or use Ensembl IDs if symbols are missing
genes_symbols_df <- getGeneSymbols(row.names(Genes))
# It's important to create a named vector to match by names not by position
gene_symbol_or_ensembl <- setNames(genes_symbols_df$hgnc_symbol, genes_symbols_df$ensembl_gene_id)

# Match and replace in Genes
Genes$Gene_Symbol <- gene_symbol_or_ensembl[ row.names(Genes) ]
Genes <- Genes %>%
  dplyr::select(Gene_Symbol, everything())

# Match and replace in Promoters
promoters_symbols_df <- getGeneSymbols(row.names(Promoters))
promoter_symbol_or_ensembl <- setNames(promoters_symbols_df$hgnc_symbol, promoters_symbols_df$ensembl_gene_id)

Promoters$Gene_Symbol <- promoter_symbol_or_ensembl[ row.names(Promoters) ]
Promoters <- Promoters %>%
  dplyr::select(Gene_Symbol, everything())
```

#some ensemble ids dont ahve symbols, so keeping their id in the symbol column instead of blank for future use
```{r}
# For Genes
Genes$Gene_Symbol <- ifelse(is.na(Genes$Gene_Symbol) | Genes$Gene_Symbol == "", row.names(Genes), Genes$Gene_Symbol)
any(duplicated(Genes$Group.1))
Genes = aggregate(Genes,FUN = mean,by=list(Genes$Gene_Symbol))
any(duplicated(Genes$Group.1))
Genes <- Genes[,-2]
Genes <- Genes %>% 
  dplyr::rename (Gene_Symbol=Group.1)
# For Promoters
Promoters$Gene_Symbol <- ifelse(is.na(Promoters$Gene_Symbol) | Promoters$Gene_Symbol == "", row.names(Promoters), Promoters$Gene_Symbol)
any(duplicated(Promoters$Gene_Symbol))
Promoters = aggregate(Promoters,FUN = mean,by=list(Promoters$Gene_Symbol))
any(duplicated(Promoters$Group.1))
Promoters <- Promoters[,-2]
Promoters <- Promoters %>% 
  dplyr::rename (Gene_Symbol=Group.1)

#write.csv(Genes, "Beta_gene_all_with_Symbols.csv", row.names = TRUE)
#write.csv(Promoters, "Beta_promoters_all_with_Symbols.csv", row.names = TRUE)

```



#separate files based on HPV status
```{r}
Genes_untreated <- Genes [,c(1:19)]
Genes_untreated_HPVneg <- Genes [,c(1:10)]
Genes_untreated_HPVpos <- Genes [, c(1,11:19)]

Promoters_untreated <- Promoters [,c(1:19)]
Promoters_untreated_HPVneg <- Promoters [,c(1:10)]
Promoters_untreated_HPVpos <- Promoters [, c(1,11:19)]

#write.csv(Genes_untreated, "Beta_gene_untreated_with_Symbols.csv", row.names = TRUE)
#write.csv(Genes_untreated_HPVneg, "Beta_gene_untreated_HPVneg_with_Symbols.csv", row.names = TRUE)
#write.csv(Genes_untreated_HPVpos, "Beta_gene_untreated_HPVpos_with_Symbols.csv", row.names = TRUE)

#write.csv(Promoters_untreated, "Beta_promoters_untreated_with_Symbols.csv", row.names = TRUE)
#write.csv(Promoters_untreated_HPVneg, "Beta_promoters_untreated_HPVneg_with_Symbols.csv", row.names = TRUE)
#write.csv(Promoters_untreated_HPVpos, "Beta_promoters_untreated_HPVpos_with_Symbols.csv", row.names = TRUE)

```


```{r}
cpg_beta <- as.data.frame(cpg_beta)
# Convert rownames to a column
cpg_beta <- cpg_beta %>% 
  tibble::rownames_to_column("row_id")

# Arrange (sort) by the new row_id column
cpg_beta_ordered <- cpg_beta %>% 
  arrange(row_id)
rownames(cpg_beta_ordered) <- cpg_beta_ordered[,1]  
cpg_beta_ordered <- cpg_beta_ordered[,-1]          


```

###Teresa code

#' Get x most variable CpGs based on a beta matrix.
#'
#' @param input_beta_matrix A matrix of beta values.
#' @param x Number of most variable CpGs to return
#'
#'
get_most_variable <- function(input_beta_matrix, x) {
  filtered_beta_matrix <- input_beta_matrix[names(head(sort(apply(X=input_beta_matrix, MARGIN=1, FUN=sd), decreasing=TRUE), x)), ]
  return(filtered_beta_matrix)
}

# Calculate the number of CpG islands that constitute the top 1%
top_1_percent_count <- ceiling(nrow(cpg_beta) * 0.01)

# Apply the get_most_variable function to select the top 1% most variable CpG islands
top_1_percent_variable_cpg <- get_most_variable(input_beta_matrix = cpg_beta, x = top_1_percent_count)

```{r}
#' Get x percent most variable CpGs based on a beta matrix.
#'
#' @param input_beta_matrix A dataframe of beta values.
#' @param x Percentage of most variable CpGs to return.
#'
get_most_variable <- function(input_beta_matrix, x) {
  # Ensure the input is treated as a dataframe
  if (is.matrix(input_beta_matrix)) {
    input_beta_matrix <- as.data.frame(input_beta_matrix)
  }
  
  # Calculate the number of rows to include based on the percentage x
  top_x_percent_count <- ceiling(nrow(input_beta_matrix) * (x / 100))
  
  # Calculate the standard deviation of each row (CpG)
  sd_values <- apply(input_beta_matrix, MARGIN=1, FUN=sd)
  
  # Sort standard deviations in decreasing order and get the top x percent names
  top_names <- names(sort(sd_values, decreasing=TRUE))[1:top_x_percent_count]
  
  # Filter the dataframe to include only the top x percent most variable CpGs
  filtered_beta_matrix <- input_beta_matrix[top_names, ]
  
  return(filtered_beta_matrix)
}

# Example usage:
# Calculate the number of CpG islands that constitute the top 3%
# Apply the get_most_variable function to select the top 3% most variable CpG islands
cpg_beta_ordered <- # (your dataframe should be defined or loaded here)
top_3_percent_variable_cpg <- get_most_variable(input_beta_matrix = cpg_beta_ordered, x = 3)


```

#conversion into M value

```{r}
# Assuming top_3_percent_variable_cpg is your data frame with beta values
epsilon <- 1e-10

# Function to convert beta to M values
convert_beta_to_M <- function(beta_value) {
  log2((beta_value + epsilon) / (1 - beta_value + epsilon))
}

# Apply the function to each value in the data frame
top_3_percent_variable_cpg_M <- apply(top_3_percent_variable_cpg, MARGIN = c(1, 2), FUN = convert_beta_to_M)
```



```{r}
library(pheatmap)

# Store the original column names
original_columns <- colnames(top_3_percent_variable_cpg_M)



new_order <- c(1:9, 19:27, 10:18, 28:length(original_columns))
# Reorder the dataframe columns
top_3_percent_variable_cpg_M_HPV <- top_3_percent_variable_cpg_M[, new_order]

beta_m <- pheatmap(scale(top_3_percent_variable_cpg_M_HPV), cluster_cols = F)
```

```{r}
top_3_percent_variable_cpg <- top_3_percent_variable_cpg[, new_order]
beta <- pheatmap(scale(top_3_percent_variable_cpg), cluster_cols = F)
```

```{r}
top_3_percent_variable_cpg_M_untreated <- top_3_percent_variable_cpg_M[,-c(19:36)]
beta_m_untreated <- pheatmap(scale(top_3_percent_variable_cpg_M_untreated), cluster_cols = F)
```

```{r}
top_3_percent_variable_cpg_untreated <- top_3_percent_variable_cpg[,-c(19:36)]
beta_untreated <- pheatmap(scale(top_3_percent_variable_cpg_untreated), cluster_cols = F)
```


###lets see al the cpg islands
```{r}
top_50_percent_variable_cpg <- get_most_variable(input_beta_matrix = cpg_beta_ordered, x = 50)   
beta_all <- pheatmap(scale(top_50_percent_variable_cpg), cluster_cols = F)
```

```{r}
top_50_percent_variable_cpg_untreated <- top_50_percent_variable_cpg[,-c(19:36)]
beta_untreated <- pheatmap(scale(top_50_percent_variable_cpg_untreated), cluster_cols = F)
```

0
