---
title: "Preclinical DESeq2 — HPV+ vs HPV- per cell line"
output: html_document
---

```{r config, include=FALSE}
# Load paths and constants (DATA_DIR, RESULTS_DIR, PRECLIN_TX_DATA, PRECLIN_TX_OUT, ...).
# Place raw input files under data/preclinical_transcriptomics/ (see data/README.md).
source(here::here("config.R"))
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Aim

## Generating Count files for all 36 HPV samples from 6 Cell Lines, FaDu, Cal33, SAS (HPV-) and UDSCC2, UPCISCC154, UMSCC47 (HPV+)


# 2. Loading packages 

```{r echo=TRUE, warning=FALSE, results='hide', message=FALSE}
#if (!require("BiocManager", quietly = TRUE))
    #install.packages("BiocManager")

#BiocManager::install("apeglm")
library(apeglm)
library(DESeq2)
library(ashr) #in order to shrink logFC if we use contrasts instead of coefficients to retrieve results
library(tidyverse)
library(ggplot2)
library(readxl)
library(pheatmap)
library(tibble)
library(dplyr)
```


# 3. Making input files ready for deseq2, start with whole comparison

```{r}
Matrix = read.csv(file.path(PRECLIN_TX_DATA, "countall.csv")) 
attach(Matrix)
#if we target the mean of counts
#Matrix = aggregate(Matrix,FUN = mean,by=list(Matrix$Genes))

#if we a trget max of counts
Matrix = aggregate(Matrix,FUN = max,by=list(Matrix$Genes))
Matrix <- Matrix[-1,-1]

#write.csv(Matrix, "Maxcountall.csv")
```


```{r}
#setting the row names (genes) for each samples

HNSCC_matrix = Matrix[,-1]
rownames(HNSCC_matrix) = Matrix[,1]

#input of the metadata
Metadata = read.csv(file.path(PRECLIN_TX_DATA, "Metadata_HNSCC.csv"))
rownames(Metadata) <- Metadata[,1]
Metadata = subset(Metadata, select = -1 )
#if the row names of metadata is in same seq as column of matrix
all(rownames(Metadata) == colnames(HNSCC_matrix))



```

# Create DESeq object
```{r}
dds_wt <- DESeqDataSetFromMatrix(countData = round(HNSCC_matrix),
                                 colData = Metadata,
                                 design = ~ Group)
```


#keeping genes with count values >= 10
```{r}
keep <- rowSums(counts(dds_wt)) >= 10
dds_wt <- dds_wt[keep,]

# Get the gene identifiers of the remaining genes
#remaining_genes <- rownames(dds_wt)
#remaining_genes <- data.frame(remaining_genes)
#remaining_genes

# Optionally, you could also determine the total number of remaining genes
#num_remaining_genes <- length(remaining_genes)
#num_remaining_genes
```


#Count normalization---> calculation
```{r}
dds_wt <- estimateSizeFactors(dds_wt)
sizeFactors(dds_wt)
```



#Count normalization---> extraction
```{r}
normalized_wt_counts <- counts(dds_wt, normalized=TRUE)
#write.csv(normalized_wt_counts,'HPV_all_normalized.csv')

```




# Run analysis
```{r, message=FALSE}
dds_wt <- DESeq(dds_wt)



wt_res <- results(dds_wt,
                  contrast = c("Group", "Untreated",
                               "Treated_4GY"),
                  alpha = 0.05,
                  lfcThreshold = .585)
wt_res
```




```{r}
result = data.frame(wt_res)

write.csv(result, file = file.path(PRECLIN_TX_OUT, "deg_untreatedvstreated_lfcchange1.5.csv"))
padj <- result$padj
any(padj <0.05)
```
So, no signficant genes between the treated vs untreated


#lets check if the treatment varies with HPV status



# 3. Making input files ready for deseq2, start with whole comparison

```{r}
Matrix = read.csv(file.path(PRECLIN_TX_DATA, "countall.csv")) 
attach(Matrix)
#if we target the mean of counts
#Matrix = aggregate(Matrix,FUN = mean,by=list(Matrix$Genes))

#if we a trget max of counts
Matrix = aggregate(Matrix,FUN = max,by=list(Matrix$Genes))
Matrix <- Matrix[-1,-1]

#write.csv(Matrix, "Maxcountall.csv")
```


```{r}
#setting the row names (genes) for each samples

HNSCC_matrix = Matrix[,-1]
rownames(HNSCC_matrix) = Matrix[,1]

#input of the metadata
Metadata = read.csv(file.path(PRECLIN_TX_DATA, "Metadata_HNSCC.csv"))
rownames(Metadata) <- Metadata[,1]
Metadata = subset(Metadata, select = -1 )
#if the row names of metadata is in same seq as column of matrix
all(rownames(Metadata) == colnames(HNSCC_matrix))



```
# Create DESeq object
```{r}
dds_wt <- DESeqDataSetFromMatrix(countData = round(HNSCC_matrix),
                                 colData = Metadata,
                                design = ~ HPV)
```

# Create DESeq object (treatment with respect to HPV)
```{r}
dds_wt <- DESeqDataSetFromMatrix(countData = round(HNSCC_matrix),
                                 colData = Metadata,
                                design = ~ HPV + Group + HPV:Group)
```


#keeping genes with count values >= 10
```{r}
keep <- rowSums(counts(dds_wt)) >= 10
dds_wt <- dds_wt[keep,]
```


```{r}
# Get the gene identifiers of the remaining genes
remaining_genes <- rownames(dds_wt)
remaining_genes <- data.frame(remaining_genes)
remaining_genes

# Optionally, you could also determine the total number of remaining genes
#num_remaining_genes <- length(remaining_genes)
#num_remaining_genes
```


#Count normalization---> calculation
```{r}
dds_wt <- estimateSizeFactors(dds_wt)
sizeFactors(dds_wt)
```



#Count normalization---> extraction
```{r}
normalized_wt_counts <- counts(dds_wt, normalized=TRUE)
#write.csv(normalized_wt_counts,'HPV_all_normalized.csv')

```




# Run analysis
```{r, message=FALSE}
dds_wt <- DESeq(dds_wt)


wt_res <- results(dds_wt, contrast = c("HPV", "HPV_Positive",
                               "HPV_Negative"),alpha = 0.05, lfcThreshold = .585)
wt_res
```





```{r}
result = data.frame(wt_res)

#write.csv(result, file = 'deg_allsamples_HPV-vsHPV+_lfcchange1.5.csv')

result %>%
  arrange(padj)
filtered_padj <- filter(result,padj<0.05)
#write.csv(filtered_padj,"significant degs allsamplesHPV-vsHPV+_lfc1.5change.csv")
dim(filtered_padj)
```

include a volcano plot

```{r}
# Create a volcano plot with different colors for upregulated and downregulated genes
volcano_plot <- ggplot(result, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = ifelse(padj < 0.05, ifelse(log2FoldChange > 0, "Upregulated", "Downregulated"), "Not Significant")), size = 2) +
  scale_color_manual(values = c("Upregulated" = "#1f78b4", "Downregulated" = "#FFB000", "Not Significant" = "gray")) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(x = "Log2 Fold Change", y = "-log10(padj)") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(-0.585, 0.585), linetype = "dashed", color = "black") +
  ggtitle("HPV- vs HPV+ (untreated) ")

# Print the plot
print(volcano_plot)

ggsave(file.path(PRECLIN_TX_OUT, "volcano_untreated_HPVneg_vs_HPVpos.png"),volcano_plot,width = 6, height = 4 )
 

```


```{r}
vsd <- vst(dds_wt, blind=FALSE)
plotPCA(vsd, intgroup=c("Cell_Line", "Group"))
```



#Analyses Untreated vs Treated cell line specific

# For individual cell line analyses
#Fadu Untreated vs treated
```{r}
#input of the count matrix as csv file
Matrix = read.csv(file.path(PRECLIN_TX_DATA, "countall.csv")) 
Matrix_fadu <- Matrix %>% select(starts_with('FaDu'))
Matrix_genes <- Matrix %>% select(starts_with('Genes'))
Matrix <- cbind(Matrix_genes, Matrix_fadu)
attach(Matrix)


#if we target the mean of counts
#Matrix = aggregate(Matrix,FUN = mean,by=list(Matrix$Genes))

#if we a trget max of counts
Matrix = aggregate(Matrix,FUN = max,by=list(Matrix$Genes))
Matrix <- Matrix[-1,-1]
#setting the row names (genes) for each samples

HNSCC_matrix = Matrix[,-1]
rownames(HNSCC_matrix) = Matrix[,1]


#input of the metadata
Metadata = read.csv(file.path(PRECLIN_TX_DATA, "Metadata_HNSCC.csv"))
fadu1 <- Metadata[1:3,]
fadu2 <- Metadata[19:21,]
Metadata <- rbind(fadu1,fadu2)
rownames(Metadata) <- Metadata[,1]
Metadata = subset(Metadata, select = -1 )
#if the row names of metadata is in same seq as column of matrix
all(rownames(Metadata) == colnames(HNSCC_matrix))


```

# Create DESeq object
```{r}
dds_wt <- DESeqDataSetFromMatrix(countData = round(HNSCC_matrix),
                                 colData = Metadata,
                                 design = ~ Group)
```


#keeping genes with count values >= 10
```{r}
keep <- rowSums(counts(dds_wt)) >= 10
dds_wt <- dds_wt[keep,]
```


#Count normalization---> calculation
```{r}
dds_wt <- estimateSizeFactors(dds_wt)
sizeFactors(dds_wt)
```



#Count normalization---> extraction
```{r}
normalized_wt_counts <- counts(dds_wt, normalized=TRUE)
write.csv(normalized_wt_counts,'HPV_Fadu_normalized.csv')

```




# Run analysis
```{r, message=FALSE}
dds_wt <- DESeq(dds_wt)

results(dds_wt, alpha = 0.05) #sample result

wt_res <- results(dds_wt,
                  contrast = c("Group", "Untreated",
                               "Treated_4GY"),
                  alpha = 0.05,
                  lfcThreshold = 0.585)
wt_res
```




```{r include=FALSE}
result = data.frame(wt_res)

write.csv(result, file = 'deg_Fadu_lfchcange1.5.csv')
padj <- result$padj
any(padj <0.05)
result %>%
  arrange(padj)
filter(result,padj<0.05)
```
#SAS Untreated vs treated
```{r}
#input of the count matrix as csv file
Matrix = read.csv(file.path(PRECLIN_TX_DATA, "countall.csv")) 
Matrix_fadu <- Matrix %>% select(starts_with('SAS'))
Matrix_genes <- Matrix %>% select(starts_with('Genes'))
Matrix <- cbind(Matrix_genes, Matrix_fadu)
attach(Matrix)


#if we target the mean of counts
#Matrix = aggregate(Matrix,FUN = mean,by=list(Matrix$Genes))

#if we a trget max of counts
Matrix = aggregate(Matrix,FUN = max,by=list(Matrix$Genes))
Matrix <- Matrix[-1,-1]
#setting the row names (genes) for each samples

HNSCC_matrix = Matrix[,-1]
rownames(HNSCC_matrix) = Matrix[,1]


#input of the metadata
Metadata = read.csv(file.path(PRECLIN_TX_DATA, "Metadata_HNSCC.csv"))
fadu1 <- Metadata[4:6,]
fadu2 <- Metadata[22:24,]
Metadata <- rbind(fadu1,fadu2)
rownames(Metadata) <- Metadata[,1]
Metadata = subset(Metadata, select = -1 )
#if the row names of metadata is in same seq as column of matrix
all(rownames(Metadata) == colnames(HNSCC_matrix))


```

# Create DESeq object
```{r}
dds_wt <- DESeqDataSetFromMatrix(countData = round(HNSCC_matrix),
                                 colData = Metadata,
                                 design = ~ Group)
```


#keeping genes with count values >= 10
```{r}
keep <- rowSums(counts(dds_wt)) >= 10
dds_wt <- dds_wt[keep,]
```


#Count normalization---> calculation
```{r}
dds_wt <- estimateSizeFactors(dds_wt)
sizeFactors(dds_wt)
```



#Count normalization---> extraction
```{r}
normalized_wt_counts <- counts(dds_wt, normalized=TRUE)
write.csv(normalized_wt_counts,'HPV_SAS_normalized.csv')

```




# Run analysis
```{r, message=FALSE}
dds_wt <- DESeq(dds_wt)

results(dds_wt, alpha = 0.05) #sample result

wt_res <- results(dds_wt,
                  contrast = c("Group", "Untreated",
                               "Treated_4GY"),
                  alpha = 0.05,
                  lfcThreshold = 0.585)
wt_res
```




```{r include=FALSE}
result = data.frame(wt_res)

write.csv(result, file = 'deg_SAS_lfchcange1.5.csv')
padj <- result$padj
any(padj <0.05)
result %>%
  arrange(padj)
filter(result,padj<0.05)
```
#Cal33 Untreated vs treated
```{r}
#input of the count matrix as csv file
Matrix = read.csv(file.path(PRECLIN_TX_DATA, "countall.csv")) 
Matrix_fadu <- Matrix %>% select(starts_with('Cal'))
Matrix_genes <- Matrix %>% select(starts_with('Genes'))
Matrix <- cbind(Matrix_genes, Matrix_fadu)
attach(Matrix)


#if we target the mean of counts
#Matrix = aggregate(Matrix,FUN = mean,by=list(Matrix$Genes))

#if we a trget max of counts
Matrix = aggregate(Matrix,FUN = max,by=list(Matrix$Genes))
Matrix <- Matrix[-1,-1]
#setting the row names (genes) for each samples

HNSCC_matrix = Matrix[,-1]
rownames(HNSCC_matrix) = Matrix[,1]


#input of the metadata
Metadata = read.csv(file.path(PRECLIN_TX_DATA, "Metadata_HNSCC.csv"))
fadu1 <- Metadata[7:9,]
fadu2 <- Metadata[25:27,]
Metadata <- rbind(fadu1,fadu2)
rownames(Metadata) <- Metadata[,1]
Metadata = subset(Metadata, select = -1 )
#if the row names of metadata is in same seq as column of matrix
all(rownames(Metadata) == colnames(HNSCC_matrix))


```

# Create DESeq object
```{r}
dds_wt <- DESeqDataSetFromMatrix(countData = round(HNSCC_matrix),
                                 colData = Metadata,
                                 design = ~ Group)
```


#keeping genes with count values >= 10
```{r}
keep <- rowSums(counts(dds_wt)) >= 10
dds_wt <- dds_wt[keep,]
```


#Count normalization---> calculation
```{r}
dds_wt <- estimateSizeFactors(dds_wt)
sizeFactors(dds_wt)
```



#Count normalization---> extraction
```{r}
normalized_wt_counts <- counts(dds_wt, normalized=TRUE)
write.csv(normalized_wt_counts,'HPV_Cal_normalized.csv')

```




# Run analysis
```{r, message=FALSE}
dds_wt <- DESeq(dds_wt)

results(dds_wt, alpha = 0.05) #sample result

wt_res <- results(dds_wt,
                  contrast = c("Group", "Untreated",
                               "Treated_4GY"),
                  alpha = 0.05,
                  lfcThreshold = 0.585)
wt_res
```




```{r include=FALSE}
result = data.frame(wt_res)

write.csv(result, file = 'deg_Cal33_lfchcange1.5.csv')
padj <- result$padj
any(padj <0.05)
result %>%
  arrange(padj)
filter(result,padj<0.05)
```
#UDSCC2 Untreated vs treated
```{r}
#input of the count matrix as csv file
Matrix = read.csv(file.path(PRECLIN_TX_DATA, "countall.csv")) 
Matrix_fadu <- Matrix %>% select(starts_with('UD'))
Matrix_genes <- Matrix %>% select(starts_with('Genes'))
Matrix <- cbind(Matrix_genes, Matrix_fadu)
attach(Matrix)


#if we target the mean of counts
#Matrix = aggregate(Matrix,FUN = mean,by=list(Matrix$Genes))

#if we a trget max of counts
Matrix = aggregate(Matrix,FUN = max,by=list(Matrix$Genes))
Matrix <- Matrix[-1,-1]
#setting the row names (genes) for each samples

HNSCC_matrix = Matrix[,-1]
rownames(HNSCC_matrix) = Matrix[,1]


#input of the metadata
Metadata = read.csv(file.path(PRECLIN_TX_DATA, "Metadata_HNSCC.csv"))
fadu1 <- Metadata[10:12,]
fadu2 <- Metadata[28:30,]
Metadata <- rbind(fadu1,fadu2)
rownames(Metadata) <- Metadata[,1]
Metadata = subset(Metadata, select = -1 )
#if the row names of metadata is in same seq as column of matrix
all(rownames(Metadata) == colnames(HNSCC_matrix))


```

# Create DESeq object
```{r}
dds_wt <- DESeqDataSetFromMatrix(countData = round(HNSCC_matrix),
                                 colData = Metadata,
                                 design = ~ Group)
```


#keeping genes with count values >= 10
```{r}
keep <- rowSums(counts(dds_wt)) >= 10
dds_wt <- dds_wt[keep,]
```


#Count normalization---> calculation
```{r}
dds_wt <- estimateSizeFactors(dds_wt)
sizeFactors(dds_wt)
```



#Count normalization---> extraction
```{r}
normalized_wt_counts <- counts(dds_wt, normalized=TRUE)
write.csv(normalized_wt_counts,'HPV_UDSCC2_normalized.csv')

```




# Run analysis
```{r, message=FALSE}
dds_wt <- DESeq(dds_wt)

results(dds_wt, alpha = 0.05) #sample result

wt_res <- results(dds_wt,
                  contrast = c("Group", "Untreated",
                               "Treated_4GY"),
                  alpha = 0.05,
                  lfcThreshold = 0.585)
wt_res


```




```{r include=FALSE}
result = data.frame(wt_res)

write.csv(result, file = 'deg_UDSCC2_lfchcange1.5.csv')
padj <- result$padj
any(padj <0.05)
result %>%
  arrange(padj)
filter(result,padj<0.05)
```
#UMSCC47 Untreated vs treated
```{r}
#input of the count matrix as csv file
Matrix = read.csv(file.path(PRECLIN_TX_DATA, "countall.csv")) 
Matrix_fadu <- Matrix %>% select(starts_with('UM'))
Matrix_genes <- Matrix %>% select(starts_with('Genes'))
Matrix <- cbind(Matrix_genes, Matrix_fadu)
attach(Matrix)


#if we target the mean of counts
#Matrix = aggregate(Matrix,FUN = mean,by=list(Matrix$Genes))

#if we a trget max of counts
Matrix = aggregate(Matrix,FUN = max,by=list(Matrix$Genes))
Matrix <- Matrix[-1,-1]
#setting the row names (genes) for each samples

HNSCC_matrix = Matrix[,-1]
rownames(HNSCC_matrix) = Matrix[,1]


#input of the metadata
Metadata = read.csv(file.path(PRECLIN_TX_DATA, "Metadata_HNSCC.csv"))
fadu1 <- Metadata[13:15,]
fadu2 <- Metadata[31:33,]
Metadata <- rbind(fadu1,fadu2)
rownames(Metadata) <- Metadata[,1]
Metadata = subset(Metadata, select = -1 )
#if the row names of metadata is in same seq as column of matrix
all(rownames(Metadata) == colnames(HNSCC_matrix))


```

# Create DESeq object
```{r}
dds_wt <- DESeqDataSetFromMatrix(countData = round(HNSCC_matrix),
                                 colData = Metadata,
                                 design = ~ Group)
```


#keeping genes with count values >= 10
```{r}
keep <- rowSums(counts(dds_wt)) >= 10
dds_wt <- dds_wt[keep,]
```


#Count normalization---> calculation
```{r}
dds_wt <- estimateSizeFactors(dds_wt)
sizeFactors(dds_wt)
```



#Count normalization---> extraction
```{r}
normalized_wt_counts <- counts(dds_wt, normalized=TRUE)
write.csv(normalized_wt_counts,'HPV_UMSCC47_normalized.csv')

```




# Run analysis
```{r, message=FALSE}
dds_wt <- DESeq(dds_wt)

results(dds_wt, alpha = 0.05) #sample result

wt_res <- results(dds_wt,
                  contrast = c("Group", "Untreated",
                               "Treated_4GY"),
                  alpha = 0.05,
                  lfcThreshold = 0.585)
wt_res
```




```{r include=FALSE}
result = data.frame(wt_res)

write.csv(result, file = 'deg_UMSCC47_lfchcange1.5.csv')
padj <- result$padj
any(padj <0.05)
result %>%
  arrange(padj)
filter(result,padj<0.05)
```
#UPCI154 Untreated vs treated
```{r}
#input of the count matrix as csv file
Matrix = read.csv(file.path(PRECLIN_TX_DATA, "countall.csv")) 
Matrix_fadu <- Matrix %>% select(starts_with('UP'))
Matrix_genes <- Matrix %>% select(starts_with('Genes'))
Matrix <- cbind(Matrix_genes, Matrix_fadu)
attach(Matrix)


#if we target the mean of counts
#Matrix = aggregate(Matrix,FUN = mean,by=list(Matrix$Genes))

#if we a trget max of counts
Matrix = aggregate(Matrix,FUN = max,by=list(Matrix$Genes))
Matrix <- Matrix[-1,-1]
#setting the row names (genes) for each samples

HNSCC_matrix = Matrix[,-1]
rownames(HNSCC_matrix) = Matrix[,1]


#input of the metadata
Metadata = read.csv(file.path(PRECLIN_TX_DATA, "Metadata_HNSCC.csv"))
fadu1 <- Metadata[16:18,]
fadu2 <- Metadata[34:36,]
Metadata <- rbind(fadu1,fadu2)
rownames(Metadata) <- Metadata[,1]
Metadata = subset(Metadata, select = -1 )
#if the row names of metadata is in same seq as column of matrix
all(rownames(Metadata) == colnames(HNSCC_matrix))


```

# Create DESeq object
```{r}
dds_wt <- DESeqDataSetFromMatrix(countData = round(HNSCC_matrix),
                                 colData = Metadata,
                                 design = ~ Group)
```


#keeping genes with count values >= 10
```{r}
keep <- rowSums(counts(dds_wt)) >= 10
dds_wt <- dds_wt[keep,]
```


#Count normalization---> calculation
```{r}
dds_wt <- estimateSizeFactors(dds_wt)
sizeFactors(dds_wt)
```



#Count normalization---> extraction
```{r}
normalized_wt_counts <- counts(dds_wt, normalized=TRUE)
write.csv(normalized_wt_counts,'HPV_UPCI154_normalized.csv')

```




# Run analysis
```{r, message=FALSE}
dds_wt <- DESeq(dds_wt)

results(dds_wt, alpha = 0.05) #sample result

wt_res <- results(dds_wt,
                  contrast = c("Group", "Untreated",
                               "Treated_4GY"),
                  alpha = 0.05,
                  lfcThreshold = 0.585)
wt_res
```




```{r include=FALSE}
result = data.frame(wt_res)

write.csv(result, file = 'deg_UPCI154_lfchcange1.5.csv')
padj <- result$padj
any(padj <0.05)
result %>%
  arrange(padj)
filter(result,padj<0.05)
```








####all cell lines



#Untreated HPV- vs +
```{r}
#input of the count matrix as csv file
Matrix = read.csv(file.path(PRECLIN_TX_DATA, "countall.csv")) 
Matrix<- Matrix[,1:19]
attach(Matrix)


#if we target the mean of counts
#Matrix = aggregate(Matrix,FUN = mean,by=list(Matrix$Genes))

#if we a trget max of counts
Matrix = aggregate(Matrix,FUN = max,by=list(Matrix$Genes))
Matrix <- Matrix[-1,-1]
#setting the row names (genes) for each samples

HNSCC_matrix = Matrix[,-1]
rownames(HNSCC_matrix) = Matrix[,1]


#input of the metadata
Metadata = read.csv(file.path(PRECLIN_TX_DATA, "Metadata_HNSCC.csv"))
Metadata <- Metadata[1:18,]
rownames(Metadata) <- Metadata[,1]
Metadata = subset(Metadata, select = -1 )
#if the row names of metadata is in same seq as column of matrix
all(rownames(Metadata) == colnames(HNSCC_matrix))


```

# Create DESeq object
```{r}
dds_wt <- DESeqDataSetFromMatrix(countData = round(HNSCC_matrix),
                                 colData = Metadata,
                                 design = ~ HPV)

#keeping genes with count values >= 10
keep <- rowSums(counts(dds_wt)) >= 10
dds_wt <- dds_wt[keep,]
```


#Count normalization---> calculation
```{r}
dds_wt <- estimateSizeFactors(dds_wt)
sizeFactors(dds_wt)

#Count normalization---> extraction
normalized_wt_counts <- counts(dds_wt, normalized=TRUE)
#write.csv(normalized_wt_counts,'Untreated_normalized.csv')
plot(normalized_wt_counts)
boxplot(log10(assays(dds_wt)[["cooks"]]), range=0, las=2,main="Cook's distance", cex.axis=0.3)
```




# Run analysis
```{r, message=FALSE}
dds_wt <- DESeq(dds_wt)


wt_res <- results(dds_wt,
                  contrast = c("HPV", "HPV_Positive",
                               "HPV_Negative"),
                  alpha = 0.05,
                  lfcThreshold = 0.585)
wt_res
result = data.frame(wt_res)
# Create a new column 'gene_symbol' with row names (gene names)
  result$gene_symbol <- rownames(result)
 # Optionally, if you want 'gene_symbol' to be the first column
  result <- result[, c("gene_symbol", setdiff(names(result), "gene_symbol"))]
```


include a volcano plot

```{r}
# Create a volcano plot with different colors for upregulated and downregulated genes
volcano_plot <- ggplot(result, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = ifelse(padj < 0.05, ifelse(log2FoldChange > 0, "Upregulated", "Downregulated"), "Not Significant")), size = 2) +
  scale_color_manual(values = c("Upregulated" = "#1f78b4", "Downregulated" = "#FFB000", "Not Significant" = "gray")) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(x = "Log2 Fold Change", y = "-log10(padj)") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(-0.585, 0.585), linetype = "dashed", color = "black") +
  ggtitle("HPV- vs HPV+ (untreated) ")

# Print the plot
print(volcano_plot)

ggsave(file.path(PRECLIN_TX_OUT, "volcano_untreated_HPVneg_vs_HPVpos.png"),volcano_plot,width = 6, height = 4 )
 

```


```{r include=FALSE}


write.csv(result, file = 'deg_Untreated_fchcange1.5.csv', row.names = F)
padj <- result$padj
any(padj <0.05)
result %>%
  arrange(padj)
filtered_padj <- filter(result,padj<0.05)
write.csv(filtered_padj, file.path(PRECLIN_TX_OUT, "significant_degs_untreated_fc1.5change.csv"))
dim(filtered_padj)
```

```{r}
vsd <- vst(dds_wt, blind=FALSE)
plotPCA(vsd, intgroup=c("Cell_Line", "Group"))
```










#Treated HPV- vs +






```{r}
#input of the count matrix as csv file
Matrix = read.csv(file.path(PRECLIN_TX_DATA, "countall.csv")) 
Matrix_fadu <- Matrix[,20:37]
Matrix_genes <- Matrix %>% dplyr::select(starts_with('Genes'))
Matrix <- cbind(Matrix_genes, Matrix_fadu)
attach(Matrix)


#if we target the mean of counts
#Matrix = aggregate(Matrix,FUN = mean,by=list(Matrix$Genes))

#if we a trget max of counts
Matrix = aggregate(Matrix,FUN = max,by=list(Matrix$Genes))
Matrix <- Matrix[-1,-1]
#setting the row names (genes) for each samples

HNSCC_matrix = Matrix[,-1]
rownames(HNSCC_matrix) = Matrix[,1]


#input of the metadata
Metadata = read.csv(file.path(PRECLIN_TX_DATA, "Metadata_HNSCC.csv"))
Metadata <- Metadata[19:36,]
rownames(Metadata) <- Metadata[,1]
Metadata = subset(Metadata, select = -1 )
#if the row names of metadata is in same seq as column of matrix
all(rownames(Metadata) == colnames(HNSCC_matrix))


```

# Create DESeq object
```{r}
dds_wt <- DESeqDataSetFromMatrix(countData = round(HNSCC_matrix),
                                 colData = Metadata,
                                 design = ~ HPV)
#keeping genes with count values >= 10
keep <- rowSums(counts(dds_wt)) >= 10
dds_wt <- dds_wt[keep,]
```


#Count normalization---> calculation
```{r}
dds_wt <- estimateSizeFactors(dds_wt)
sizeFactors(dds_wt)

#Count normalization---> extraction
normalized_wt_counts <- counts(dds_wt, normalized=TRUE)
#write.csv(normalized_wt_counts,'Treated_normalized.csv')
plot(normalized_wt_counts)
```




# Run analysis
```{r, message=FALSE}
dds_wt <- DESeq(dds_wt)

wt_res <- results(dds_wt,
                  contrast = c("HPV", "HPV_Positive",
                               "HPV_Negative"),
                  alpha = 0.05,
                  lfcThreshold = 0.585)
wt_res
result = data.frame(wt_res)

#wt_res_shrunk <- lfcShrink(dds_wt, 
                           #contrast=c("HPV", "HPV_Negative", "HPV_Positive"),
                           #res=wt_res, 
                           #type='normal')
#result_shrunk = data.frame(wt_res_shrunk)

#plotMA(wt_res, ylim=c(-2,2))

# Create a new column 'gene_symbol' with row names (gene names)
  result$gene_symbol <- rownames(result)
  # Optionally, if you want 'gene_symbol' to be the first column
  result <- result[, c("gene_symbol", setdiff(names(result), "gene_symbol"))]
  


```

```{r}
# Create a volcano plot with different colors for upregulated and downregulated genes
volcano_plot <- ggplot(result, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = ifelse(padj < 0.05, ifelse(log2FoldChange > 0, "Upregulated", "Downregulated"), "Not Significant")), size = 2) +
  scale_color_manual(values = c("Upregulated" = "#1f78b4", "Downregulated" = "#FFB000", "Not Significant" = "gray")) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(x = "Log2 Fold Change", y = "-log10(padj)") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(-0.585, 0.585), linetype = "dashed", color = "black") +
  ggtitle("HPV- vs HPV+ (treated) ")

# Print the plot
print(volcano_plot)

ggsave(file.path(PRECLIN_TX_OUT, "volcano_treated_HPVneg_vs_HPVpos.png"),volcano_plot,width = 6, height = 4 )
```



```{r include=FALSE}


write.csv(result, file = 'deg_treated_lfchcange1.5.csv', row.names = F)
padj <- result$padj
any(padj <0.05)
result %>%
  arrange(padj)
filtered_padj <- filter(result,padj<0.05)
write.csv(filtered_padj, file.path(PRECLIN_TX_OUT, "significant_degs_treated_lfc1.5change.csv"))
dim(filtered_padj)
```






# 3. All samples, HPV- vs HPV+

```{r}
Matrix = read.csv(file.path(PRECLIN_TX_DATA, "countall.csv")) 
attach(Matrix)
#if we target the mean of counts
#Matrix = aggregate(Matrix,FUN = mean,by=list(Matrix$Genes))

#if we a trget max of counts
Matrix = aggregate(Matrix,FUN = max,by=list(Matrix$Genes))
Matrix <- Matrix[-1,-1]
#setting the row names (genes) for each samples

HNSCC_matrix = Matrix[,-1]
rownames(HNSCC_matrix) = Matrix[,1]

#input of the metadata
Metadata = read.csv(file.path(PRECLIN_TX_DATA, "Metadata_HNSCC.csv"))
rownames(Metadata) <- Metadata[,1]
Metadata = subset(Metadata, select = -1 )
#if the row names of metadata is in same seq as column of matrix
all(rownames(Metadata) == colnames(HNSCC_matrix))



```

# Create DESeq object
```{r}
dds_wt <- DESeqDataSetFromMatrix(countData = round(HNSCC_matrix),
                                 colData = Metadata,
                                 design = ~ HPV)
```


#keeping genes with count values >= 10
```{r}
keep <- rowSums(counts(dds_wt)) >= 10
dds_wt <- dds_wt[keep,]
```


#Count normalization---> calculation
```{r}
dds_wt <- estimateSizeFactors(dds_wt)
sizeFactors(dds_wt)
```



#Count normalization---> extraction
```{r}
normalized_wt_counts <- counts(dds_wt, normalized=TRUE)
#write.csv(normalized_wt_counts,'HPV_all_normalized_esignbyHPVstatus.csv')

```




# Run analysis
```{r, message=FALSE}
dds_wt <- DESeq(dds_wt)

wt_res <- results(dds_wt,
                  contrast = c("HPV", "HPV_Positive",
                               "HPV_Negative"),
                  alpha = 0.05,
                  lfcThreshold = .585)
wt_res
result = data.frame(wt_res)
```


```{r}
# Create a volcano plot with different colors for upregulated and downregulated genes
volcano_plot <- ggplot(result, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = ifelse(padj < 0.05, ifelse(log2FoldChange > 0, "Upregulated", "Downregulated"), "Not Significant")), size = 2) +
  scale_color_manual(values = c("Upregulated" = "#1f78b4", "Downregulated" = "#FFB000", "Not Significant" = "gray")) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(x = "Log2 Fold Change", y = "-log10(padj)") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(-0.585, 0.585), linetype = "dashed", color = "black") +
  ggtitle("HPV- vs HPV+ (All samples) ")

# Print the plot
print(volcano_plot)

ggsave(file.path(PRECLIN_TX_OUT, "volcano_allsamples_HPVneg_vs_HPVpos.png"),volcano_plot,width = 6, height = 4 )
```

```{r}


#write.csv(result, file = 'deg_allsaples_HPV-vsHPV+_lfcchange1.5.csv')
padj <- result$padj
any(padj <0.05)
result %>%
  arrange(padj)
filtered_padj <- filter(result,padj<0.05)
#write.csv(filtered_padj,"significant degs allsamplesHPV-vsHPV+_lfc1.5change.csv")
dim(filtered_padj)
```




```{r}
vsd <- vst(dds_wt, blind=FALSE)
plotPCA(vsd, intgroup=c("Cell_Line"))
```
##another method for subsetting data

```{r}
#convert results into df
res_control <- as.data.frame(wt_res)

#subset
res.untreated_HPV_neg_vsHPV_pos_sub <- subset(res_control, padj < 0.05)
res.untreated_HPV_neg_vsHPV_pos_sub_final <- na.omit(res.untreated_HPV_neg_vsHPV_pos_sub)
res.untreated_HPV_neg_vsHPV_pos_sub_final <- cbind(gene_symbol = rownames(res.untreated_HPV_neg_vsHPV_pos_sub_final ), res.untreated_HPV_neg_vsHPV_pos_sub_final ) #we add a column with the gene symbol because we will need it for GSEA
head(res.untreated_HPV_neg_vsHPV_pos_sub_final)
write.csv(res.untreated_HPV_neg_vsHPV_pos_sub_final, file.path(PRECLIN_TX_OUT, "res_untreated_HPVneg_vs_HPVpos_sub_final.csv"))

```



```{r}
vsd <- vst(dds_wt, blind=FALSE)
plotPCA(vsd, intgroup=c("Cell_Line", "Group"))
```


```{r}
library(hrbrthemes)
pcaData <- plotPCA(vsd, intgroup=c("Cell_Line", "Group"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=Cell_Line, shape=Group)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()+ theme_ipsum()

```

```{r, message=FALSE}

library (genefilter)
topVarGenes <- head(order(rowVars(assay(vsd)), decreasing = TRUE), 20)

mat  <- assay(vsd)[ topVarGenes, ]
mat  <- mat - rowMeans(mat)
anno <- as.data.frame(colData(vsd)[, c("Cell_Line", "Group")])
pheatmap(mat, annotation_col = anno, cluster_cols= T, color=colorRampPalette(c("#104e8b","#e8e8e8","#b22222"))(n = 100), fontsize = 6.5)
```

###heatmap of vst files
```{r}
sig <- read.csv(file.path(PRECLIN_TX_OUT, "count_untreated_HPVneg_vs_HPVpos_significantgenes.csv"),header = TRUE, stringsAsFactors = FALSE, sep = ",")
vsdsig <-varianceStabilizingTransformation(object = as.matrix(sig[,-1]))
breaksList = seq(-3, 3, by = 1)
anno <- read.csv(file.path(PRECLIN_TX_DATA, "Metadata_HNSCC_untreated.csv"))
rownames(anno) <- colnames(vsdsig)
pheatmap(vsdsig,  show_colnames = F, 
         show_rownames = T,

         annotation_col = anno,
      
         annotation_colors = NA,
         clustering_distance_cols = 'euclidean',
         clustering_distance_rows = 'euclidean',
         breaks = breaksList,
         color = colorRampPalette(c("navy", "white", "firebrick3"))(length(breaksList)),
         cluster_cols = T,
         cluster_rows = T,
         border_color = NA)

```
#getting vst files

```{r}

#untreated files
Matrix <- read.csv(file.path(PRECLIN_TX_DATA, "count_untreated.csv"),header = TRUE, stringsAsFactors = FALSE, sep = ",")
#if we a trget max of counts
Matrix = aggregate(Matrix,FUN = max,by=list(Matrix$Genes))
Matrix <- Matrix[-1,-1]
rownames(Matrix) <- Matrix[ , 1]
vsd_untreated <-varianceStabilizingTransformation(object = as.matrix(Matrix[,-1]))
vsd_untreated <- write.csv(vsd_untreated, file.path(PRECLIN_TX_OUT, "Vsd_untreated.csv"))
#treated files

Matrix <- read.csv(file.path(PRECLIN_TX_DATA, "count_untreated.csv"),header = TRUE, stringsAsFactors = FALSE, sep = ",")
#if we a trget max of counts
Matrix = aggregate(Matrix,FUN = max,by=list(Matrix$Genes))
Matrix <- Matrix[-1,-1]
rownames(Matrix) <- Matrix[ , 1]
vsd_untreated <-varianceStabilizingTransformation(object = as.matrix(Matrix[,-1]))
vsd_untreated <- write.csv(vsd_untreated, file.path(PRECLIN_TX_OUT, "Vsd_untreated.csv"))


```



```{r}
Matrix <- read.csv(file.path(PRECLIN_TX_DATA, "countall.csv"),header = TRUE, stringsAsFactors = FALSE, sep = ",")
#if we a trget max of counts
Matrix = aggregate(Matrix,FUN = max,by=list(Matrix$Genes))
Matrix <- Matrix[-1,-1]
rownames(Matrix) <- Matrix[ , 1]
vsd_all <-varianceStabilizingTransformation(object = as.matrix(Matrix[,-1]))
#vsd_all<- write.csv(vsd_all,"Vsd_all.csv")
transposed_vsd <- t(vsd_all)
vsd_metadata <- read.csv(file.path(PRECLIN_TX_DATA, "Metadata_HNSCC.csv"))
vsd_metadata <- vsd_metadata[,2:5]
vsd_dataplusmetadata<- cbind(transposed_vsd, vsd_metadata)
plotpca <- prcomp(transposed_vsd)

library(ggfortify)
autoplot(plotpca, data = vsd_dataplusmetadata, colour = 'Cell_Line')
```


```{r}
library(factoextra)
plotpca2 <- prcomp(vsd_dataplusmetadata[, 1:55765])
fviz_pca_ind(plotpca2,
             col.ind = (vsd_metadata[,1]) # Color by the quality of representation
      
             )

```






```{r}

library(tidyverse)
library(hrbrthemes)
library(viridis)
#Maria_risk-score
file <- read.csv(file.path(PRECLIN_TX_DATA, "Maria_Risk_Score.csv"))
attach(file)
file %>%
  ggplot( aes(x=file$Cell_Lines, y=file$Risk.Score, fill=file$Cell_Lines)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() 

```




#Only HPV-




```{r}
#input of the count matrix as csv file
Matrix = read.csv(file.path(PRECLIN_TX_DATA, "countall.csv")) 
Matrix_fadu <- Matrix[,c(2:10,20:28)]
Matrix_genes <- Matrix %>% dplyr::select(starts_with('Genes'))
Matrix <- cbind(Matrix_genes, Matrix_fadu)
attach(Matrix)


#if we target the mean of counts
#Matrix = aggregate(Matrix,FUN = mean,by=list(Matrix$Genes))

#if we a trget max of counts
Matrix = aggregate(Matrix,FUN = max,by=list(Matrix$Genes))
Matrix <- Matrix[-1,-1]
#setting the row names (genes) for each samples

HNSCC_matrix = Matrix[,-1]
rownames(HNSCC_matrix) = Matrix[,1]


#input of the metadata
Metadata = read.csv(file.path(PRECLIN_TX_DATA, "Metadata_HNSCC.csv"))
Metadata <- Metadata[c(1:9,19:27),]
rownames(Metadata) <- Metadata[,1]
Metadata = subset(Metadata, select = -1 )
#if the row names of metadata is in same seq as column of matrix
all(rownames(Metadata) == colnames(HNSCC_matrix))


```

# Create DESeq object
```{r}
dds_wt <- DESeqDataSetFromMatrix(countData = round(HNSCC_matrix),
                                 colData = Metadata,
                                 design = ~ Group)
#keeping genes with count values >= 10
keep <- rowSums(counts(dds_wt)) >= 10
dds_wt <- dds_wt[keep,]
```


#Count normalization---> calculation
```{r}
dds_wt <- estimateSizeFactors(dds_wt)
sizeFactors(dds_wt)

#Count normalization---> extraction
normalized_wt_counts <- counts(dds_wt, normalized=TRUE)
#write.csv(normalized_wt_counts,'Treated_normalized.csv')

```




# Run analysis
```{r, message=FALSE}
dds_wt <- DESeq(dds_wt)

wt_res <- results(dds_wt,
                  contrast = c("Group", "Treated_4GY",
                               "Untreated"),
                  alpha = 0.05,
                  lfcThreshold = 0.585)
wt_res
result = data.frame(wt_res)

#wt_res_shrunk <- lfcShrink(dds_wt, 
                           #contrast=c("HPV", "HPV_Negative", "HPV_Positive"),
                           #res=wt_res, 
                           #type='normal')
#result_shrunk = data.frame(wt_res_shrunk)

#plotMA(wt_res, ylim=c(-2,2))

# Create a new column 'gene_symbol' with row names (gene names)
  result$gene_symbol <- rownames(result)
  # Optionally, if you want 'gene_symbol' to be the first column
  result <- result[, c("gene_symbol", setdiff(names(result), "gene_symbol"))]
  


```

```{r}
# Create a volcano plot with different colors for upregulated and downregulated genes
volcano_plot <- ggplot(result, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = ifelse(padj < 0.05, ifelse(log2FoldChange > 0, "Upregulated", "Downregulated"), "Not Significant")), size = 2) +
  scale_color_manual(values = c("Upregulated" = "#1f78b4", "Downregulated" = "#FFB000", "Not Significant" = "gray")) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(x = "Log2 Fold Change", y = "-log10(padj)") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(-0.585, 0.585), linetype = "dashed", color = "black") +
  ggtitle("HPV- vs HPV+ (treated) ")

# Print the plot
print(volcano_plot)

ggsave(file.path(PRECLIN_TX_OUT, "volcano_treated_HPVneg_vs_HPVpos.png"),volcano_plot,width = 6, height = 4 )
```



```{r include=FALSE}


#write.csv(result, file = 'deg_treated_lfchcange1.5.csv', row.names = F)
padj <- result$padj
any(padj <0.05)
result %>%
  arrange(padj)
filtered_padj <- filter(result,padj<0.05)
#write.csv(filtered_padj, file.path(PRECLIN_TX_OUT, "significant_degs_treated_lfc1.5change.csv"))
dim(filtered_padj)
```


#Only HPV+




```{r}
#input of the count matrix as csv file
Matrix = read.csv(file.path(PRECLIN_TX_DATA, "countall.csv")) 
Matrix_fadu <- Matrix[,c(11:19,29:37)]
Matrix_genes <- Matrix %>% dplyr::select(starts_with('Genes'))
Matrix <- cbind(Matrix_genes, Matrix_fadu)
attach(Matrix)


#if we target the mean of counts
#Matrix = aggregate(Matrix,FUN = mean,by=list(Matrix$Genes))

#if we a trget max of counts
Matrix = aggregate(Matrix,FUN = max,by=list(Matrix$Genes))
Matrix <- Matrix[-1,-1]
#setting the row names (genes) for each samples

HNSCC_matrix = Matrix[,-1]
rownames(HNSCC_matrix) = Matrix[,1]


#input of the metadata
Metadata = read.csv(file.path(PRECLIN_TX_DATA, "Metadata_HNSCC.csv"))
Metadata <- Metadata[c(10:18,28:36),]
rownames(Metadata) <- Metadata[,1]
Metadata = subset(Metadata, select = -1 )
#if the row names of metadata is in same seq as column of matrix
all(rownames(Metadata) == colnames(HNSCC_matrix))


```

# Create DESeq object
```{r}
dds_wt <- DESeqDataSetFromMatrix(countData = round(HNSCC_matrix),
                                 colData = Metadata,
                                 design = ~ Group)
#keeping genes with count values >= 10
keep <- rowSums(counts(dds_wt)) >= 10
dds_wt <- dds_wt[keep,]
```


#Count normalization---> calculation
```{r}
dds_wt <- estimateSizeFactors(dds_wt)
sizeFactors(dds_wt)

#Count normalization---> extraction
normalized_wt_counts <- counts(dds_wt, normalized=TRUE)
#write.csv(normalized_wt_counts,'Treated_normalized.csv')

```




# Run analysis
```{r, message=FALSE}
dds_wt <- DESeq(dds_wt)

wt_res <- results(dds_wt,
                  contrast = c("Group", "Treated_4GY",
                               "Untreated"),
                  alpha = 0.05,
                  lfcThreshold = 0.585)
wt_res
result = data.frame(wt_res)

#wt_res_shrunk <- lfcShrink(dds_wt, 
                           #contrast=c("HPV", "HPV_Negative", "HPV_Positive"),
                           #res=wt_res, 
                           #type='normal')
#result_shrunk = data.frame(wt_res_shrunk)

#plotMA(wt_res, ylim=c(-2,2))

# Create a new column 'gene_symbol' with row names (gene names)
  result$gene_symbol <- rownames(result)
  # Optionally, if you want 'gene_symbol' to be the first column
  result <- result[, c("gene_symbol", setdiff(names(result), "gene_symbol"))]
  


```

```{r}
# Create a volcano plot with different colors for upregulated and downregulated genes
volcano_plot <- ggplot(result, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = ifelse(padj < 0.05, ifelse(log2FoldChange > 0, "Upregulated", "Downregulated"), "Not Significant")), size = 2) +
  scale_color_manual(values = c("Upregulated" = "#1f78b4", "Downregulated" = "#FFB000", "Not Significant" = "gray")) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(x = "Log2 Fold Change", y = "-log10(padj)") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(-0.585, 0.585), linetype = "dashed", color = "black") +
  ggtitle("HPV- vs HPV+ (treated) ")

# Print the plot
print(volcano_plot)

ggsave(file.path(PRECLIN_TX_OUT, "volcano_treated_HPVneg_vs_HPVpos.png"),volcano_plot,width = 6, height = 4 )
```



```{r include=FALSE}


#write.csv(result, file = 'deg_treated_lfchcange1.5.csv', row.names = F)
padj <- result$padj
any(padj <0.05)
result %>%
  arrange(padj)
filtered_padj <- filter(result,padj<0.05)
#write.csv(filtered_padj, file.path(PRECLIN_TX_OUT, "significant_degs_treated_lfc1.5change.csv"))
dim(filtered_padj)
```
