s---
title: "Spatial Metaprograms: Foundation Data Preparation"
subtitle: "Clean spot-level master table with MP scores for downstream spatial analysis"
output: 
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE, fig.width = 12, fig.height = 8)
```

# Overview

**File 1 Purpose:** Create a clean, one-row-per-spot master table with:
- Spatial coordinates (barcode, pixel positions)
- Sample and cell line metadata
- Hypoxia classification and percentages
- Continuous MP scores (MP1-MP4) from AddModuleScore
- Dominant metaprogram with confidence metrics
- Global MP composition entropy (sample-level diversity)

**This is a FOUNDATION file.** Detailed spatial analysis (neighborhoods, gradients, domains, heterogeneity interpretation) happens in File 2.

**Output:** 
- `master_spot_table.csv` — one row per spot, ready for File 2
- `mp_hypoxia_analysis_results.rds` — all objects for reference

---

# Section 1: Setup & Data Loading

```{r load-libraries}
library(tidyverse)
library(Seurat)
library(ggplot2)
library(patchwork)
library(readr)
```

## Load Consensus Metaprograms

```{r load-metaprograms}
# Load final consensus MP definitions
source(here::here("config.R"))  # defines DATA_DIR, RESULTS_DIR, ANALYSIS_SUBTYPES_DIR
analysis_subtypes_dir <- ANALYSIS_SUBTYPES_DIR
results_dir <- file.path(analysis_subtypes_dir, "data", "parallel_results")
out_dir <- file.path(results_dir, "integrated")

metaprograms_file <- file.path(out_dir, "metaprograms_final.rds")
if (!file.exists(metaprograms_file)) {
  metaprograms_file <- file.path(out_dir, "metaprogram_genes.rds")
}
metaprograms <- readRDS(metaprograms_file)

# Drop MPs with no genes
metaprograms <- metaprograms[sapply(metaprograms, length) > 0]
cat("Loaded", length(metaprograms), "consensus metaprograms (empty MPs removed)\n")
cat("MP names:", names(metaprograms), "\n\n")

# Show MP gene compositions
for (mp_name in names(metaprograms)) {
  genes <- metaprograms[[mp_name]]
  cat(mp_name, ": ", length(genes), " genes\n")
  cat("  ", paste(head(genes, 5), collapse=", "), "...\n\n")
}
```

## Load Hypoxia Data

```{r load-hypoxia}
# Load hypoxia data
hypoxia_file <- file.path(analysis_subtypes_dir, "2_METAPROGRAMS_INTEGRATION", "data", "all_samples_aligned_hypoxia.csv")
hypoxia_data <- read_csv(hypoxia_file, show_col_types = FALSE)
cat("Loaded hypoxia data:", nrow(hypoxia_data), "spots\n")
```

## Compute MP Scores from Seurat Objects

```{r compute-mp-scores}
# Compute MP scores directly from gene expression
mp_score_list <- list()

qc_dir <- file.path(analysis_subtypes_dir, "0_QC_AND_PREPROCESSING", "data")

for (sample_name in unique(hypoxia_data$sample)) {
  seu_file <- file.path(qc_dir, paste0(sample_name, "_qc_with_metadata.rds"))
  if (!file.exists(seu_file)) {
    cat("", sample_name, "- QC file not found:", seu_file, "\n")
    next
  }
  
  cat("Processing:", sample_name, "\n")
  seu <- readRDS(seu_file)
  
  # Extract normalized expression matrix (Seurat v4/v5 compatible)
  mat <- tryCatch(
    GetAssayData(seu, assay = "Spatial", layer = "data"),
    error = function(e) GetAssayData(seu, assay = "Spatial", slot = "data")
  )

  if (nrow(mat) == 0 || ncol(mat) == 0) {
    cat("  Normalizing from counts...\n")
    seu <- NormalizeData(seu, assay = "Spatial", verbose = FALSE)
    mat <- tryCatch(
      GetAssayData(seu, assay = "Spatial", layer = "data"),
      error = function(e) GetAssayData(seu, assay = "Spatial", slot = "data")
    )
  }

  # Compute mean expression of MP genes per spot
  mp_df <- data.frame(row.names = colnames(mat))
  for (mp_name in names(metaprograms)) {
    mp_genes_present <- intersect(metaprograms[[mp_name]], rownames(mat))
    if (length(mp_genes_present) > 5) {
      mp_df[[mp_name]] <- colMeans(mat[mp_genes_present, , drop = FALSE])
    } else {
      mp_df[[mp_name]] <- NA_real_
    }
  }

  mp_df$barcode <- colnames(mat)
  mp_df$sample  <- sample_name
  mp_score_list[[sample_name]] <- mp_df
  cat("  Done:", nrow(mp_df), "spots,", sum(!is.na(mp_df$MP1)), "with MP1 score\n")
}

if (length(mp_score_list) == 0) {
  stop("No QC files found. Check path: ", qc_dir)
}
mp_all_samples <- do.call(rbind, mp_score_list)
rownames(mp_all_samples) <- NULL

# Merge hypoxia data with MP scores
analysis_data <- hypoxia_data |>
  left_join(mp_all_samples, by = c("barcode", "sample")) |>
  as.data.frame()

# Verify cell_line is present (from alignment file)
if (!"cell_line" %in% colnames(analysis_data)) {
  stop("cell_line column not found. Check alignment file structure.")
}

# Assign dominant metaprogram + confidence scores
mp_cols <- intersect(names(metaprograms), colnames(analysis_data))
if(length(mp_cols) > 0) {
  mp_matrix <- as.matrix(analysis_data[, mp_cols])
  
  # Top score — guard against all-NA rows (which.max returns integer(0) → apply returns list)
  max_idx <- apply(mp_matrix, 1, function(x) {
    idx <- which.max(x)
    if (length(idx) == 0L) NA_integer_ else as.integer(idx)
  })
  analysis_data$dominant_metaprogram <- colnames(mp_matrix)[max_idx]
  analysis_data$top_score <- apply(mp_matrix, 1, function(x) if (all(is.na(x))) NA_real_ else max(x, na.rm = TRUE))
  
  # Second highest score
  second_idx <- apply(mp_matrix, 1, function(x) {
    sorted_idx <- order(x, decreasing = TRUE)
    if(length(sorted_idx) > 1) return(sorted_idx[2]) else return(1)
  })
  analysis_data$second_score <- mp_matrix[cbind(1:nrow(mp_matrix), second_idx)]
  
  # Dominance margin: confidence in dominant call
  analysis_data$dominance_margin <- analysis_data$top_score - analysis_data$second_score
}

cat("Computed MP scores from Seurat objects\n")
cat("Dimensions:", nrow(analysis_data), "spots ×", ncol(analysis_data), "variables\n")
cat("Samples:", n_distinct(analysis_data$sample), "\n")
cat("Cell lines:", n_distinct(analysis_data$cell_line), "\n")
cat("Spots with MP assignment:", sum(!is.na(analysis_data$dominant_metaprogram)), "\n")
cat("Mean dominance margin:", round(mean(analysis_data$dominance_margin, na.rm=TRUE), 3), "\n\n")
```

---

# Section 2: Metaprogram Distribution

## MP Presence & Abundance

```{r mp-distribution}
# For each sample: count spots per program
mp_distribution <- analysis_data |>
  filter(!is.na(dominant_metaprogram)) |>
  group_by(sample) |>
  summarize(
    n_spots = n(),
    mp1_count = sum(dominant_metaprogram == "MP1"),
    mp2_count = sum(dominant_metaprogram == "MP2"),
    mp3_count = sum(dominant_metaprogram == "MP3"),
    mp4_count = sum(dominant_metaprogram == "MP4"),
    mp5_count = sum(dominant_metaprogram == "MP5"),
    mp1_pct = 100 * mp1_count / n_spots,
    mp2_pct = 100 * mp2_count / n_spots,
    mp3_pct = 100 * mp3_count / n_spots,
    mp4_pct = 100 * mp4_count / n_spots,
    mp5_pct = 100 * mp5_count / n_spots,
    cell_line = first(cell_line),
    .groups = "drop"
  ) |>
  arrange(desc(mp1_pct + mp2_pct))

cat("METAPROGRAM DISTRIBUTION BY SAMPLE\n")
cat("====================================\n\n")
cat("Samples with complete data:", n_distinct(mp_distribution$sample), "\n")
cat("Cell lines: ", paste(unique(mp_distribution$cell_line), collapse = ", "), "\n\n")
print(mp_distribution |> dplyr::select(sample, n_spots, mp1_pct, mp2_pct, mp3_pct, mp4_pct, mp5_pct, cell_line))
```

## Visualization: MP Composition

```{r mp-composition-plot, fig.width=14, fig.height=6}
# Convert to long format for stacked bar plot
mp_comp_long <- mp_distribution |>
  dplyr::select(sample, mp1_pct, mp2_pct, mp3_pct, mp4_pct, mp5_pct) |>
  pivot_longer(cols = starts_with("mp"), names_to = "program", values_to = "pct") |>
  mutate(program = sub("_pct", "", program) |> toupper()) |>
  arrange(sample)

# Order samples by MP1 dominance
sample_order <- mp_distribution |> 
  arrange(desc(mp1_pct)) |> 
  pull(sample)

p_mp_comp <- ggplot(mp_comp_long |> mutate(sample = fct_relevel(sample, sample_order)), 
                     aes(x = sample, y = pct, fill = program)) +
  geom_col(alpha = 0.8) +
  coord_flip() +
  labs(
    title = "Metaprogram Composition per Sample",
    x = "Sample",
    y = "% of Spots",
    fill = "Program"
  ) +
  scale_fill_manual(values = c("MP1" = "#0072B2", "MP2" = "#E69F00", "MP3" = "#009E73", "MP4" = "#CC79A7", "MP5" = "#D55E00")) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 13))

p_mp_comp
```



---

# Section 4: Data Quality Summary

## Spot-level Completeness

```{r data-quality}
cat("SPOT-LEVEL DATA QUALITY\n")
cat("==========================\n\n")

cat("MP Score Completeness:\n")
mp_missing <- analysis_data |>
  summarise(
    complete_mp1 = sum(!is.na(MP1)),
    complete_mp2 = sum(!is.na(MP2)),
    complete_mp3 = sum(!is.na(MP3)),
    complete_mp4 = sum(!is.na(MP4)),
    complete_mp5 = sum(!is.na(MP5)),
    all_five = sum(!is.na(MP1) & !is.na(MP2) & !is.na(MP3) & !is.na(MP4) & !is.na(MP5))
  )

print(mp_missing)

cat("\nDominance Confidence (dominance_margin):\n")
cat("  Mean margin:", round(mean(analysis_data$dominance_margin, na.rm=TRUE), 3), "\n")
cat("  Min margin: ", round(min(analysis_data$dominance_margin, na.rm=TRUE), 3), "\n")
cat("  Max margin: ", round(max(analysis_data$dominance_margin, na.rm=TRUE), 3), "\n")
cat("  (Higher margin = more confident MP call)\n\n")

cat("Hypoxia Data:\n")
cat("  Spots with hypoxia_label:", sum(!is.na(analysis_data$hypoxia_label)), "\n")
cat("  Normoxic (label=0):", sum(analysis_data$hypoxia_label == 0, na.rm=TRUE), "\n")
cat("  Hypoxic (label=1):", sum(analysis_data$hypoxia_label == 1, na.rm=TRUE), "\n")
```

## One Row Per Spot Verification

```{r spot-uniqueness}
# Verify exactly one row per barcode per sample
spot_counts <- analysis_data |>
  group_by(sample, barcode) |>
  summarise(n = n(), .groups = "drop") |>
  filter(n > 1)

if(nrow(spot_counts) > 0) {
  cat(" WARNING: Found", nrow(spot_counts), "barcodes with multiple rows\n")
  print(spot_counts)
} else {
  cat("Each barcode appears exactly once per sample\n")
  cat("   Total unique spots:", n_distinct(analysis_data$barcode), "\n")
  cat("   Across", n_distinct(analysis_data$sample), "samples\n")
}
```

---

# Section 5: Save Master Spot Table & Results

## Create Master Spot Table (One row per spot)

```{r create-master-table}
# Debug: confirm which MP columns are in analysis_data
mp_present <- intersect(c("MP1","MP2","MP3","MP4","MP5"), colnames(analysis_data))
cat("MP columns in analysis_data:", paste(mp_present, collapse=", "), "\n")
if (length(mp_present) < 5) warning("Not all 5 MPs present — check score computation above")

# Use any_of so missing columns don't cause error
master_spot_table <- analysis_data |>
  dplyr::select(any_of(c(
    # Metadata
    "sample", "barcode", "cell_line",
    # Spatial coordinates
    "pxl_row_in_fullres", "pxl_col_in_fullres",
    "array_row", "array_col",
    # Hypoxia classification
    "hypoxia_label", "hypoxia_percentage",
    # MP scores (continuous)
    "MP1", "MP2", "MP3", "MP4", "MP5",
    # Dominance metrics
    "dominant_metaprogram", "top_score", "second_score", "dominance_margin"
  ))) |>
  arrange(sample, barcode)

# Merge pathologist annotations if available (resolve file, with fallback)
annot_file <- file.path(analysis_subtypes_dir, "3_CELL_TYPE_ANNOTATION", "results",
                         "pathologist_labels_all_samples.csv")
if (!file.exists(annot_file)) {
  annot_file <- file.path(ANALYSIS_SUBTYPES_DIR, "3_CELL_TYPE_ANNOTATION/results/pathologist_labels_all_samples.csv")
}
if (file.exists(annot_file)) {
  cat("Merging pathologist annotations...\n")
  annot <- read_csv(annot_file, show_col_types = FALSE) |>
    rename(barcode = spot) |>
    dplyr::select(barcode, sample, pathologist_label)
  master_spot_table <- master_spot_table |>
    left_join(annot, by = c("barcode", "sample"))
  cat("  Annotation coverage:", sum(!is.na(master_spot_table$pathologist_label)),
      "of", nrow(master_spot_table), "spots\n")
  cat("  Labels:", paste(sort(unique(master_spot_table$pathologist_label)), collapse=", "), "\n")
} else {
  cat("No pathologist annotations found — run 3_CELL_TYPE_ANNOTATION/1_assign_pathologist_labels.R first\n")
  master_spot_table$pathologist_label <- NA_character_
}

cat("\nMASTER SPOT TABLE CREATED\n")
cat("============================\n\n")
cat("Dimensions:", nrow(master_spot_table), "spots ×", ncol(master_spot_table), "columns\n")
cat("Columns:", paste(colnames(master_spot_table), collapse=", "), "\n")
```

## Save Master Spot Table as CSV

```{r save-master-table}
output_dir <- file.path(analysis_subtypes_dir, "2_METAPROGRAMS_INTEGRATION", "data")
dir.create(output_dir, showWarnings = FALSE)

master_file <- file.path(output_dir, "master_spot_table.csv")
write_csv(master_spot_table, master_file)

cat("\nSaved master spot table:\n")
cat(master_file, "\n\n")
cat("This table is ready for File 2 (spatial analysis)\n")
cat("Each row is one unique spot with all essential data\n")
```

## Save All Objects to RDS (for reference)

```{r save-rds}
# Bundle all analysis results into a single list
analysis_results <- list(
  master_spot_table = master_spot_table,
  analysis_data = analysis_data,
  metaprograms = metaprograms,
  mp_distribution = mp_distribution,
  mp_entropy_sample = mp_entropy_sample
)

results_file <- file.path(output_dir, "mp_hypoxia_analysis_results.rds")
saveRDS(analysis_results, results_file)

cat("\nAll results saved to:\n")
cat(results_file, "\n\n")
cat("Objects saved for reference:\n")
cat("  - master_spot_table: one-row-per-spot master data (also saved as CSV)\n")
cat("  - analysis_data: full spot-level data with all variables\n")
cat("  - metaprograms: MP gene definitions\n")
cat("  - mp_distribution: MP counts/percentages per sample\n")
cat("  - mp_entropy_sample: global MP composition entropy (sample-level)\n\n")
cat("To load in File 2:\n")
cat("  results <- readRDS('mp_hypoxia_analysis_results.rds')\n")
cat("  master_spot_table <- results$master_spot_table\n")
cat("  # OR directly from CSV:\n")
cat("  master_spot_table <- read_csv('master_spot_table.csv')\n")
```

