---
title: "Metaprogram Gene Set Enrichment Analysis"
author: "Local Analysis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 8)
```

# Overview

This notebook performs gene set enrichment analysis on the metaprograms identified from the hypoxia Visium data.

**Input required:** `metaprograms_final.rds` file from the main integration analysis.

---

# Setup

```{r install-packages, eval=FALSE}
# Install packages if needed (run once)
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# Update ggplot2 first (required version >= 3.5.2)
install.packages("ggplot2")

# Install Bioconductor packages
if (!requireNamespace("clusterProfiler", quietly = TRUE)) {
  BiocManager::install("clusterProfiler")
}
if (!requireNamespace("enrichplot", quietly = TRUE)) {
  BiocManager::install("enrichplot")
}
if (!requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
  BiocManager::install("org.Hs.eg.db")
}

# Install CRAN packages
if (!requireNamespace("msigdbr", quietly = TRUE)) {
  install.packages("msigdbr")
}
if (!requireNamespace("dplyr", quietly = TRUE)) {
  install.packages("dplyr")
}
```

```{r load-libraries}
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
library(msigdbr)
library(dplyr)
library(ggplot2)

cat("Libraries loaded\n")
```

```{r set-directories}
source(here::here("config.R"))  # defines DATA_DIR, RESULTS_DIR, ANALYSIS_SUBTYPES_DIR
analysis_subtypes_dir <- ANALYSIS_SUBTYPES_DIR

mp_labels <- c(
  MP1 = "MP1 — EMT/Mesenchymal",
  MP2 = "MP2 — Hypoxia Response",
  MP3 = "MP3 — Squamous Differentiation",
  MP4 = "MP4 — ECM/Stromal",
  MP5 = "MP5 — Keratinocyte/Squamous-epithelial"
)

input_file <- file.path(analysis_subtypes_dir, "data", "parallel_results", "integrated", "metaprograms_final.rds")

out_dir <- file.path(analysis_subtypes_dir, "4_METAPROGRAMS_SPATIAL_ANALYSIS", "results")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

cat("Input file:", input_file, "\n")
cat("Output directory:", out_dir, "\n")
```

---

# Load Metaprograms

```{r load-metaprograms}
cat("=== Loading metaprograms ===\n\n")

# Load the metaprograms
metaprograms <- readRDS(input_file)

cat("Number of metaprograms:", length(metaprograms), "\n\n")

# Display summary
for (mp_name in names(metaprograms)) {
  cat(mp_name, ":", length(metaprograms[[mp_name]]), "genes\n")
  cat("  Top genes:", paste(head(metaprograms[[mp_name]], 10), collapse = ", "), "\n\n")
}
```

---

# Gene Set Enrichment Analysis

```{r gsea-analysis}
cat("=== Running Gene Set Enrichment Analysis ===\n\n")

# Store all enrichment results
enrichment_results <- list()

for (mp_name in names(metaprograms)) {
  cat("\n### Analyzing", mp_name, "###\n")
  
  genes <- metaprograms[[mp_name]]
  cat("  Total genes:", length(genes), "\n")
  
  # Convert gene symbols to Entrez IDs
  gene_entrez <- bitr(
    genes, 
    fromType = "SYMBOL",
    toType = "ENTREZID", 
    OrgDb = org.Hs.eg.db
  )
  
  cat("  Mapped to Entrez:", nrow(gene_entrez), "\n")
  
  # 1. GO Biological Process
  cat("  Running GO Biological Process...\n")
  go_bp <- enrichGO(
    gene = gene_entrez$ENTREZID,
    OrgDb = org.Hs.eg.db,
    ont = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.2,
    readable = TRUE
  )
  
  # 2. GO Molecular Function
  cat("  Running GO Molecular Function...\n")
  go_mf <- enrichGO(
    gene = gene_entrez$ENTREZID,
    OrgDb = org.Hs.eg.db,
    ont = "MF",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.2,
    readable = TRUE
  )
  
  # 3. GO Cellular Component
  cat("  Running GO Cellular Component...\n")
  go_cc <- enrichGO(
    gene = gene_entrez$ENTREZID,
    OrgDb = org.Hs.eg.db,
    ont = "CC",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.2,
    readable = TRUE
  )
  
  # 4. KEGG Pathways
  cat("  Running KEGG Pathways...\n")
  kegg <- enrichKEGG(
    gene = gene_entrez$ENTREZID,
    organism = "hsa",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.2
  )
  
  # 5. Hallmark Gene Sets (MSigDB)
  cat("  Running Hallmark Gene Sets...\n")
  hallmark_sets <- msigdbr(species = "Homo sapiens", category = "H")
  hallmark_t2g <- hallmark_sets %>% dplyr::select(gs_name, gene_symbol)
  
  hallmark <- enricher(
    gene = genes,
    TERM2GENE = hallmark_t2g,
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.2
  )
  
  # Store results
  enrichment_results[[mp_name]] <- list(
    GO_BP = go_bp,
    GO_MF = go_mf,
    GO_CC = go_cc,
    KEGG = kegg,
    Hallmark = hallmark,
    genes = genes,
    gene_entrez = gene_entrez
  )
  
  cat("  Complete\n")
}

cat("\nAll metaprograms analyzed\n\n")
```

```{r save-results}
# Save all enrichment results
saveRDS(enrichment_results, file.path(out_dir, "enrichment_results.rds"))
cat("Results saved to:", file.path(out_dir, "enrichment_results.rds"), "\n\n")
```

---

# Enrichment Summary

```{r summary-tables}
cat("=== Enrichment Summary ===\n\n")

for (mp_name in names(enrichment_results)) {
  cat("\n## ", mp_labels[mp_name], " ##\n\n", sep = "")
  
  results <- enrichment_results[[mp_name]]
  
  # GO Biological Process
  if (!is.null(results$GO_BP) && nrow(results$GO_BP@result) > 0) {
    cat("### GO Biological Process (Top 10) ###\n")
    top_bp <- head(results$GO_BP@result[, c("Description", "pvalue", "p.adjust", "Count")], 10)
    print(knitr::kable(top_bp, row.names = FALSE, digits = 4))
    cat("\n")
  } else {
    cat("### GO Biological Process ###\nNo significant terms\n\n")
  }
  
  # GO Molecular Function
  if (!is.null(results$GO_MF) && nrow(results$GO_MF@result) > 0) {
    cat("### GO Molecular Function (Top 10) ###\n")
    top_mf <- head(results$GO_MF@result[, c("Description", "pvalue", "p.adjust", "Count")], 10)
    print(knitr::kable(top_mf, row.names = FALSE, digits = 4))
    cat("\n")
  } else {
    cat("### GO Molecular Function ###\nNo significant terms\n\n")
  }
  
  # KEGG Pathways
  if (!is.null(results$KEGG) && nrow(results$KEGG@result) > 0) {
    cat("### KEGG Pathways (Top 10) ###\n")
    top_kegg <- head(results$KEGG@result[, c("Description", "pvalue", "p.adjust", "Count")], 10)
    print(knitr::kable(top_kegg, row.names = FALSE, digits = 4))
    cat("\n")
  } else {
    cat("### KEGG Pathways ###\nNo significant pathways\n\n")
  }
  
  # Hallmark Gene Sets
  if (!is.null(results$Hallmark) && nrow(results$Hallmark@result) > 0) {
    cat("### Hallmark Gene Sets (Top 10) ###\n")
    top_hallmark <- head(results$Hallmark@result[, c("ID", "pvalue", "p.adjust", "Count")], 10)
    print(knitr::kable(top_hallmark, row.names = FALSE, digits = 4))
    cat("\n")
  } else {
    cat("### Hallmark Gene Sets ###\nNo significant gene sets\n\n")
  }
  
  cat("\n---\n\n")
}
```

---

# Visualizations

## Dotplots

```{r dotplots, fig.width=12, fig.height=8}
cat("=== Creating dotplots ===\n\n")

for (mp_name in names(enrichment_results)) {
  cat("Plotting", mp_name, "\n")
  
  results <- enrichment_results[[mp_name]]
  
  # GO BP Dotplot
  if (!is.null(results$GO_BP) && nrow(results$GO_BP@result) > 0) {
    # Filter for significant results
    sig_results <- results$GO_BP@result[results$GO_BP@result$p.adjust < 0.05, ]
    if (nrow(sig_results) > 0) {
      tryCatch({
        p1 <- dotplot(results$GO_BP, showCategory = 15, title = paste(mp_labels[mp_name], "- GO Biological Process"))
        print(p1)
        
        ggsave(
          filename = file.path(out_dir, paste0(mp_name, "_GO_BP_dotplot.pdf")),
          plot = p1,
          width = 10,
          height = 8
        )
        cat("  GO BP dotplot saved\n")
      }, error = function(e) {
        cat("  Skipping GO BP dotplot:", conditionMessage(e), "\n")
      })
    } else {
      cat("  No significant GO BP results to plot\n")
    }
  }
  
  # KEGG Dotplot
  if (!is.null(results$KEGG) && nrow(results$KEGG@result) > 0) {
    sig_results <- results$KEGG@result[results$KEGG@result$p.adjust < 0.05, ]
    if (nrow(sig_results) > 0) {
      tryCatch({
        p2 <- dotplot(results$KEGG, showCategory = 15, title = paste(mp_labels[mp_name], "- KEGG Pathways"))
        print(p2)
        
        ggsave(
          filename = file.path(out_dir, paste0(mp_name, "_KEGG_dotplot.pdf")),
          plot = p2,
          width = 10,
          height = 8
        )
        cat("  KEGG dotplot saved\n")
      }, error = function(e) {
        cat("  Skipping KEGG dotplot:", conditionMessage(e), "\n")
      })
    } else {
      cat("  No significant KEGG results to plot\n")
    }
  }
  
  # Hallmark Dotplot
  if (!is.null(results$Hallmark) && nrow(results$Hallmark@result) > 0) {
    sig_results <- results$Hallmark@result[results$Hallmark@result$p.adjust < 0.05, ]
    if (nrow(sig_results) > 0) {
      tryCatch({
        p3 <- dotplot(results$Hallmark, showCategory = 15, title = paste(mp_labels[mp_name], "- Hallmark Gene Sets"))
        print(p3)
        
        ggsave(
          filename = file.path(out_dir, paste0(mp_name, "_Hallmark_dotplot.pdf")),
          plot = p3,
          width = 10,
          height = 8
        )
        cat("  Hallmark dotplot saved\n")
      }, error = function(e) {
        cat("  Skipping Hallmark dotplot:", conditionMessage(e), "\n")
      })
    } else {
      cat("  No significant Hallmark results to plot\n")
    }
  }
}

cat("\nDotplots completed\n\n")
```

## Bar Plots

```{r barplots, fig.width=10, fig.height=6}
cat("=== Creating barplots ===\n\n")

for (mp_name in names(enrichment_results)) {
  cat("Barplots for", mp_name, "\n")
  results <- enrichment_results[[mp_name]]
  
  # Hallmark barplot
  if (!is.null(results$Hallmark) && nrow(results$Hallmark@result) > 0) {
    sig_results <- results$Hallmark@result[results$Hallmark@result$p.adjust < 0.05, ]
    if (nrow(sig_results) > 0) {
      tryCatch({
        p <- barplot(results$Hallmark, showCategory = 10, title = paste(mp_labels[mp_name], "- Top Hallmark Gene Sets"))
        print(p)
        
        ggsave(
          filename = file.path(out_dir, paste0(mp_name, "_Hallmark_barplot.pdf")),
          plot = p,
          width = 10,
          height = 6
        )
        cat("  Hallmark barplot saved\n")
      }, error = function(e) {
        cat("  Skipping Hallmark barplot:", conditionMessage(e), "\n")
      })
    } else {
      cat("  No significant Hallmark results to plot\n")
    }
  }
  
  # GO BP barplot
  if (!is.null(results$GO_BP) && nrow(results$GO_BP@result) > 0) {
    sig_results <- results$GO_BP@result[results$GO_BP@result$p.adjust < 0.05, ]
    if (nrow(sig_results) > 0) {
      tryCatch({
        p <- barplot(results$GO_BP, showCategory = 10, title = paste(mp_labels[mp_name], "- GO Biological Process"))
        print(p)
        
        ggsave(
          filename = file.path(out_dir, paste0(mp_name, "_GO_BP_barplot.pdf")),
          plot = p,
          width = 10,
          height = 6
        )
        cat("  GO BP barplot saved\n")
      }, error = function(e) {
        cat("  Skipping GO BP barplot:", conditionMessage(e), "\n")
      })
    } else {
      cat("  No significant GO BP results to plot\n")
    }
  }
}

cat("\nBarplots completed\n\n")
```



## Enrichment Map

```{r enrichment-maps, fig.width=12, fig.height=10}
cat("=== Creating enrichment maps ===\n\n")

for (mp_name in names(enrichment_results)) {
  cat("Enrichment map for", mp_name, "\n")
  results <- enrichment_results[[mp_name]]
  
  # Hallmark enrichment map - with reduced complexity
  if (!is.null(results$Hallmark) && nrow(results$Hallmark@result) > 0) {
    sig_results <- results$Hallmark@result[results$Hallmark@result$p.adjust < 0.05, ]
    
    if (nrow(sig_results) >= 2) {
      tryCatch({
        # Limit to top 10 most significant terms to reduce memory usage
        top_terms <- head(sig_results[order(sig_results$p.adjust), ], 10)
        
        # Create a subset enrichment result object
        hallmark_subset <- results$Hallmark
        hallmark_subset@result <- hallmark_subset@result[hallmark_subset@result$ID %in% top_terms$ID, ]
        
        # Pairwise term similarity on subset
        hallmark_subset <- pairwise_termsim(hallmark_subset)
        
        # Create enrichment map with reduced complexity
        p <- emapplot(hallmark_subset, 
                      showCategory = min(10, nrow(hallmark_subset@result)), 
                      cex_label_category = 0.6,
                      layout = "kk") +  # Use Kamada-Kawai layout (less memory intensive)
          ggtitle(paste(mp_name, "- Hallmark Enrichment Map (Top 10)"))
        
        print(p)
        
        ggsave(
          filename = file.path(out_dir, paste0(mp_name, "_Hallmark_emap.pdf")),
          plot = p,
          width = 12,
          height = 10
        )
        cat("  Enrichment map saved (top 10 terms)\n")
      }, error = function(e) {
        cat("  Skipping enrichment map:", conditionMessage(e), "\n")
        cat("    Try reducing the number of gene sets or increasing memory allocation\n")
      })
    } else {
      cat("  Not enough significant Hallmark results for enrichment map (need ≥2)\n")
    }
  }
}

cat("\nEnrichment maps completed\n\n")
```

---

# Export Results to CSV

```{r export-csv}
cat("=== Exporting results to CSV ===\n\n")

for (mp_name in names(enrichment_results)) {
  results <- enrichment_results[[mp_name]]
  
  # GO BP
  if (!is.null(results$GO_BP) && nrow(results$GO_BP@result) > 0) {
    write.csv(
      results$GO_BP@result,
      file.path(out_dir, paste0(mp_name, "_GO_BP_enrichment.csv")),
      row.names = FALSE
    )
  }
  
  # GO MF
  if (!is.null(results$GO_MF) && nrow(results$GO_MF@result) > 0) {
    write.csv(
      results$GO_MF@result,
      file.path(out_dir, paste0(mp_name, "_GO_MF_enrichment.csv")),
      row.names = FALSE
    )
  }
  
  # GO CC
  if (!is.null(results$GO_CC) && nrow(results$GO_CC@result) > 0) {
    write.csv(
      results$GO_CC@result,
      file.path(out_dir, paste0(mp_name, "_GO_CC_enrichment.csv")),
      row.names = FALSE
    )
  }
  
  # KEGG
  if (!is.null(results$KEGG) && nrow(results$KEGG@result) > 0) {
    write.csv(
      results$KEGG@result,
      file.path(out_dir, paste0(mp_name, "_KEGG_enrichment.csv")),
      row.names = FALSE
    )
  }
  
  # Hallmark
  if (!is.null(results$Hallmark) && nrow(results$Hallmark@result) > 0) {
    write.csv(
      results$Hallmark@result,
      file.path(out_dir, paste0(mp_name, "_Hallmark_enrichment.csv")),
      row.names = FALSE
    )
  }
  
}

cat("Enrichment tables exported to:", out_dir, "\n")
cat("  Files: <MP>_<database>_enrichment.csv\n\n")
```

---

# Export Gene Lists

```{r export-gene-lists}
cat("=== Exporting gene lists ===\n\n")

for (mp_name in names(metaprograms)) {
  gene_df <- data.frame(
    Metaprogram = mp_name,
    Rank = 1:length(metaprograms[[mp_name]]),
    Gene = metaprograms[[mp_name]]
  )
  
  write.csv(
    gene_df,
    file.path(out_dir, paste0(mp_name, "_genes.csv")),
    row.names = FALSE
  )
}

cat("Gene lists exported to:", out_dir, "\n\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```

---

# Summary

**Analysis Complete! **

All enrichment results have been saved to: **`r out_dir`**

**Output files:**
- `enrichment_results.rds` - Complete R object with all results
- `<MP>_GO_BP_enrichment.csv` - GO Biological Process results
- `<MP>_GO_MF_enrichment.csv` - GO Molecular Function results
- `<MP>_GO_CC_enrichment.csv` - GO Cellular Component results
- `<MP>_KEGG_enrichment.csv` - KEGG pathway results
- `<MP>_Hallmark_enrichment.csv` - Hallmark gene set results
- `<MP>_genes.csv` - Complete gene lists per metaprogram
- Various PDF plots (dotplots, barplots, networks, enrichment maps)

**Databases analyzed:**
1. GO Biological Process
2. GO Molecular Function
3. GO Cellular Component
4. KEGG Pathways
5. Hallmark Gene Sets (MSigDB)
