---
title: "Metaprogram Heterogeneity: Cell Line × Hypoxia Analysis"
subtitle: "Intra and Inter-cell Line Heterogeneity in Spatial Metaprogram Organization"

date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 8)
```

# Overview

Using NMF loading-weighted metaprograms to show:
1. **Inter-cell line heterogeneity**: Different cell lines have distinct MP composition patterns
2. **Intra-cell line heterogeneity**: Hypoxia modulates MP distribution within cell lines


---

# Setup

```{r libraries}
library(dplyr)
library(ggplot2)
library(tidyr)
library(RColorBrewer)
library(cowplot)
library(ggpubr)
library(pheatmap)
```

```{r load-data}
# Resolve paths relative to this Rmd file
rmd_dir            <- dirname(knitr::current_input(dir = TRUE))
analysis_dir       <- dirname(rmd_dir)
master_spot_path   <- file.path(analysis_dir, "2_METAPROGRAMS_INTEGRATION", "data", "master_spot_table.csv")
gene_ranked_path   <- file.path(analysis_dir, "2_METAPROGRAMS_INTEGRATION", "data", "metaprogram_all_genes_ranked.csv")

# Create output folders — Writing folder lives two levels above Analysis_subtypes
base_dir    <- dirname(dirname(analysis_dir))
figures_dir <- file.path(base_dir, "Writing", "Heterogeneity_Figures")
results_dir <- file.path(rmd_dir, "results")
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)

# Load master spot table
master_spot_table <- read.csv(master_spot_path)

# Convert hypoxia_label 0/1 → "Normoxic"/"Hypoxic" for readable facets and pivots
master_spot_table <- master_spot_table %>%
  mutate(hypoxia_label = case_when(
    hypoxia_label == 1 ~ "Hypoxic",
    hypoxia_label == 0 ~ "Normoxic",
    TRUE ~ NA_character_
  ))

cat("Loaded master spot table:", nrow(master_spot_table), "spots\n")
cat("  Cell lines:", n_distinct(master_spot_table$cell_line), "\n")
cat("  Samples:", n_distinct(master_spot_table$sample), "\n")

cat("\nMetaprogram assignment summary:\n")
print(table(master_spot_table$dominant_metaprogram, useNA = "ifany"))

cat("\nHypoxia label summary:\n")
print(table(master_spot_table$hypoxia_label, useNA = "ifany"))
```

---

# FIGURE 1: Inter-Cell Line Heterogeneity

Metaprogram composition differs dramatically by cell line.

```{r figure1-cellline-composition}
mp_by_cellline <- master_spot_table %>%
  filter(!is.na(dominant_metaprogram)) %>%
  group_by(cell_line, dominant_metaprogram) %>%
  summarise(n_spots = n(), .groups = "drop") %>%
  group_by(cell_line) %>%
  mutate(
    total_spots = sum(n_spots),
    pct = 100 * n_spots / total_spots
  ) %>%
  ungroup()

cat("\n=== Metaprogram Composition by Cell Line ===\n\n")
print(mp_by_cellline)

p1 <- ggplot(mp_by_cellline, aes(x = cell_line, y = pct, fill = dominant_metaprogram)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.3) +
  scale_fill_manual(
    values = c(MP1 = "#0072B2", MP2 = "#E69F00", MP3 = "#009E73", MP4 = "#CC79A7", MP5 = "#D55E00"),
    name = "Metaprogram"
  ) +
  labs(
    title = "INTER-CELL LINE HETEROGENEITY: Metaprogram Composition",
    x = "Cell Line",
    y = "% Spots",
    subtitle = "Different cell lines exhibit distinct MP profiles"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    legend.position = "right"
  )

p1

ggsave(
  file.path(figures_dir, "FIGURE1_InterCellLine_MPComposition.pdf"),
  plot = p1,
  width = 10, height = 6
)
cat("FIGURE 1 saved\n")
```

---

# FIGURE 2: Intra-Cell Line Heterogeneity × Hypoxia

How hypoxia modulates MP distribution within cell lines.

```{r figure2-hypoxia-effect}
mp_by_cellline_hypoxia <- master_spot_table %>%
  filter(!is.na(dominant_metaprogram), !is.na(hypoxia_label)) %>%
  group_by(cell_line, hypoxia_label, dominant_metaprogram) %>%
  summarise(n_spots = n(), .groups = "drop") %>%
  group_by(cell_line, hypoxia_label) %>%
  mutate(
    total_spots = sum(n_spots),
    pct = 100 * n_spots / total_spots
  ) %>%
  ungroup()

cat("\n=== Metaprogram Composition by Cell Line × Hypoxia ===\n\n")
print(mp_by_cellline_hypoxia)

p2 <- ggplot(
  mp_by_cellline_hypoxia,
  aes(x = cell_line, y = pct, fill = dominant_metaprogram)
) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.3) +
  facet_wrap(~ hypoxia_label, nrow = 1, labeller = labeller(
    hypoxia_label = c(
      "Hypoxic"  = "HYPOXIC REGIONS",
      "Normoxic" = "NORMOXIC REGIONS"
    )
  )) +
  scale_fill_manual(
    values = c(MP1 = "#0072B2", MP2 = "#E69F00", MP3 = "#009E73", MP4 = "#CC79A7", MP5 = "#D55E00"),
    name = "Metaprogram"
  ) +
  labs(
    title = "INTRA-CELL LINE HETEROGENEITY: Hypoxia-Driven MP Shifts",
    x = "Cell Line",
    y = "% Spots",
    subtitle = "MP1 & MP4 enrich in hypoxia; MP2 & MP3 deplete"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
    plot.title = element_text(size = 14, face = "bold", hjust = 0),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    strip.text = element_text(size = 11, face = "bold"),
    legend.position = "bottom"
  )

p2

ggsave(
  file.path(figures_dir, "FIGURE2_IntraCellLine_Hypoxia_MPShifts.pdf"),
  plot = p2,
  width = 12, height = 6
)
cat("FIGURE 2 saved\n")
```

---

# Quantitative Analysis: Statistical Summary

```{r stats-summary}
cat("\n=== INTRA-CELL LINE HETEROGENEITY: Hypoxia Effect Size ===\n\n")

mp_shift <- master_spot_table %>%
  filter(!is.na(dominant_metaprogram), !is.na(hypoxia_label)) %>%
  group_by(cell_line, hypoxia_label, dominant_metaprogram) %>%
  summarise(n_spots = n(), .groups = "drop") %>%
  group_by(cell_line, hypoxia_label) %>%
  mutate(pct = 100 * n_spots / sum(n_spots)) %>%
  ungroup() %>%
  pivot_wider(
    id_cols     = c(cell_line, dominant_metaprogram),
    names_from  = hypoxia_label,
    values_from = pct,
    values_fill = 0
  ) %>%
  mutate(
    shift     = Hypoxic - Normoxic,
    abs_shift = abs(shift)
  ) %>%
  arrange(cell_line, desc(abs_shift))

print(mp_shift)

mp_shift_summary <- mp_shift %>%
  group_by(dominant_metaprogram) %>%
  summarise(
    mean_shift     = mean(shift, na.rm = TRUE),
    mean_abs_shift = mean(abs_shift, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_abs_shift))

cat("\n=== Mean Hypoxia Effect per Metaprogram ===\n")
cat("(Positive = enriched in Hypoxia, Negative = enriched in Normoxia)\n\n")
print(mp_shift_summary)
```

---

# FIGURE 3: Heatmap — Cell Line × Metaprogram Composition

```{r figure3-heatmap}
comp_matrix <- master_spot_table %>%
  filter(!is.na(dominant_metaprogram)) %>%
  group_by(cell_line, dominant_metaprogram) %>%
  summarise(n_spots = n(), .groups = "drop") %>%
  group_by(cell_line) %>%
  mutate(pct = 100 * n_spots / sum(n_spots)) %>%
  ungroup() %>%
  pivot_wider(
    id_cols     = cell_line,
    names_from  = dominant_metaprogram,
    values_from = pct,
    values_fill = 0
  ) %>%
  tibble::column_to_rownames("cell_line") %>%
  as.matrix()

# Reorder columns MP1–MP5
comp_matrix <- comp_matrix[, intersect(c("MP1","MP2","MP3","MP4","MP5"), colnames(comp_matrix))]

pdf(
  file.path(figures_dir, "FIGURE3_CellLine_MPComposition_Heatmap.pdf"),
  width = 10, height = 6
)
pheatmap(
  comp_matrix,
  color            = colorRampPalette(c("white", "yellow", "orange", "red"))(50),
  cluster_rows     = TRUE,
  cluster_cols     = FALSE,
  display_numbers  = TRUE,
  number_format    = "%.1f",
  fontsize         = 12,
  fontsize_number  = 10,
  main             = "Inter-Cell Line Heterogeneity\nMetaprogram % Composition",
  border_color     = "gray40"
)
dev.off()

cat("FIGURE 3 saved\n")
```

---

# FIGURE 4: Sample-Level Hypoxia Shift

Quantitative shift in metaprogram abundance per sample (% Hypoxic − % Normoxic).

```{r figure4-sample-level-shifts}
shift_long <- master_spot_table %>%
  filter(!is.na(dominant_metaprogram), !is.na(hypoxia_label)) %>%
  group_by(sample, hypoxia_label, dominant_metaprogram) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(sample, hypoxia_label) %>%
  mutate(pct = 100 * count / sum(count)) %>%
  ungroup()

shift_wide <- shift_long %>%
  pivot_wider(
    id_cols     = c(sample, dominant_metaprogram),
    names_from  = hypoxia_label,
    values_from = pct,
    values_fill = 0
  ) %>%
  mutate(Shift = Hypoxic - Normoxic)

p4 <- ggplot(shift_wide, aes(x = dominant_metaprogram, y = Shift, fill = Shift > 0)) +
  geom_col(color = "black", linewidth = 0.2, alpha = 0.85) +
  facet_wrap(~sample, nrow = 4) +
  scale_fill_manual(
    values = c("TRUE" = "#2E86AB", "FALSE" = "#A23B72"),
    labels = c("TRUE" = "Hypoxia-Enriched", "FALSE" = "Normoxia-Enriched"),
    guide  = "legend"
  ) +
  labs(
    title    = "Sample-Level Metaprogram Hypoxia Sensitivity",
    subtitle = "Shift = % in Hypoxic regions − % in Normoxic regions\nPositive (blue) = hypoxia-enriched | Negative (purple) = normoxia-enriched",
    x        = "Metaprogram",
    y        = "Shift (percentage points)",
    fill     = "Enrichment"
  ) +
  theme_minimal() +
  theme(
    plot.title         = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle      = element_text(size = 11, hjust = 0.5, color = "gray40"),
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text          = element_text(size = 9),
    strip.text         = element_text(size = 9, face = "bold"),
    legend.position    = "bottom",
    legend.title       = element_text(face = "bold")
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.8)

ggsave(
  file.path(figures_dir, "FIGURE4_Sample_Level_Hypoxia_Shifts.pdf"),
  plot = p4, width = 16, height = 12, dpi = 300
)
print(p4)
cat("FIGURE 4 saved\n")

# Summary statistics
shift_stats <- shift_wide %>%
  group_by(dominant_metaprogram) %>%
  summarise(
    Mean_Shift   = mean(Shift),
    Median_Shift = median(Shift),
    SD_Shift     = sd(Shift),
    .groups = "drop"
  ) %>%
  arrange(desc(Mean_Shift))

cat("\n=== Sample-Level Shift Summary ===\n\n")
print(shift_stats)
```

---

# FIGURE 5: Intra-Sample Composition (Hypoxic vs Normoxic)

```{r figure5-sample-composition}
sample_comp <- master_spot_table %>%
  filter(!is.na(dominant_metaprogram), !is.na(hypoxia_label)) %>%
  group_by(sample, hypoxia_label, dominant_metaprogram) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(sample, hypoxia_label) %>%
  mutate(pct = 100 * count / sum(count)) %>%
  ungroup()

norm_data <- sample_comp %>% filter(hypoxia_label == "Normoxic")
hyp_data  <- sample_comp %>% filter(hypoxia_label == "Hypoxic")

mp_fill <- scale_fill_manual(
  values = c(MP1 = "#0072B2", MP2 = "#E69F00", MP3 = "#009E73", MP4 = "#CC79A7", MP5 = "#D55E00"),
  name   = "Metaprogram"
)

p_norm <- ggplot(norm_data, aes(x = reorder(sample, -pct), y = pct, fill = dominant_metaprogram)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.2) +
  mp_fill +
  labs(title = "Normoxic Regions", x = "Sample", y = "% Spots") +
  ylim(0, 100) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
        plot.title  = element_text(face = "bold", size = 12))

p_hyp <- ggplot(hyp_data, aes(x = reorder(sample, -pct), y = pct, fill = dominant_metaprogram)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.2) +
  mp_fill +
  labs(title = "Hypoxic Regions", x = "Sample", y = "% Spots") +
  ylim(0, 100) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
        plot.title  = element_text(face = "bold", size = 12))

fig5 <- plot_grid(p_norm, p_hyp, nrow = 2, labels = c("A", "B"), label_size = 12)

ggsave(
  file.path(figures_dir, "FIGURE5_IntraSample_Hypoxia_MPComposition.pdf"),
  plot = fig5, width = 16, height = 10, dpi = 300
)
print(fig5)
cat("FIGURE 5 saved\n")
```

---

# Signature Genes per Metaprogram

```{r signature-genes}
if (file.exists(gene_ranked_path)) {
  genes_df <- read.csv(gene_ranked_path)

  signature_genes <- genes_df %>%
    filter(Metaprogram %in% c("MP1","MP2","MP3","MP4","MP5")) %>%
    group_by(Metaprogram) %>%
    arrange(Rank) %>%
    summarise(genes = list(Gene), .groups = "drop") %>%
    tibble::deframe()

  hypoxia_mps  <- c("MP1", "MP4")
  normoxia_mps <- c("MP2", "MP3")

  for (mp in names(signature_genes)) {
    assoc <- if (mp %in% hypoxia_mps) "HYPOXIA-CORRELATED" else "NORMOXIA-PREFERRED"
    cat(sprintf("\n%s | %s | %d genes\n", mp, assoc, length(signature_genes[[mp]])))
    cat("Top 10:", paste(head(signature_genes[[mp]], 10), collapse = ", "), "\n")
  }

  sig_table <- genes_df %>%
    filter(Metaprogram %in% c("MP1","MP2","MP3","MP4","MP5")) %>%
    mutate(Hypoxia_Association = ifelse(Metaprogram %in% hypoxia_mps,
                                        "HYPOXIA-CORR", "NORMOXIA-PREFER"))

  write.csv(sig_table,
            file.path(results_dir, "Signature_Genes_Summary.csv"),
            row.names = FALSE)
  cat("\nSignature genes table saved\n")
} else {
  cat("Gene ranked file not found:", gene_ranked_path, "\n")
  cat("  Skipping signature gene section.\n")
}
```

---

# Summary Table

```{r summary-heterogeneity}
# Compute from data — not hardcoded
dominant_mp_per_cl <- master_spot_table %>%
  filter(!is.na(dominant_metaprogram)) %>%
  group_by(cell_line, dominant_metaprogram) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(cell_line) %>%
  slice_max(n, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  rename(Dominant_MP = dominant_metaprogram) %>%
  select(cell_line, Dominant_MP)

mp1_shift_per_cl <- mp_shift %>%
  filter(dominant_metaprogram == "MP1") %>%
  select(cell_line, MP1_shift = shift)

heterogeneity_summary <- dominant_mp_per_cl %>%
  left_join(mp1_shift_per_cl, by = "cell_line") %>%
  mutate(MP1_shift = round(MP1_shift, 1))

knitr::kable(heterogeneity_summary,
             caption = "Cell Line Dominant Metaprogram and MP1 Hypoxia Shift (pp)")

write.csv(heterogeneity_summary,
          file.path(results_dir, "Heterogeneity_Summary.csv"),
          row.names = FALSE)
cat("Summary table saved\n")
```

---

# Output Files

```{r output-summary}
cat("\nGenerated Files:\n\n")
cat("Figures (figures/):\n")
cat("  FIGURE1_InterCellLine_MPComposition.pdf\n")
cat("  FIGURE2_IntraCellLine_Hypoxia_MPShifts.pdf\n")
cat("  FIGURE3_CellLine_MPComposition_Heatmap.pdf\n")
cat("  FIGURE4_Sample_Level_Hypoxia_Shifts.pdf\n")
cat("  FIGURE5_IntraSample_Hypoxia_MPComposition.pdf\n\n")
cat("Results (results/):\n")
cat("  Signature_Genes_Summary.csv\n")
cat("  Heterogeneity_Summary.csv\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
