---
title: "DKTK-ROG validation: clinical features, TP53 and visualisations"
output: html_document
---

```{r setup, include=FALSE}
# Clinical / demographic summary of the DKTK-ROG cohort, TP53 mutation and
# p53 IHC correlations with HPV status and CT-panel clusters, plus the
# accompanying visualisation figures. Inputs: DKTK_DATA; outputs: DKTK_OUT.
source(here::here("config.R"))
knitr::opts_chunk$set(fig.width = 14, fig.height = 8)
library(oligo)
library(biomaRt)
library(dplyr)
library(pheatmap)
library(ggplot2)
library(reshape2)
library(readxl)
library(tidyr)
library(tableone)
library(gridExtra)
library(UpSetR)
library(grid)
library(rstatix)
library(data.table)
library(survival)
library(survminer)
library(purrr)
```

# Load and preprocess expression data

```{r}
load(file = file.path(DKTK_DATA, "datarma.rda"))
boxplot(exprs(data.rma))
```

```{r}
probesets <- rownames(data.rma) 
probesets.corr <- gsub("\\.1", "", probesets)
rownames(data.rma) <- probesets.corr
fData(data.rma) <- as.data.frame(probesets.corr)
```

```{r}
ensembl <- biomaRt::useMart("ensembl", dataset = "hsapiens_gene_ensembl")
annotLookup <- biomaRt::getBM(attributes = c("affy_hta_2_0",
                              "ensembl_gene_id",
                              "external_gene_name"
                            ),
               filters = "with_affy_hta_2_0",
               values = TRUE,
               mart = ensembl) 
```

```{r}
indicesLookup <- match(probesets.corr, annotLookup$affy_hta_2_0)
indicesLookup <- data.frame(probesets.corr, annotLookup[indicesLookup,])

indicesLookup$control <- ifelse(
  is.na(indicesLookup$external_gene_name),
  ifelse(grepl("(TC.*hg)", probesets.corr), 
         FALSE, 
         TRUE),
  FALSE)

data_dktk.final <- subset(data.rma, !indicesLookup$control)
```

```{r}
fData(data_dktk.final) <- left_join(fData(data_dktk.final), indicesLookup[1:4])
```

```{r}
result <- tibble(fData(data_dktk.final)$probesets.corr, 
                 fData(data_dktk.final)$external_gene_name, 
                 .name_repair = ~ c("Probeset", "External_gene_name"))
result <- bind_cols(result, as_tibble(exprs(data_dktk.final)))

format_sample_id <- function(id) {
  prefix <- gsub("(DKTKRO1[ab])(\\d+)", "\\1", id)
  num <- gsub("(DKTKRO1[ab])(\\d+)", "\\2", id)
  num_padded <- stringr::str_pad(num, width = 3, pad = "0")
  formatted_id <- paste0(prefix, num_padded)
  return(formatted_id)
}

original_colnames <- colnames(result)
new_colnames <- gsub("(.+)\\.CEL", "DKTKRO1\\1", original_colnames)
formatted_colnames <- sapply(new_colnames, format_sample_id)
colnames(result) <- formatted_colnames
result <- result %>%
  dplyr::rename(Probeset = ProbesetProbeset, External_gene_name = External_gene_nameExternal_gene_name)
```

```{r}
data_dktk <- result
data_dktk <- data_dktk %>% filter(External_gene_name != "")
```

```{r}
data_dktk <- data.table::data.table(data_dktk)

nondups <- data_dktk[!(
  duplicated(data_dktk, by = "External_gene_name") | 
    duplicated(data_dktk, fromLast = TRUE, by="External_gene_name")
),]
dups <- data_dktk[(
  duplicated(data_dktk, by = "External_gene_name") | 
                 duplicated(data_dktk, fromLast = TRUE, by="External_gene_name")
),]

log_median <- function(x) {
  n <- length(x)
  s <- sort(x)
  r <- 0:1
  ifelse(n %% 2 == 1, 
         s[(n+1)/2], 
         exp(median(log(s[n/2+r])))
  )
}

patient_cols <- which(!(colnames(dups) %in% c("Probeset", "External_gene_name")))

dups <- dups %>% 
  group_by(External_gene_name) %>%
  summarise_at(.vars = names(.)[patient_cols],
               .funs = log_median)

data_dktk <- rbind(nondups[,-1], dups)
```

# Load clinical data

```{r}
dktk_adjuvant <- read.csv(file.path(DKTK_DATA, "dktk_adjuvant.csv"), check.names = FALSE)

samples <- colnames(data_dktk)[-1]

dktk_adjuvant_microarray <- dktk_adjuvant %>%
  filter(`Participant ID` %in% samples)
```

# Prepare expression and metadata

```{r}
metadata_adjuvant <- as.data.frame(dktk_adjuvant_microarray)
rownames(metadata_adjuvant) <- metadata_adjuvant[,1]
metadata_adjuvant <- metadata_adjuvant[, -1]

data_adjuvant <- as.data.frame(data_dktk)
rownames(data_adjuvant) <- data_adjuvant[,1]
data_adjuvant <- data_adjuvant[,-1]

common_row_names <- intersect(colnames(data_adjuvant), rownames(metadata_adjuvant))
data_adjuvant <- data_adjuvant[, common_row_names]

expression_data_adjuvant <- data.frame(lapply(data_adjuvant, as.numeric), row.names = rownames(data_adjuvant))
expression_data_adjuvant_t <- t(expression_data_adjuvant)

metadata_adjuvant <- metadata_adjuvant[match(colnames(expression_data_adjuvant), rownames(metadata_adjuvant)), ]
```

# K-means clustering

```{r}
final_gene <- c("RIBC2", "TCAM1P", "SMC1B", "STAG3", "SYCP2")

scaled_patients <- scale(expression_data_adjuvant_t[, final_gene], center = TRUE, scale = TRUE)

kmeans_clustering <- function(data) {
  set.seed(42)
  kmeans_result <- kmeans(data, centers = 3, nstart = 100, iter.max = 10, algorithm = "Hartigan-Wong")
  cluster_means <- tapply(rowMeans(data), kmeans_result$cluster, mean)
  
  if (cluster_means[1] > cluster_means[2] && cluster_means[1] > cluster_means[3]) {
    high <- 1; medium <- ifelse(cluster_means[2] > cluster_means[3], 2, 3); low <- ifelse(cluster_means[2] > cluster_means[3], 3, 2)
  } else if (cluster_means[2] > cluster_means[1] && cluster_means[2] > cluster_means[3]) {
    high <- 2; medium <- ifelse(cluster_means[1] > cluster_means[3], 1, 3); low <- ifelse(cluster_means[1] > cluster_means[3], 3, 1)
  } else {
    high <- 3; medium <- ifelse(cluster_means[1] > cluster_means[2], 1, 2); low <- ifelse(cluster_means[1] > cluster_means[2], 2, 1)
  }
  
  cluster <- as.factor(kmeans_result$cluster)
  levels(cluster)[high] <- "High CT Panel Expression"
  levels(cluster)[medium] <- "Intermediate CT Panel Expression"
  levels(cluster)[low] <- "Low CT Panel Expression"
  
  return(list(cluster_labels = cluster, centroids = kmeans_result$centers))
}

clustering_results <- kmeans_clustering(scaled_patients[, final_gene])
expression_data_adjuvant_t <- as.data.frame(expression_data_adjuvant_t)
expression_data_adjuvant_t <- tibble::rownames_to_column(expression_data_adjuvant_t, "Patient_ID")
expression_data_adjuvant_t$DEG.kmeans <- clustering_results$cluster_labels
```

# Prepare data for clinical analysis

```{r}
LTC_adjuvant <- read.csv(file.path(DKTK_DATA, "adjuvant_merged.csv"))

cluster <- cbind(scaled_patients[, final_gene], expression_data_adjuvant_t[, c("DEG.kmeans", "Patient_ID")])

clusters_df <- data.frame(Participant.ID = rownames(cluster), 
                          clusters_elastic_adjuvant = cluster[, "DEG.kmeans"])

filtered_LTC_adjuvant <- LTC_adjuvant[LTC_adjuvant$Participant.ID %in% clusters_df$Participant.ID, ]

selected_columns <- filtered_LTC_adjuvant[, c("Participant.ID", colnames(LTC_adjuvant)[14:19])]

merged_df_elastic_adjuvant <- merge(clusters_df, selected_columns, by = "Participant.ID")

# Prepare metadata_adjuvant_filtered from original metadata
metadata_adjuvant$sample_id <- rownames(metadata_adjuvant)
metadata_adjuvant_filtered <- metadata_adjuvant

metadata_adjuvant_filtered$Participant.ID <- rownames(metadata_adjuvant_filtered)
merged_with_metadata <- merge(merged_df_elastic_adjuvant,
                              metadata_adjuvant_filtered,
                              by = "Participant.ID")

annets <- read.csv(file.path(DKTK_DATA, "tp53_status_dktk_adjuvant.csv"))
merged_with_metadata <- merge(merged_with_metadata,
                              annets[, c("Participant.ID", "TP53.status", "TP53.variant")],
                              by = "Participant.ID", all.x = TRUE)

merged_with_metadata <- merged_with_metadata %>%
  mutate(TP53_variant_group = case_when(
    is.na(TP53.variant) ~ NA_character_,
    TP53.variant == "wt" ~ "WT",
    grepl("fs$|\\*$", TP53.variant) ~ "Frameshift/Nonsense Mutation",
    TRUE ~ "Missense Mutation"
  ))

merged_with_metadata <- merged_with_metadata %>%
  mutate(across(c(P16, `HPV16 DNS`, `HPV16 RNA`), ~ ifelse(. == 1, "Positive", "Negative")))
colnames(merged_with_metadata)[colnames(merged_with_metadata) == "HPV16 DNS"] <- "HPV16 DNA"

merged_with_metadata <- merged_with_metadata %>%
  mutate(
    P16_Positive               = ifelse(P16 == "Positive", "Positive", "Negative"),
    HPV16DNA_Positive          = ifelse(`HPV16 DNA` == "Positive", "Positive", "Negative"),
    HPV16RNA_Positive          = ifelse(`HPV16 RNA` == "Positive", "Positive", "Negative"),
    Double_P16_HPV16DNA        = ifelse(P16 == "Positive" & `HPV16 DNA` == "Positive", "Positive", "Negative"),
    Double_P16_HPV16RNA        = ifelse(P16 == "Positive" & `HPV16 RNA` == "Positive", "Positive", "Negative"),
    Double_HPV16DNA_HPV16RNA   = ifelse(`HPV16 DNA` == "Positive" & `HPV16 RNA` == "Positive", "Positive", "Negative"),
    Triple_Positive            = ifelse(P16 == "Positive" & `HPV16 DNA` == "Positive" & `HPV16 RNA` == "Positive", "Positive", "Negative"),
    Any_Positive               = ifelse(P16 == "Positive" | `HPV16 DNA` == "Positive" | `HPV16 RNA` == "Positive", "Positive", "Negative")
  )
```

# Table One: Clinical characteristics

```{r}
# Rename columns for clarity (from original file)
colnames(merged_with_metadata)[colnames(merged_with_metadata) == "clusters_elastic_adjuvant"] <- "K-means cluster"
colnames(merged_with_metadata)[colnames(merged_with_metadata) == "HPV16 DNS"]            <- "HPV16 DNA"
colnames(merged_with_metadata)[colnames(merged_with_metadata) == "T E1 C1"]              <- "Tumor stage"
colnames(merged_with_metadata)[colnames(merged_with_metadata) == "Sex.x"]                <- "Gender"
colnames(merged_with_metadata)[colnames(merged_with_metadata) == "AGE E1 C1"]            <- "Age"
colnames(merged_with_metadata)[colnames(merged_with_metadata) == "LOCALISATION E1 C1"]   <- "Tumor localisation"
colnames(merged_with_metadata)[colnames(merged_with_metadata) == "N E1 C1"]              <- "N status"
colnames(merged_with_metadata)[colnames(merged_with_metadata) == "R E1 C1"]              <- "R status"
colnames(merged_with_metadata)[colnames(merged_with_metadata) == "GRADING E1 C1"]        <- "Grading"
colnames(merged_with_metadata)[colnames(merged_with_metadata) == "SUNCF TABACCO E3 C3"] <- "Smoking"
colnames(merged_with_metadata)[colnames(merged_with_metadata) == "SUABUSE ALCOHOL E3 C3"] <- "Alcohol"
colnames(merged_with_metadata)[colnames(merged_with_metadata) == "UICCSTATE E1 C1"]      <- "UICC status"
colnames(merged_with_metadata)[colnames(merged_with_metadata) == "ECESTATE E1 C1"]       <- "ECE status"
colnames(merged_with_metadata)[colnames(merged_with_metadata) == "1TSTDOSE E2 C2"]       <- "Chemotherapy dose (Gy)"

# Standardize Gender values
merged_with_metadata$Gender[merged_with_metadata$Gender == "f"] <- "Female"
merged_with_metadata$Gender[merged_with_metadata$Gender == "m"] <- "Male"

# Recode binary and categorical markers
merged_with_metadata$P53[merged_with_metadata$P53 == 0] <- "Negative"
merged_with_metadata$P53[merged_with_metadata$P53 == 1] <- "Positive"

merged_with_metadata$`HPV16 RNA`[merged_with_metadata$`HPV16 RNA` == 0] <- "Negative"
merged_with_metadata$`HPV16 RNA`[merged_with_metadata$`HPV16 RNA` == 1] <- "Positive"

merged_with_metadata$`HPV16 DNA`[merged_with_metadata$`HPV16 DNA` == 0] <- "Negative"
merged_with_metadata$`HPV16 DNA`[merged_with_metadata$`HPV16 DNA` == 1] <- "Positive"

# Recode Tumor stage
merged_with_metadata$`Tumor stage`[merged_with_metadata$`Tumor stage` == "0"] <- "Tx"
merged_with_metadata$`Tumor stage`[merged_with_metadata$`Tumor stage` == "1"] <- "T0"
merged_with_metadata$`Tumor stage`[merged_with_metadata$`Tumor stage` == "2"] <- "Tis"
merged_with_metadata$`Tumor stage`[merged_with_metadata$`Tumor stage` == "3"] <- "T1"
merged_with_metadata$`Tumor stage`[merged_with_metadata$`Tumor stage` == "4"] <- "T2"
merged_with_metadata$`Tumor stage`[merged_with_metadata$`Tumor stage` == "5"] <- "T3"
merged_with_metadata$`Tumor stage`[merged_with_metadata$`Tumor stage` == "6"] <- "T4"

# Recode N status
merged_with_metadata$`N status`[merged_with_metadata$`N status` == "0"] <- "Nx"
merged_with_metadata$`N status`[merged_with_metadata$`N status` == "1"] <- "N0"
merged_with_metadata$`N status`[merged_with_metadata$`N status` == "2"] <- "N1"
merged_with_metadata$`N status`[merged_with_metadata$`N status` == "3"] <- "N2"
merged_with_metadata$`N status`[merged_with_metadata$`N status` == "4"] <- "N2a"
merged_with_metadata$`N status`[merged_with_metadata$`N status` == "5"] <- "N2b"
merged_with_metadata$`N status`[merged_with_metadata$`N status` == "6"] <- "N2c"
merged_with_metadata$`N status`[merged_with_metadata$`N status` == "7"] <- "N3"

# Recode R status
merged_with_metadata$`R status`[merged_with_metadata$`R status` == "0"] <- "Rx"
merged_with_metadata$`R status`[merged_with_metadata$`R status` == "1"] <- "R0"
merged_with_metadata$`R status`[merged_with_metadata$`R status` == "2"] <- "R1"
merged_with_metadata$`R status`[merged_with_metadata$`R status` == "3"] <- "R2"

# Recode Grading
merged_with_metadata$`Grading`[merged_with_metadata$`Grading` == "0"] <- "Gx"
merged_with_metadata$`Grading`[merged_with_metadata$`Grading` == "1"] <- "G1"
merged_with_metadata$`Grading`[merged_with_metadata$`Grading` == "2"] <- "G2"
merged_with_metadata$`Grading`[merged_with_metadata$`Grading` == "3"] <- "G3"
merged_with_metadata$`Grading`[merged_with_metadata$`Grading` == "4"] <- "G4"

# Recode UICC status
merged_with_metadata$`UICC status`[merged_with_metadata$`UICC status` == "0"] <- "0"
merged_with_metadata$`UICC status`[merged_with_metadata$`UICC status` == "1"] <- "I"
merged_with_metadata$`UICC status`[merged_with_metadata$`UICC status` == "2"] <- "II"
merged_with_metadata$`UICC status`[merged_with_metadata$`UICC status` == "3"] <- "III"
merged_with_metadata$`UICC status`[merged_with_metadata$`UICC status` == "4"] <- "IV"
merged_with_metadata$`UICC status`[merged_with_metadata$`UICC status` == "5"] <- "IVA"
merged_with_metadata$`UICC status`[merged_with_metadata$`UICC status` == "6"] <- "IVB"
merged_with_metadata$`UICC status`[merged_with_metadata$`UICC status` == "7"] <- "IVC"

# Recode ECE status
merged_with_metadata$`ECE status`[merged_with_metadata$`ECE status` == "0"] <- "No"
merged_with_metadata$`ECE status`[merged_with_metadata$`ECE status` == "1"] <- "Yes"

# Recode Smoking
merged_with_metadata$`Smoking`[merged_with_metadata$`Smoking` == "0"] <- "Yes"
merged_with_metadata$`Smoking`[merged_with_metadata$`Smoking` == "1"] <- "Ex-smoker since 1 Year"
merged_with_metadata$`Smoking`[merged_with_metadata$`Smoking` == "2"] <- "No"
merged_with_metadata$`Smoking`[merged_with_metadata$`Smoking` == "3"] <- "Unknown"

# Recode Alcohol
merged_with_metadata$`Alcohol`[merged_with_metadata$`Alcohol` == "0"] <- "Yes"
merged_with_metadata$`Alcohol`[merged_with_metadata$`Alcohol` == "1"] <- "Non-Alcoholic since 1 Year"
merged_with_metadata$`Alcohol`[merged_with_metadata$`Alcohol` == "2"] <- "No"
merged_with_metadata$`Alcohol`[merged_with_metadata$`Alcohol` == "3"] <- "Unknown"

# Filter to Any_Positive cohort
final_merged_data <- merged_with_metadata %>%
  filter(`Any_Positive` == "Positive")

# Ensure Participant.ID exists
if (!"Participant.ID" %in% colnames(final_merged_data)) {
  final_merged_data$Participant.ID <- rownames(final_merged_data)
}

vars <- c("P53", "TP53.status", "TP53.variant", "HPV16 DNA", "HPV16 RNA", "P16",
          "Gender", "Tumor stage", "Age", "Chemotherapy dose (Gy)", 
          "Smoking", "Alcohol", "N status", "R status", "Grading", "UICC status", 
          "ECE status")
          
categorical_vars <- c("P53", "Gender", "Tumor stage", "Smoking", 
                      "Alcohol", "N status", "R status", "Grading", "UICC status", 
                      "ECE status", "TP53.status", "HPV16 DNA", "HPV16 RNA", "P16")

table1 <- CreateTableOne(vars = vars, 
                         strata = "K-means cluster", 
                         data = final_merged_data, 
                         factorVars = categorical_vars)

final_table <- as.data.frame(print(table1, showAllLevels = TRUE, quote = TRUE, noSpaces = TRUE))

rownames(final_table) <- gsub("^X\\.|\\.+$", "", rownames(final_table))
rownames(final_table) <- gsub("\\.+", " ", rownames(final_table))
rownames(final_table) <- gsub("^\\d+\\s+", "", rownames(final_table))
final_table$row_labels <- rownames(final_table)

print(final_table)
write.csv(final_table, file.path(DKTK_OUT, "table1_dktkadjuvant.csv"))
```

# TP53 correlation analysis

```{r}
# Define the gene-of-interest (GOI) list for TP53 correlations
# Only significant CT panel genes from DEG analysis
final_gene_extended <- c(
  "TP53", "RIBC2", "STAG3", "SMC1B", "SYCP2", "TCAM1P", "MYO3A", 
  "YBX2", "TAF7L", "ZNF541", "SYCE2", "SOX30", "ZCWPW1", "MAGEB2"
)

cor_strength <- function(r) {
  a <- abs(r)
  dplyr::case_when(
    a < 0.3 ~ "Weak (<0.3)",
    a < 0.6 ~ "Moderate (0.3ŌĆō0.6)",
    TRUE    ~ "Strong (>0.6)"
  )
}

tp53_vs_all <- function(expr_mat, method = c("spearman", "pearson")) {
  method <- match.arg(method)

  stopifnot("TP53" %in% rownames(expr_mat))
  genes <- rownames(expr_mat)

  tp53_vec <- as.numeric(expr_mat["TP53", ])

  res <- map_dfr(setdiff(genes, "TP53"), function(g) {
    x <- tp53_vec
    y <- as.numeric(expr_mat[g, ])

    ok <- is.finite(x) & is.finite(y)
    if (sum(ok) < 5) return(NULL)

    ct <- suppressWarnings(stats::cor.test(x[ok], y[ok], method = method, exact = FALSE))

    tibble(
      Gene = g,
      Correlation = unname(ct$estimate),
      P_Value = ct$p.value
    )
  })

  res %>%
    mutate(
      FDR = p.adjust(P_Value, method = "BH"),
      Direction = ifelse(Correlation >= 0, "Positive", "Negative"),
      Strength = cor_strength(Correlation),
      Method = method
    )
}

selected_dets <- c("Any", "P16", "HPV16DNA")

tp53_all <- list()

for (name in c("Any_Positive", "P16_Positive", "HPV16DNA_Positive")) {

  md_sub <- final_merged_data %>% filter((!!sym(name)) == "Positive")
  if (nrow(md_sub) < 10) next

  common_samples <- intersect(md_sub$Participant.ID, colnames(expression_data_adjuvant))
  if (length(common_samples) < 5) next

  expr_sub <- expression_data_adjuvant[, common_samples, drop = FALSE]
  # Include TP53 plus the extended CT panel genes
  genes_to_use <- union("TP53", intersect(rownames(expr_sub), final_gene_extended))
  expr_sub <- expr_sub[genes_to_use, , drop = FALSE]

  if (!"TP53" %in% rownames(expr_sub)) next

  det_name <- gsub("_Positive", "", name)
  tp53_res <- tp53_vs_all(expr_sub, method = "spearman") %>%
    mutate(Detection = det_name)

  tp53_all[[det_name]] <- tp53_res
}

tp53_tbl <- bind_rows(tp53_all)
if(nrow(tp53_tbl) == 0) {
  cat("Warning: tp53_tbl is empty. Check that samples have been properly filtered.\n")
}
```

# TP53 correlation plot

```{r}
if(nrow(tp53_tbl) > 0) {
  tp53_tbl_plot <- tp53_tbl %>%
    mutate(
      Gene = reorder(Gene, Correlation),
      signif = case_when(
        FDR < 0.001 ~ "***",
        FDR < 0.01  ~ "**",
        FDR < 0.05  ~ "*",
        TRUE        ~ ""
      )
    )

  p <- ggplot(tp53_tbl_plot, aes(x = Correlation, y = Gene, fill = Strength)) +
    geom_col(width = 0.75) +
    geom_text(
      aes(label = sprintf("%.2f", Correlation)),
      hjust = ifelse(tp53_tbl_plot$Correlation >= 0, 1.1, -0.1),
      color = "black",
      size = 3.5
    ) +
    geom_text(
      aes(label = signif),
      hjust = ifelse(tp53_tbl_plot$Correlation >= 0, -0.2, 1.2),
      size = 5,
      fontface = "bold"
    ) +
    facet_wrap(~ Detection, nrow = 1) +
    coord_cartesian(xlim = c(-1, 1), clip = "off") +
    theme_minimal() +
    theme(
      legend.position = "top",
      panel.grid.major.y = element_blank(),
      plot.margin = margin(5.5, 20, 5.5, 5.5)
    ) +
    labs(
      x = "Spearman correlation with TP53 (Žü)",
      y = NULL,
      fill = "Strength"
    )

  print(p)
  write.csv(tp53_tbl, file.path(DKTK_OUT, "TP53_correlations_all_detections.csv"))
} else {
  cat("No TP53 correlations to plot - empty dataset\n")
}
```

# Add gene expression to merged metadata

```{r}
gene_list <- c("TP53", "SMC1B", "STAG3", "SYCP2", "TCAM1P", "RIBC2")

for(g in gene_list) {
  if(g %in% rownames(expression_data_adjuvant)) {
    merged_with_metadata[[g]] <- as.numeric(expression_data_adjuvant[g, merged_with_metadata$Participant.ID])
  } else {
    merged_with_metadata[[g]] <- NA
  }
}

# Also add genes to final_merged_data
for(g in gene_list) {
  if(g %in% rownames(expression_data_adjuvant)) {
    final_merged_data[[g]] <- as.numeric(expression_data_adjuvant[g, final_merged_data$Participant.ID])
  } else {
    final_merged_data[[g]] <- NA
  }
}

print(head(merged_with_metadata[, c("Participant.ID", gene_list)]))
```

# TP53 expression by cluster

```{r}
final_merged_data$`K-means cluster` <- factor(
  final_merged_data$`K-means cluster`,
  levels = c("High CT Panel Expression", "Intermediate CT Panel Expression", "Low CT Panel Expression")
)

filtered_ct <- final_merged_data %>%
  dplyr::filter(
    !is.na(TP53),
    !is.na(`K-means cluster`),
    !is.na(P53),
    !is.na(TP53_variant_group),
    TP53_variant_group != "Unknown Mutation Status"
  )

p_tp53_all <- ggplot(filtered_ct,
                     aes(x = `K-means cluster`, y = TP53, fill = `K-means cluster`)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5) +
  geom_jitter(aes(color = TP53_variant_group), width = 0.2, size = 2, alpha = 0.8) +
  facet_wrap(~ P53) +
  labs(title = "TP53 Expression by CT Panel Group",
       x = "CT Panel Group",
       y = "TP53 Expression",
       fill = "CT Panel Group",
       color = "TP53 Variant Group") +
  scale_fill_manual(values = c("High CT Panel Expression" = "#7AB2D3", 
                               "Intermediate CT Panel Expression" = "#E7B800", 
                               "Low CT Panel Expression" = "#DE3163")) +
  theme_minimal() +
  theme(text = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 1))

print(p_tp53_all)

# Tukey HSD post-hoc test per P53 status
get_tukey_one_facet <- function(df_one_p53) {
  # Create clean helper column for stats
  df_one_p53$K_means_cluster <- as.factor(df_one_p53$`K-means cluster`)
  
  if (dplyr::n_distinct(df_one_p53$K_means_cluster) < 2) return(tibble())

  fit <- aov(TP53 ~ K_means_cluster, data = df_one_p53)
  tk <- TukeyHSD(fit)$K_means_cluster

  out <- as.data.frame(tk)
  out$comparison <- rownames(out)

  out %>%
    as_tibble() %>%
    separate(comparison, into = c("group1", "group2"), sep = "-", remove = TRUE) %>%
    transmute(group1 = group1, group2 = group2, p.adj = `p adj`)
}

tukey_results_ct <- filtered_ct %>%
  group_by(P53) %>%
  group_modify(~ get_tukey_one_facet(.x)) %>%
  ungroup() %>%
  mutate(
    p.adj.signif = case_when(
      p.adj <= 0.0001 ~ "****",
      p.adj <= 0.001  ~ "***",
      p.adj <= 0.01   ~ "**",
      p.adj <= 0.05   ~ "*",
      TRUE            ~ "ns"
    )
  )

max_tp53 <- filtered_ct %>%
  group_by(P53) %>%
  summarise(max_TP53 = max(TP53, na.rm = TRUE), .groups = "drop")

tukey_results_ct <- tukey_results_ct %>%
  left_join(max_tp53, by = "P53") %>%
  group_by(P53) %>%
  mutate(y.position = max_TP53 + seq(0.1, 0.1 * n(), length.out = n())) %>%
  ungroup()

p_tp53_all <- p_tp53_all +
  stat_pvalue_manual(
    tukey_results_ct,
    label = "p.adj.signif",
    y.position = "y.position",
    tip.length = 0.02
  )

print(p_tp53_all)

# Save plot
ggsave(file.path(DKTK_OUT, "TP53_correlation_by_detection.pdf"), plot = p_tp53_all, width = 14, height = 10, units = "in")
```

# HPV detection and TP53 interaction plot

```{r}
library(scales)

# 1) Define short labels for detection methods (simplified to 5 main categories)
short_labs <- c(
  Any_Positive                   = "Any+",
  Triple_Positive                = "Triple+",
  P16_Positive                   = "P16+",
  HPV16DNA_Positive              = "DNA+",
  HPV16RNA_Positive              = "RNA+"
)

# Pivot to long and compute variant/HPV interaction
long_data3 <- final_merged_data %>%
  dplyr::select(Participant.ID,
         TP53_variant_group,
         P53,
         `K-means cluster`,
         all_of(names(short_labs))) %>%
  pivot_longer(
    cols      = names(short_labs),
    names_to  = "Detection_Method",
    values_to = "HPV_Status"
  ) %>%
  filter(!is.na(HPV_Status)) %>%
  mutate(
    Detection_Short = short_labs[Detection_Method],
    P53             = factor(P53,
                             levels = c("Negative","Positive"),
                             labels = c("P53ŌĆō","P53+")),
    `K-means cluster` = factor(`K-means cluster`,
      levels = c("High CT Panel Expression",
                 "Intermediate CT Panel Expression",
                 "Low CT Panel Expression")
    ),
    Variant = case_when(
      stringr::str_detect(TP53_variant_group, "WT")            ~ "WT",
      stringr::str_detect(TP53_variant_group, "Missense")      ~ "Missense",
      stringr::str_detect(TP53_variant_group, "Frameshift|Nonsense") ~ "Frameshift/Nonsense",
      TRUE                                            ~ "Unknown"
    ),
    HPV_Status = factor(HPV_Status, levels = c("Positive","Negative")),
    VarHPV = factor(
      paste(Variant, HPV_Status, sep = "."),
      levels = c(
        "WT.Positive","WT.Negative",
        "Missense.Positive","Missense.Negative",
        "Frameshift/Nonsense.Positive","Frameshift/Nonsense.Negative",
        "Unknown.Positive","Unknown.Negative"
      )
    )
  )

# Build 8-color palette
base_cols <- c(
  WT                     = "#2b8cbe",
  Missense               = "#8856a7",
  `Frameshift/Nonsense`  = "#41b6c4",
  Unknown                = "#000000"
)
pal <- setNames(
  sapply(levels(long_data3$VarHPV), function(lvl) {
    parts <- strsplit(lvl, "\\.")[[1]]
    col   <- base_cols[parts[1]]
    if (parts[2] == "Negative") col <- scales::alpha(col, 0.4)
    col
  }),
  levels(long_data3$VarHPV)
)

# Horizontal bar plot
p_final_axis <- ggplot(long_data3, aes(x = Detection_Short, fill = VarHPV)) +
  geom_bar(position = "stack", color = "black", size = 0.2) +
  facet_grid(`K-means cluster` ~ P53, switch = "x") +
  coord_flip() +
  scale_fill_manual(values = pal,
                    name   = "TP53 var ┬Ę HPV",
                    drop   = FALSE) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.05)),
    breaks = pretty_breaks(n = 5)
  ) +
  labs(
    x     = NULL,
    y     = "Number of patients",
    title = "HPV detection methods ├Ś TP53 variants,\nby CT-panel & P53 stain"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    strip.placement    = "outside",
    strip.text.x       = element_text(size = 12, face = "bold"),
    strip.text.y       = element_text(size = 12, face = "bold"),
    axis.text.y        = element_text(size = 10, colour = "black"),
    axis.text.x        = element_text(size = 12, colour = "black"),
    panel.grid.major.y = element_blank(),
    legend.position    = "bottom",
    plot.margin        = margin(t = 10, r = 50, b = 10, l = 10)  # Add right margin to prevent cutoff
  )

print(p_final_axis)

# Save HPV interaction plot
ggsave(file.path(DKTK_OUT, "HPV_detection_TP53_interaction.pdf"), plot = p_final_axis, width = 16, height = 10, units = "in")
```

# HPV detection counts summary

```{r}
# Create upset_data from final_merged_data
upset_data <- final_merged_data %>%
  select(Any_Positive, Triple_Positive, P16_Positive, HPV16DNA_Positive, HPV16RNA_Positive, P53)

# HPV detection counts summary
hpv_detection_summary <- data.frame(
  Detection_Method = c("Any Positive", "Triple Positive", "P16+", "DNA+", "RNA+"),
  Count = c(
    sum(upset_data$Any_Positive == "Positive", na.rm = TRUE),
    sum(upset_data$Triple_Positive == "Positive", na.rm = TRUE),
    sum(upset_data$P16_Positive == "Positive", na.rm = TRUE),
    sum(upset_data$HPV16DNA_Positive == "Positive", na.rm = TRUE),
    sum(upset_data$HPV16RNA_Positive == "Positive", na.rm = TRUE)
  )
)

# P53 positive/negative by HPV detection method
p53_hpv_breakdown <- data.frame(
  Detection_Method = rep(c("Any Positive", "Triple Positive", "P16+", "DNA+", "RNA+"), 2),
  P53_Status = rep(c("P53 Positive", "P53 Negative"), each = 5),
  Count = c(
    # P53 Positive
    sum(upset_data$Any_Positive == "Positive" & upset_data$P53 == "Positive", na.rm = TRUE),
    sum(upset_data$Triple_Positive == "Positive" & upset_data$P53 == "Positive", na.rm = TRUE),
    sum(upset_data$P16_Positive == "Positive" & upset_data$P53 == "Positive", na.rm = TRUE),
    sum(upset_data$HPV16DNA_Positive == "Positive" & upset_data$P53 == "Positive", na.rm = TRUE),
    sum(upset_data$HPV16RNA_Positive == "Positive" & upset_data$P53 == "Positive", na.rm = TRUE),
    # P53 Negative
    sum(upset_data$Any_Positive == "Positive" & upset_data$P53 == "Negative", na.rm = TRUE),
    sum(upset_data$Triple_Positive == "Positive" & upset_data$P53 == "Negative", na.rm = TRUE),
    sum(upset_data$P16_Positive == "Positive" & upset_data$P53 == "Negative", na.rm = TRUE),
    sum(upset_data$HPV16DNA_Positive == "Positive" & upset_data$P53 == "Negative", na.rm = TRUE),
    sum(upset_data$HPV16RNA_Positive == "Positive" & upset_data$P53 == "Negative", na.rm = TRUE)
  )
)

cat("\n=== HPV DETECTION METHOD COUNTS ===\n")
print(hpv_detection_summary)

cat("\n=== HPV DETECTION BY P53 STATUS ===\n")
print(p53_hpv_breakdown)

# Save to CSV
write.csv(hpv_detection_summary, file.path(DKTK_OUT, "summary_hpv_detection_counts.csv"), row.names = FALSE)
write.csv(p53_hpv_breakdown, file.path(DKTK_OUT, "summary_hpv_p53_breakdown.csv"), row.names = FALSE)
```



