---
title: "Align Tissue Positions with Hypoxia Data"
subtitle: "All Available Samples - Dynamic Discovery"
output: 
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
```

## Overview

This document automatically discovers and aligns all available samples from the parallel_results directory with hypoxia registration data. It scans for tissue position files and corresponding hypoxia files.

---

## Load Libraries

```{r load-libraries}
library(tidyverse)
```

---

## Define Alignment Function

```{r define-function}
# Function to align and merge spatial data
align_spots_with_hypoxia <- function(tissue_pos_file, hypoxia_file, sample_name) {
  
  # Read files
  tissue_pos <- read_csv(tissue_pos_file, show_col_types = FALSE)
  hypoxia <- read_csv(hypoxia_file, show_col_types = FALSE)
  
  # Filter tissue positions for in_tissue == 1 (tissue-only)
  tissue_in <- tissue_pos |>
    filter(in_tissue == 1) |>
    mutate(sample = sample_name)
  
  # Rename updated column to match downstream expectations
  if ("hypoxia_label_25" %in% colnames(hypoxia)) {
    hypoxia <- hypoxia |> rename(hypoxia_label = hypoxia_label_25)
  }

  # Add sample to hypoxia data to prevent cross-sample contamination
  hypoxia <- hypoxia |>
    mutate(sample = sample_name)
  
  # Merge on pixel coordinates (unique spatial identifier within sample)
  # Join on: pixel_row, pixel_col, and sample name
  merged <- tissue_in |>
    left_join(
      hypoxia,
      by = c("pxl_row_in_fullres", "pxl_col_in_fullres", "sample"),
      relationship = "one-to-one"
    )
  
  return(merged)
}
```

---

## Discover Available Samples

```{r discover-samples}
# Define base paths
source(here::here("config.R"))  # defines DATA_DIR, RESULTS_DIR, ANALYSIS_SUBTYPES_DIR
tissue_pos_dir <- file.path(DATA_DIR, "SPT/samples_3.3.1")
hypoxia_dir <- file.path(DATA_DIR, "SPT/Hypoxia_registration_SPT_results/hypoxia_spots_updated")

# Get all sample directories (one level deep from samples_3.3.1)
all_sample_dirs <- list.dirs(tissue_pos_dir, full.names = FALSE, recursive = FALSE)
all_sample_dirs <- all_sample_dirs[all_sample_dirs != ""]  # Remove empty

cat("Found", length(all_sample_dirs), "sample directories\n\n")

# Discover all hypoxia files
hypoxia_files <- list.files(hypoxia_dir, pattern = "^tp_.*\\.csv$", 
                            full.names = TRUE)
cat("Found", length(hypoxia_files), "hypoxia files\n")
cat("Hypoxia samples:", paste(sort(gsub("tp_|.csv", "", basename(hypoxia_files))), collapse = ", "), "\n\n")

# Extract sample names and build mapping - ONLY include if hypoxia data exists
samples <- list()

for (sample_name in all_sample_dirs) {
  # Build tissue_pos path for this sample - ONLY use outs/spatial path
  tissue_pos_file <- file.path(tissue_pos_dir, sample_name, "outs", "spatial", "tissue_positions.csv")
  
  if (file.exists(tissue_pos_file)) {
    # Extract identifier from sample name
    # E.g., N150a014 from N150a014rep_FaDu_SPT_A1 or N165a002 from N165a002_SAT_SPT_C1
    sample_id <- sub("^(N[0-9]+[a-z][0-9]+).*", "\\1", sample_name)
    
    # Look for EXACT matching hypoxia file
    hyp_file <- file.path(hypoxia_dir, paste0("tp_", sample_id, ".csv"))
    
    if (file.exists(hyp_file)) {
      # Extract cell line using SAME pattern as QC script
      extracted_cell_line <- gsub(".*_(FaDu|UT5|Cal33|UT8|SAS|UT45|SAT).*", "\\1", sample_name)
      
      # If extraction failed, mark as "OTHER"
      if (extracted_cell_line == sample_name) {
        cell_line <- "OTHER"
      } else {
        cell_line <- extracted_cell_line
      }
      
      samples[[length(samples) + 1]] <- list(
        name = sample_name,
        tissue_pos = tissue_pos_file,
        hypoxia = hyp_file,
        cell_line = cell_line
      )
      
      cat(sprintf("%s (%s) - has hypoxia data\n", sample_name, cell_line))
    } else {
      cat(sprintf("%s - NO hypoxia data\n", sample_name))
    }
  } else {
    cat(sprintf("%s - NO tissue_positions.csv\n", sample_name))
  }
}

cat("\nTotal samples with BOTH tissue and hypoxia data:", length(samples), "\n")
```

---

## Run Alignment

```{r run-alignment}
# Create sample info dataframe first (to avoid redundant reads)
sample_info <- map_df(samples, ~ data.frame(
  sample_name = .$name,
  cell_line = .$cell_line,
  stringsAsFactors = FALSE
))

# Run alignment and add sample-level metadata
results <- map_df(samples, ~ align_spots_with_hypoxia(
  tissue_pos_file = .$tissue_pos,
  hypoxia_file = .$hypoxia,
  sample_name = .$name
)) |>
  left_join(sample_info, by = c("sample" = "sample_name"))

cat("Alignment complete\n")
cat("Total spots:", nrow(results), "\n")
cat("Spots with hypoxia data:", sum(!is.na(results$hypoxia_label)), "\n")
cat("Spots missing hypoxia data:", sum(is.na(results$hypoxia_label)), "\n")
```

---

## Quick Check - First 10 Rows

```{r check-alignment}
results |>
  select(sample, cell_line, barcode, pxl_row_in_fullres, pxl_col_in_fullres, hypoxia_label, hypoxia_percentage) |>
  head(10)
```

---

## Summary Statistics

```{r summary-stats}
summary_stats <- results |>
  group_by(cell_line, sample, hypoxia_label) |>
  summarise(
    n_spots = n(),
    mean_hypoxia_pct = mean(hypoxia_percentage, na.rm = TRUE),
    sd_hypoxia_pct = sd(hypoxia_percentage, na.rm = TRUE),
    .groups = "drop"
  )

summary_stats
```

---

## Summary by Cell Line

```{r summary-by-cellline}
results |>
  group_by(cell_line, hypoxia_label) |>
  summarise(
    n_spots = n(),
    mean_hypoxia_pct = mean(hypoxia_percentage, na.rm = TRUE),
    .groups = "drop"
)
```

## Validate Alignment Quality

```{r validate-join}
# Check for duplicate pixel coordinate-sample combinations (shouldn't happen)
duplicates <- results |>
  group_by(pxl_row_in_fullres, pxl_col_in_fullres, sample) |>
  summarise(n = n(), .groups = "drop") |>
  filter(n > 1)

if (nrow(duplicates) > 0) {
  cat(" WARNING: Found", nrow(duplicates), "duplicate pixel-sample combinations\n")
  print(duplicates)
} else {
  cat("Join validation: Pixel coordinates are unique per sample\n")
}

# Summary of data quality
cat("\n=== ALIGNMENT QUALITY SUMMARY ===\n")
cat("Total spots after alignment:", nrow(results), "\n")
cat("Spots with hypoxia data:", sum(!is.na(results$hypoxia_label)), 
    " (", round(100 * sum(!is.na(results$hypoxia_label)) / nrow(results), 1), "%)\n")
cat("Spots missing hypoxia data:", sum(is.na(results$hypoxia_label)), "\n")

cat("\n=== HYPOXIA CLASSIFICATION ===\n")
cat("Hypoxic spots (label = 1):", sum(results$hypoxia_label == 1, na.rm = TRUE), "\n")
cat("Normoxic spots (label = 0):", sum(results$hypoxia_label == 0, na.rm = TRUE), "\n")

cat("\nBreakdown by sample:\n")
results |>
  group_by(sample) |>
  summarise(
    n_spots = n(),
    n_hypoxic = sum(hypoxia_label == 1, na.rm = TRUE),
    n_normoxic = sum(hypoxia_label == 0, na.rm = TRUE),
    pct_hypoxic = round(100 * sum(hypoxia_label == 1, na.rm = TRUE) / sum(!is.na(hypoxia_label)), 1),
    .groups = "drop"
  ) |>
  print()
```

---

## Save Output

```{r save-output}
# Only save spots that have complete hypoxia data
results_clean <- results |>
  filter(!is.na(hypoxia_label))

output_file <- "data/all_samples_aligned_hypoxia.csv"
write_csv(results_clean, file = output_file)

cat("\nSaved clean aligned data to:", output_file, "\n")
cat("Total spots (with complete hypoxia data):", nrow(results_clean), "\n")
cat("Cell lines:", paste(unique(results_clean$cell_line), collapse = ", "), "\n")
cat("Samples:", n_distinct(results_clean$sample), "\n")
```

---

## Data Columns

The output CSV contains:

| Column | Description |
|--------|-------------|
| `barcode` | 10X Visium spot barcode identifier |
| `in_tissue` | Flag (0/1) for spots in tissue region |
| `array_row`, `array_col` | Array coordinates |
| `pxl_row_in_fullres`, `pxl_col_in_fullres` | Pixel coordinates (join key for hypoxia alignment) |
| `sample` | Sample name |
| `cell_line` | Cell line (FaDu, UT5, Cal33, UT8, SAS, SAT, etc.) |
| `hypoxia_label` | Binary classification: 1 = hypoxic, 0 = normoxic |
| `hypoxia_percentage` | % hypoxic pixels in spot (0-100) |
| `hypoxic_pixels` | Count of hypoxic pixels |
| `total_pixels` | Total pixels in spot (typically 1681) |

**Note:** Alignment is performed using pixel coordinates (pxl_row_in_fullres, pxl_col_in_fullres) as the spatial identifier within each sample. This ensures no cross-sample contamination.

