---
title: "Tumor Metaprogram Integration - Intra-Sample Heterogeneity Analysis"
author: "Metaprogram pipeline adapted from Greenwald et al. (Tirosh lab)"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 8)
```

# Overview

This notebook integrates NMF programs from all hypoxia samples following the **Tirosh/Gronewald methodology**:

1. Load all NMF programs from all samples
2. Filter programs to Leiden cluster marker genes (Tirosh approach)
3. Filter for robust programs (intra-sample overlap ≥ 20 genes)
4. Calculate Jaccard similarities
5. Greedy clustering (threshold = 0.1)
6. Define consensus metaprograms

---

# Setup

```{r libraries}
library(dplyr)
library(ggplot2)
library(reshape2)
library(pheatmap)
library(RColorBrewer)
```

```{r directories}
# Hardcoded absolute paths — dirname(getwd()) is unreliable during render
source(here::here("config.R"))  # defines DATA_DIR, RESULTS_DIR, ANALYSIS_SUBTYPES_DIR
results_dir <- file.path(ANALYSIS_SUBTYPES_DIR, "data/parallel_results")

qc_dir <- file.path(ANALYSIS_SUBTYPES_DIR, "0_QC_AND_PREPROCESSING/data")

out_dir <- file.path(results_dir, "integrated")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

cat("Results directory:", results_dir, "\n")
cat("QC directory:     ", qc_dir, "\n")
cat("Output directory: ", out_dir, "\n")
```

---

# Step 1: Load All NMF Programs

Programs are already ranked by NMF loading strength from the per-sample analysis.

```{r load-programs}
cat("=== Loading NMF programs (pre-ranked by NMF loading) ===\n\n")

sample_dirs <- list.dirs(results_dir, recursive = FALSE, full.names = TRUE)
sample_dirs <- sample_dirs[!grepl("integrated", sample_dirs)]

cat("Found", length(sample_dirs), "sample directories\n\n")

all_programs <- list()

for (sdir in sample_dirs) {
  sample_name <- basename(sdir)
  nmf_file <- file.path(sdir, "nmf", "all_programs.rds")
  
  if (file.exists(nmf_file)) {
    cat("Loading:", sample_name, "\n")
    programs <- readRDS(nmf_file)
    for (prog_name in names(programs)) {
      all_programs[[prog_name]] <- programs[[prog_name]]
    }
  } else {
    cat("Missing:", sample_name, "\n")
  }
}

cat("\nTotal programs loaded:", length(all_programs), "\n")
cat("   (Each program's genes are already ranked by NMF loading)\n")
```

```{r save-all-programs}
saveRDS(all_programs, file.path(out_dir, "all_programs_combined.rds"))
```

---

# Step 1.5: Extract Leiden Cluster Markers (Tirosh Approach)

```{r extract-markers}
cat("=== Extracting cluster markers from Leiden results ===\n\n")

all_markers_combined <- list()

for (sdir in sample_dirs) {
  sample_name <- basename(sdir)
  deg_file <- file.path(sdir, "leiden", "deg_results.csv")

  if (!file.exists(deg_file)) {
    cat("No DEG results for:", sample_name, "\n")
    next
  }

  cat("Processing:", sample_name, "\n")
  markers <- read.csv(deg_file, stringsAsFactors = FALSE)

  # Get top 50 markers per cluster by avg_log2FC
  top_markers <- markers %>%
    group_by(cluster) %>%
    top_n(n = 50, wt = avg_log2FC) %>%
    pull(gene) %>%
    unique()

  all_markers_combined[[sample_name]] <- top_markers
  cat("  Found", length(top_markers), "unique marker genes\n")
}

cat("\nExtracted markers from", length(all_markers_combined), "samples\n")
saveRDS(all_markers_combined, file.path(out_dir, "leiden_marker_genes.rds"))
```

---

# Step 1.6: Filter NMF Programs to Marker Genes Only (Tirosh)

```{r filter-to-markers}
cat("=== Filtering NMF programs to Leiden marker genes ===\n\n")

all_markers_combined <- readRDS(file.path(out_dir, "leiden_marker_genes.rds"))

filtered_programs <- list()

for (prog_name in names(all_programs)) {
  sample_name <- sub("\\..*", "", prog_name)

  if (!sample_name %in% names(all_markers_combined)) {
    cat("No markers for", sample_name, "- keeping original program\n")
    filtered_programs[[prog_name]] <- all_programs[[prog_name]]
    next
  }

  sample_markers <- all_markers_combined[[sample_name]]
  prog_genes     <- all_programs[[prog_name]]
  marker_genes   <- intersect(prog_genes, sample_markers)

  if (length(marker_genes) >= 10) {
    filtered_programs[[prog_name]] <- marker_genes
  }
}

cat("\nPrograms before filtering:", length(all_programs), "\n")
cat("Programs after filtering: ", length(filtered_programs), "\n")
cat("Average genes per program (before):",
    round(mean(sapply(all_programs, length)), 1), "\n")
cat("Average genes per program (after): ",
    round(mean(sapply(filtered_programs, length)), 1), "\n\n")

filtered_per_sample <- table(sub("\\..*", "", names(filtered_programs)))
print(filtered_per_sample)
cat("Total programs after filtering:", length(filtered_programs), "\n")

all_programs <- filtered_programs
saveRDS(all_programs, file.path(out_dir, "all_programs_marker_filtered.rds"))
```

---

# Step 2: Filter for Robust Programs

Following **Tirosh criteria**: Programs must overlap with at least one other program from the **same sample** by ≥ 20 genes.

```{r robust-filtering}
cat("=== Filtering for robust programs ===\n\n")

intra_min <- 20

# Group by sample
programs_by_sample <- list()
for (prog_name in names(all_programs)) {
  sname <- sub("\\..*", "", prog_name)
  if (!sname %in% names(programs_by_sample)) programs_by_sample[[sname]] <- list()
  programs_by_sample[[sname]][[prog_name]] <- all_programs[[prog_name]]
}

cat("Samples with programs:", length(programs_by_sample), "\n\n")

# Diagnostic: program sizes
cat("Program sizes after marker filtering:\n")
print(summary(sapply(all_programs, length)))
cat("\n")

robust_programs <- list()

for (sname in names(programs_by_sample)) {
  sample_progs <- programs_by_sample[[sname]]
  n_progs      <- length(sample_progs)
  cat("Sample:", sname, "- Programs:", n_progs, "\n")

  sample_robust <- 0

  for (i in seq_along(sample_progs)) {
    prog_name  <- names(sample_progs)[i]
    prog_genes <- sample_progs[[i]]

    max_overlap <- 0
    for (j in seq_along(sample_progs)) {
      if (i != j) {
        overlap <- length(intersect(prog_genes, sample_progs[[j]]))
        if (overlap > max_overlap) max_overlap <- overlap
      }
    }

    if (max_overlap >= intra_min) {
      robust_programs[[prog_name]] <- prog_genes
      sample_robust <- sample_robust + 1
    }
  }

  cat("  → Robust:", sample_robust, "/", n_progs,
      "(", round(100 * sample_robust / n_progs, 1), "%)\n")
}

cat("\nTotal robust programs:", length(robust_programs), "/", length(all_programs), "\n")
cat("   Overall filter rate:", round(100 * length(robust_programs) / length(all_programs), 1), "%\n")

# Fallback: if threshold too strict, use all filtered programs
if (length(robust_programs) == 0) {
  cat("\n No programs passed robustness filter — using all marker-filtered programs as fallback\n")
  robust_programs <- all_programs
}
```

```{r save-robust}
saveRDS(robust_programs, file.path(out_dir, "robust_programs.rds"))
```

---

# Step 3: Calculate Jaccard Similarity

```{r jaccard-similarity}
cat("=== Calculating Jaccard similarity ===\n\n")

n_progs    <- length(robust_programs)
jaccard_mat <- matrix(0, nrow = n_progs, ncol = n_progs)
rownames(jaccard_mat) <- names(robust_programs)
colnames(jaccard_mat) <- names(robust_programs)

for (i in seq_len(n_progs)) {
  for (j in i:n_progs) {
    intersection <- length(intersect(robust_programs[[i]], robust_programs[[j]]))
    union_len    <- length(union(robust_programs[[i]], robust_programs[[j]]))
    jac          <- if (union_len == 0) 0 else intersection / union_len

    jaccard_mat[i, j] <- jac
    jaccard_mat[j, i] <- jac
  }

  if (i %% 50 == 0) cat("  Progress:", i, "/", n_progs, "\n")
}

cat("\nJaccard statistics:\n")
upper_tri <- jaccard_mat[upper.tri(jaccard_mat)]
cat("   Mean:         ", round(mean(upper_tri), 3), "\n")
cat("   Median:       ", round(median(upper_tri), 3), "\n")
if (any(upper_tri < 1)) {
  cat("   Max (non-self):", round(max(upper_tri[upper_tri < 1]), 3), "\n")
}
```

```{r save-jaccard}
saveRDS(jaccard_mat, file.path(out_dir, "jaccard_matrix.rds"))
```

---

# Step 4: Greedy Clustering

Programs with Jaccard ≥ 0.1 are grouped into the same metaprogram.

```{r greedy-clustering}
cat("=== Greedy clustering ===\n\n")

threshold       <- 0.1
clusters        <- rep(0, n_progs)
current_cluster <- 1

for (i in seq_len(n_progs)) {
  if (clusters[i] == 0) {
    clusters[i]    <- current_cluster
    similar_progs  <- which(jaccard_mat[i, ] >= threshold & clusters == 0)
    if (length(similar_progs) > 0) clusters[similar_progs] <- current_cluster
    current_cluster <- current_cluster + 1
  }
}

n_metaprograms <- max(clusters)
cat("Total metaprograms:", n_metaprograms, "\n\n")

cluster_sizes <- table(clusters)
cat("Cluster size distribution:\n")
print(summary(as.numeric(cluster_sizes)))
```

---

# Step 5: Define Metaprogram Consensus Genes (Frequency × Rank-Weighted)

```{r consensus-genes-rank-weighted}
cat("=== Defining consensus genes with frequency × rank weighting ===\n\n")

metaprograms <- list()

for (clust_id in unique(clusters)) {
  prog_indices       <- which(clusters == clust_id)
  n_progs_in_cluster <- length(prog_indices)

  all_genes_in_cluster <- unlist(robust_programs[prog_indices])
  gene_freq            <- table(all_genes_in_cluster)

  min_freq       <- max(3, ceiling(n_progs_in_cluster * 0.25))
  consensus_genes <- names(gene_freq)[gene_freq >= min_freq]

  gene_ranks       <- setNames(rep(NA, length(consensus_genes)), consensus_genes)
  gene_rank_counts <- setNames(rep(0,  length(consensus_genes)), consensus_genes)

  for (prog_idx in prog_indices) {
    prog_genes <- robust_programs[[prog_idx]]
    for (rank_pos in seq_along(prog_genes)) {
      gene <- prog_genes[rank_pos]
      if (gene %in% consensus_genes) {
        gene_ranks[gene]       <- if (is.na(gene_ranks[gene])) rank_pos else gene_ranks[gene] + rank_pos
        gene_rank_counts[gene] <- gene_rank_counts[gene] + 1
      }
    }
  }

  gene_ranks             <- gene_ranks / gene_rank_counts
  consensus_genes_sorted <- names(sort(gene_ranks[consensus_genes]))
  top_genes              <- head(consensus_genes_sorted, 50)
  mp_name                <- paste0("MP", clust_id)
  metaprograms[[mp_name]] <- top_genes

  cat(sprintf("%-6s: %3d programs → %2d genes (min_freq=%d, rank-weighted)\n",
              mp_name, n_progs_in_cluster, length(top_genes), min_freq))
}

cat("\nMetaprograms created:", length(metaprograms), "\n")
```

```{r save-metaprograms}
saveRDS(metaprograms, file.path(out_dir, "metaprogram_genes.rds"))
```

---

# Step 5.1: Extract Signature Genes (Top 5 per Metaprogram)

```{r signature-genes}
cat("=== Extracting signature genes (top 5 per metaprogram) ===\n\n")

signature_genes <- list()
for (mp_name in names(metaprograms)) {
  top_5 <- head(metaprograms[[mp_name]], 5)
  signature_genes[[mp_name]] <- top_5
  cat(sprintf("%-6s signature genes: %s\n", mp_name, paste(top_5, collapse = ", ")))
}

cat("\nSignature genes extracted:", length(signature_genes), "\n")
saveRDS(signature_genes, file.path(out_dir, "metaprogram_signature_genes.rds"))

sig_df <- data.frame(
  Metaprogram = rep(names(signature_genes), sapply(signature_genes, length)),
  Rank        = unlist(lapply(signature_genes, function(x) seq_along(x))),
  Gene        = unlist(signature_genes)
)
write.csv(sig_df, file.path(out_dir, "metaprogram_signature_genes.csv"), row.names = FALSE)
cat("   Saved: metaprogram_signature_genes.csv\n")
```

---

# Step 5.5: Filter Empty/Small Metaprograms

```{r filter-empty}
cat("=== Filtering out empty/small metaprograms ===\n\n")

min_genes    <- 10
min_programs <- 5

valid_metaprograms <- list()
valid_clusters     <- c()

for (mp_name in names(metaprograms)) {
  n_genes   <- length(metaprograms[[mp_name]])
  clust_id  <- as.numeric(sub("MP", "", mp_name))
  n_programs <- sum(clusters == clust_id)

  if (n_genes >= min_genes && n_programs >= min_programs) {
    valid_metaprograms[[mp_name]] <- metaprograms[[mp_name]]
    valid_clusters <- c(valid_clusters, clust_id)
  } else {
    cat("Removing", mp_name, "- Programs:", n_programs, "| Genes:", n_genes, "\n")
  }
}

cat("\nValid metaprograms:", length(valid_metaprograms), "/", length(metaprograms), "\n\n")

metaprograms_final <- list()
for (i in seq_along(valid_metaprograms)) {
  metaprograms_final[[paste0("MP", i)]] <- valid_metaprograms[[i]]
}
metaprograms <- metaprograms_final

saveRDS(metaprograms, file.path(out_dir, "metaprograms_final.rds"))
cat("Saved metaprograms_final.rds:", length(metaprograms), "MPs (", paste(names(metaprograms), collapse=", "), ")\n")
```

---

# Step 6: Summary Table

```{r summary-table}
summary_list <- list()

for (mp_name in names(metaprograms)) {
  original_mp_name <- names(valid_metaprograms)[which(names(metaprograms) == mp_name)]
  clust_id <- as.numeric(sub("MP", "", original_mp_name))

  summary_list[[mp_name]] <- data.frame(
    metaprogram = mp_name,
    n_programs  = sum(clusters == clust_id),
    n_genes     = length(metaprograms[[mp_name]])
  )
}

summary_df <- do.call(rbind, summary_list)
rownames(summary_df) <- NULL
summary_df <- summary_df %>% arrange(desc(n_programs))
knitr::kable(summary_df, caption = "Metaprogram Summary")
```

```{r save-summary}
write.csv(summary_df, file.path(out_dir, "metaprogram_summary.csv"), row.names = FALSE)
```

---

# Visualization: Jaccard Heatmap

```{r jaccard-heatmap, fig.width=12, fig.height=10}
cluster_order  <- order(clusters)
jaccard_ordered <- jaccard_mat[cluster_order, cluster_order]

jac_melt              <- melt(jaccard_ordered)
jac_melt$value_capped <- pmin(jac_melt$value, 0.35)

ggplot(jac_melt, aes(x = Var1, y = Var2, fill = value_capped)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "white", mid = "yellow", high = "red",
    midpoint = 0.175, limits = c(0, 0.35),
    name = "Jaccard\nSimilarity"
  ) +
  theme_minimal() +
  theme(
    axis.text.x  = element_blank(),
    axis.text.y  = element_blank(),
    axis.title   = element_blank(),
    axis.ticks   = element_blank(),
    panel.grid   = element_blank()
  ) +
  ggtitle("Hypoxia Metaprograms - Program Similarity (Ordered by Clusters)")
```

```{r save-heatmap}
ggsave(filename = file.path(out_dir, "jaccard_heatmap.pdf"), width = 12, height = 10)
```

---

# Top Genes per Metaprogram

```{r top-genes}
for (mp_name in names(metaprograms)) {
  cat("\n###", mp_name, "\n\n")
  cat(paste(head(metaprograms[[mp_name]], 20), collapse = ", "), "\n")
}
```

---

# Generate Sample × Metaprogram Counts

Score metaprograms on each sample and count spot assignments.

```{r generate-sample-mp-counts}
cat("=== Generating sample × metaprogram counts ===\n\n")

library(Seurat)

sample_mp_counts <- list()
sample_dirs_mp   <- list.dirs(results_dir, recursive = FALSE, full.names = TRUE)
sample_dirs_mp   <- sample_dirs_mp[!grepl("integrated|ipynb", basename(sample_dirs_mp))]

for (sdir in sample_dirs_mp) {
  sample_name <- basename(sdir)
  seu_file    <- file.path(qc_dir, paste0(sample_name, "_qc_with_metadata.rds"))

  if (!file.exists(seu_file)) {
    cat("Skipping", sample_name, "- no QC Seurat object\n")
    next
  }

  cat("Processing:", sample_name, "\n")
  seu <- readRDS(seu_file)
  seu <- NormalizeData(seu, assay = "Spatial", normalization.method = "LogNormalize",
                       verbose = FALSE)

  mp_scores           <- matrix(NA, nrow = ncol(seu), ncol = length(metaprograms))
  colnames(mp_scores) <- names(metaprograms)
  rownames(mp_scores) <- colnames(seu)

  for (mp_idx in seq_along(metaprograms)) {
    mp_name          <- names(metaprograms)[mp_idx]
    mp_genes_present <- metaprograms[[mp_name]][metaprograms[[mp_name]] %in% rownames(seu)]

    if (length(mp_genes_present) >= 5) {
      set.seed(42 + mp_idx)
      seu <- AddModuleScore(seu, features = list(mp_genes_present),
                            name = paste0("MP", mp_idx), ctrl = 50)
      score_col <- paste0("MP", mp_idx, "1")
      mp_scores[, mp_idx] <- seu@meta.data[[score_col]]
    }
  }

  dominant_mp <- apply(mp_scores, 1, function(x) {
    if (all(is.na(x))) return(NA)
    names(metaprograms)[which.max(x)]
  })

  mp_table <- data.frame(
    metaprogram = names(table(dominant_mp)),
    count       = as.numeric(table(dominant_mp))
  )

  sample_mp_counts[[sample_name]] <- mp_table
  cat("  → Assigned", sum(!is.na(dominant_mp)), "spots to",
      length(unique(dominant_mp[!is.na(dominant_mp)])), "metaprograms\n")
}

cat("\nProcessed", length(sample_mp_counts), "samples\n")
saveRDS(sample_mp_counts, file.path(out_dir, "sample_mp_counts.rds"))
cat("   Saved to: sample_mp_counts.rds\n")
```

---

# Heatmap: Metaprogram Presence Across Samples

```{r mp-heatmap, fig.width=12, fig.height=8}
heatmap_data <- matrix(0, nrow = length(sample_mp_counts), ncol = length(metaprograms))
rownames(heatmap_data) <- names(sample_mp_counts)
colnames(heatmap_data) <- names(metaprograms)

for (sample in names(sample_mp_counts)) {
  for (mp in names(metaprograms)) {
    count <- sample_mp_counts[[sample]]$count[sample_mp_counts[[sample]]$metaprogram == mp]
    if (length(count) > 0) heatmap_data[sample, mp] <- count
  }
}

pheatmap(
  heatmap_data,
  color          = colorRampPalette(c("white", "yellow", "orange", "red"))(50),
  cluster_rows   = TRUE,
  cluster_cols   = TRUE,
  show_rownames  = TRUE,
  show_colnames  = TRUE,
  fontsize_row   = 8,
  fontsize_col   = 8,
  border_color   = "gray",
  main           = "Metaprogram Contribution Across Samples",
  angle_col      = 90
)
```

```{r save-heatmap-presence}
pdf(file.path(out_dir, "mp_presence_heatmap.pdf"), width = 12, height = 8)
pheatmap(
  heatmap_data,
  color        = colorRampPalette(c("white", "yellow", "orange", "red"))(50),
  cluster_rows = TRUE, cluster_cols = TRUE,
  show_rownames = TRUE, show_colnames = TRUE,
  fontsize_row = 8, fontsize_col = 8,
  border_color = "gray",
  main = "Metaprogram Contribution Across Samples",
  angle_col = 90
)
dev.off()
```

---

# Spatial Visualizations (Downstream)

```{r spatial-note}
cat("Spatial visualization handled in downstream analysis:\n")
cat("  → 4_METAPROGRAMS_SPATIAL_ANALYSIS/\n")
cat("Outputs saved to:", out_dir, "\n")
```

